---
title: "Nematostella gastrula scATAC ArchR project"
execute:
  eval: false
  message: false
  warning: false
output: html_notebook
---

# Setup

Load packages and functions.

```{r}
#| eval:true
#| warning: false
#| message: false

# load in-house functions
source("metacell_downstream_functions/Downstream_functions.R")
source("metacell_downstream_functions/utils.R")
source("motif-analysis/mta_downstream_functions.R")
source("scatac_helper_functions.R")

# global settings
options(stringsAsFactors = FALSE)

# load packages
library(BSgenome.jaNemVect1.1.DToL.Assembly)
library(GenomicRanges)
library(trackViewer)
library(GSEABase)
library(data.table)
library(tidyr)
library(stringr)
library(ggplot2)
library(patchwork)
library(ggrepel)
library(ggside)
library(grid)
library(ArchR)
library(Seurat)
library(metacell)
library(lisi)

# set ggplot2 theme
theme_blank <- theme_minimal() + theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  text = element_text(size=20)
)
theme_py <- theme_light() + theme(
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.border = element_rect(colour = "black", fill = NA),
  text = element_text(size=20),
  strip.placement = "outside", 
  strip.text = element_text(size=20, color="black"),
  strip.background = element_rect(fill="white")
)
theme_set(theme_py)
```

If you are continuing analysis, you can load existing project.

```{r}
sample <- "Nvec_gastrula"
proj <- loadArchRProject(sprintf("ArchRProj_%s", sample))
plot_dir <- file.path(file.path(sprintf("ArchRProj_%s", sample), "Plots"))
dir.create(plot_dir, showWarnings = FALSE)
peak_dir <- file.path(file.path(sprintf("ArchRProj_%s", sample), "Peaks"))
dir.create(peak_dir, showWarnings = FALSE)
ann_dir <- file.path("annotation")
```

# Inputs and QCs

Input samples and files

```{r fig.width=12}
input_dir <- "~/cluster/aelek/proj/scATAC_nvec_gastrula/scatacseq-pipeline/"
samples <- c("G26_fx", "G26_fs")
cell_caller <- "EmptyDrop"
cell_caller <- "filter"
# files
input_fns <- structure(
    file.path(
      input_dir, samples, "filtered_alignments",
      cell_caller, "cell_barcodes.fragments.gz"
    ),
    names = samples
)

# assign colors
orig_cols <- hsv(
  h = seq(0.2, 0.8, length.out = length(samples)),
  s = 1,
  v = 1
)
orig_cols <- structure(orig_cols, names = samples)
```

# Clustering

Genome

```{r}
require(BSgenome.jaNemVect1.1.DToL.Assembly)
genome_dir <- "genome"
geneAnnotationFile <- file.path(
  genome_dir, "Nvec_v4_merged_annotation_sort.gtf.gz"
)
geneAnnotationDT <- fread(geneAnnotationFile)
geneAnnotationDT <- geneAnnotationDT[
  V1 %in% BSgenome.jaNemVect1.1.DToL.Assembly@seqinfo@seqnames[1:15]
]
geneAnnotationDT[, gene := str_extract(
  V9, '(?<=transcript_id ").+(?="; gene_id)'
)]
geneAnnotationDT[,TSS := V4][V7 == "-", TSS := V5]

# genes
genes <- geneAnnotationDT[V3 == "transcript"]
geneAnnotationGR <- GRanges(
  seqnames = genes[[1]],
  ranges = IRanges(genes[[4]], genes[[5]]),
  strand = genes[[7]],
  symbol = genes$gene
)

# tss
geneAnnotationTSS <- GRanges(
  seqnames = genes[[1]],
  ranges = IRanges(genes$TSS, genes$TSS),
  strand = genes[[7]],
  symbol = genes$gene
)

# exons
exons <- geneAnnotationDT[V3 == 'exon']
geneAnnotationEXON <- GRanges(
  seqnames = exons[[1]],
  ranges = IRanges(exons[[4]], exons[[5]]),
  strand = exons[[7]],
  symbol = exons$gene
)
```

Set up ArchR

```{r warning=FALSE, message=FALSE}
require(ArchR)
require(GenomicRanges)

set.seed(1)
addArchRThreads(threads = c(parallel::detectCores() - 1))
addArchRChrPrefix(chrPrefix = FALSE)

# genome annotation
genomeAnnotation <- createGenomeAnnotation(
  genome = BSgenome.jaNemVect1.1.DToL.Assembly,
  chromSizes = GRanges(
    seqnames = BSgenome.jaNemVect1.1.DToL.Assembly@seqinfo@seqnames[1:15],
    ranges = IRanges(
      start = 1, end = BSgenome.jaNemVect1.1.DToL.Assembly@seqinfo@seqlengths
    )[1:15]
  ),
  filter = FALSE
)
addArchRChrPrefix(chrPrefix = FALSE)

# gene annotations
geneAnnotation <- createGeneAnnotation(
  TSS = geneAnnotationTSS,
  exons = geneAnnotationEXON,
  genes = geneAnnotationGR
)
```

Create Arrow files

```{r}
# create arrow files
ArrowFiles <- createArrowFiles(
  inputFiles = input_fns,
  sampleNames = samples,
  # using EmptyDrop filtering + frags thresholding 200
  minTSS = 4, minFrags = 200,
  addTileMat = TRUE,
  addGeneScoreMat = TRUE,
  # https://github.com/GreenleafLab/ArchR/issues/352
  subThreading = FALSE,
  geneAnnotation = geneAnnotation,
  genomeAnnotation = genomeAnnotation,
  force = TRUE
)
```

Doublets

```{r}
ArrowFiles <- list.files(".", pattern = "arrow$", full.names = TRUE)
for (ArrowFile in ArrowFiles) {
  if (grepl("PFA", ArrowFile)) {
    knnMethod <- "UMAP"
    k <- 5
    varFeatures <- 1000
  } else {
    knnMethod <- "UMAP"
    k <- 10
    varFeatures <- 5000
  }
  doubScores <- addDoubletScores(
    input = ArrowFile,
    k = k,
    knnMethod = "LSI",
    LSIMethod = 1,
    dimsToUse = 1:20,
    nTrials = 10,
    scaleDims = TRUE,
    LSIParams = list(seed = 1, dimsToUse = 1:20, varFeatures = varFeatures),
    UMAPParams = list(n_neighbors = 30, min_dist = 0.1, metric = "cosine"),
    verbose = TRUE, force = TRUE
  )
}
```

Create ArchR project

```{r}
ArrowFiles <- list.files(".", pattern = "arrow$", full.names = TRUE)
sample <- "Nvec_gastrula"

proj <- ArchRProject(
  ArrowFiles = ArrowFiles,
  outputDirectory = sprintf("output_%s", paste(sample, collapse = "_")),
  copyArrows = FALSE,
  geneAnnotation = geneAnnotation,
  genomeAnnotation = genomeAnnotation
)

saveArchRProject(
  ArchRProj = proj,
  outputDirectory = sprintf("ArchRProj_%s", sample),
  load = TRUE
)
```

Iterative LSI

```{r}
# LSI
proj <- addIterativeLSI(
    ArchRProj = proj,
    useMatrix = "TileMatrix",
    name = "IterativeLSI",
    iterations = 4,
    varFeatures = 10000,
    selectionMethod = "var",
    dimsToUse = 1:30,
    clusterParams = list( # Seurat::FindClusters
        resolution = c(0.3),
        sampleCells = 2000,
        n.start = 10
    ),
    UMAPParams = list( # uwot::umap
      n_neighbors = 10, 
      min_dist = 0.2, spread = 1.5,
      metric = "cosine"
    ),
    force = TRUE
)
saveArchRProject(
  ArchRProj = proj,
  outputDirectory = sprintf("ArchRProj_%s", sample),
  load = TRUE
)
```

Before dublet removal:  
- numberOfCells(1): 7148  
- medianTSS(1): 17.228  
- medianFrags(1): 1420  

After doublet removal:  
- numberOfCells(1): 6882  
- medianTSS(1): 17.183  
- medianFrags(1): 1362  


```{r batch_correction, include = FALSE, eval = FALSE}
# Batch correction
proj <- addHarmony(
    ArchRProj = proj,
    reducedDims = "IterativeLSI",
    name = "Harmony",
    groupBy = "Sample"
)
```

Clustering and dimensionality reduction

```{r}
# Clustering
# dim reduction to use, IterativeLSI or Harmony
rdim <- "IterativeLSI"
# resolution, increase for more clusters (1.5 before)
resolution <- 1
proj <- addClusters(
    input = proj,
    reducedDims = rdim,
    method = "Seurat",
    name = "Clusters",
    resolution = resolution,
    dimsToUse = NULL,
    scaleDims = NULL,
    maxClusters = 40,
    testBias = TRUE,
    filterBias = TRUE,
    biasClusters = 0.01,
    biasCol = "TSSEnrichment", # nFrags
    force = TRUE
)

# UMAP
umap_params <- c(nNeighbors = 30, minDist = 0.2, spread = 1.8)
proj <- addUMAP(
    ArchRProj = proj,
    reducedDims = rdim,
    name = "UMAP",
    nNeighbors = umap_params["nNeighbors"],
    minDist = umap_params["minDist"], # affects local spread more
    spread = umap_params["spread"], # affects global spread more
    metric = "cosine",
    force = TRUE
)

# tSNE
tsne_params <- list(perplexity = 50)
proj <- addTSNE(
    ArchRProj = proj,
    reducedDims = rdim,
    name = "TSNE",
    perplexity = tsne_params[["perplexity"]],
    force = TRUE,
    check_duplicates = FALSE
)
saveArchRProject(
  ArchRProj = proj,
  outputDirectory = sprintf("ArchRProj_%s", sample),
  load = TRUE
)
```

Embedding plots

```{r}
plot_dir <- file.path(sprintf("ArchRProj_%s", sample), "Plots")
dir.create(plot_dir, showWarnings = FALSE)

# samples
orig <- unique(proj@cellColData$Sample)
orig_ord <- orig[order(as.integer(stringr::str_extract(orig, "\\d+")))]
orig_pal <- orig_cols[orig_ord]
cdt <- as.data.table(as.data.frame(proj@cellColData))[, .(Sample)]
cdt[, Sample := factor(Sample, levels = samples)]
ctab <- table(droplevels(cdt$Sample))
# clusters
cdt <- as.data.table(as.data.frame(proj@cellColData))[
  , .(Clusters, nFrags, ReadsInTSS, ReadsInPromoter, TSSEnrichment)
]
clusters <- paste0("C", seq_along(unique(cdt$Clusters)))
cluster_colors <- structure(
  paletteDiscrete(values = clusters, set = "stallion"),
  names = clusters
)

# UMAP/tSNE plots
emb <- "TSNE"
rdim <- "IterativeLSI"

p1 <- plotEmbedding(
  ArchRProj = proj,
  colorBy = "cellColData",
  pal = orig_pal,
  name = "Sample",
  embedding = emb,
  plotAs = "points",
  size = 1,
  labelSize = 0
) +
  theme(legend.title = element_blank(), legend.position="bottom") +
  guides(
    colour = guide_legend(
      nrow = 4, byrow = TRUE,
      override.aes = list(size = 3)
    )
  ) +
  labs(
    title = "Sample",
    x = paste0(emb, 1),
    y = paste0(emb, 2)
  )
p1$data$color <- str_remove(p1$data$color, "\\d+-")
p1 <- p1 +
  scale_color_manual(
    values = orig_pal,
    limits = names(ctab),
    labels = sprintf("%s (%s)", names(ctab), format(ctab, big.mark=","))
  )

p2 <- plotEmbedding(
  ArchRProj = proj,
  colorBy = "cellColData",
  name = "nFrags",
  embedding = emb,
  plotAs = "points",
  size = 1
) +
  theme(legend.title = element_blank(), legend.position = "bottom") +
  labs(title = "Unique fragments", x = paste0(emb, 1), y = paste0(emb, 2))
p2$data$alpha <- 0.5

p3 <- plotEmbedding(
  ArchRProj = proj,
  colorBy = "cellColData",
  name = "Clusters",
  pal = cluster_colors,
  embedding = emb,
  plotAs = "points",
  size = 1
) +
  theme(legend.position = "none") +
  labs(title = "Clusters", x = paste0(emb, 1), y = paste0(emb, 2))

p4 <- plotEmbedding(
  ArchRProj = proj,
  colorBy = "cellColData",
  name = "DoubletEnrichment",
  pal = ArchRPalettes$greyMagma,
  embedding = emb,
  plotAs = "points",
  size = 1
) +
  theme(legend.position = "none") +
  labs(title = "Doublet enrichment", x = paste0(emb, 1), y = paste0(emb, 2))

p5 <- plotEmbedding(
  ArchRProj = proj,
  colorBy = "cellColData",
  name = "TSSEnrichment",
  pal = ArchRPalettes$blueYellow,
  embedding = emb,
  plotAs = "points",
  size = 1
) +
  theme(legend.position = "bottom") +
  labs(title = "TSS enrichment", x = paste0(emb, 1), y = paste0(emb, 2))

pdf(
  file.path(plot_dir, sprintf("Plot-%s-%s.pdf", emb, rdim)),
  width = 12, height = 12
)
(p2 + p1) / (p3 + p5)
dev.off()

pdf(
  file.path(plot_dir, sprintf("Plot-%s-Samples-%s.pdf", emb, rdim)),
  width = 6, height = 6
)
p1
dev.off()

pdf(
  file.path(plot_dir, sprintf("Plot-%s-Fragments-%s.pdf", emb, rdim)),
  width = 6, height = 6
)
p2
dev.off()

pdf(
  file.path(plot_dir, sprintf("Plot-%s-Clusters-%s.pdf", emb, rdim)),
  width = 6, height = 6
)
p3
dev.off()

pdf(
  file.path(plot_dir, sprintf("Plot-%s-Doublets-%s.pdf", emb, rdim)),
  width = 6, height = 6
)
p4
dev.off()

pdf(
  file.path(plot_dir, sprintf("Plot-%s-TSSEnrich-%s.pdf", emb, rdim)),
  width = 6, height = 6
)
p5
dev.off()

```

Filter doublets. After doing that, re-do LSI and clustering.

```{r}
proj <- filterDoublets(proj, filterRatio = 1)
#Filtering 266 cells from ArchRProject!
#        G26_fs : 187 of 4326 (4.3%)
#        G26_fx : 79 of 2822 (2.8%)
saveArchRProject(
  ArchRProj = proj,
  outputDirectory = sprintf("ArchRProj_%s", sample),
  load = TRUE
)
```

Cluster stats

```{r}
rdim <- "IterativeLSI"

# number of cells per sample
cdt <- as.data.table(as.data.frame(proj@cellColData))[
  , .(Clusters, nFrags, ReadsInTSS, ReadsInPromoter, TSSEnrichment)
]
clusters <- paste0("C", seq_along(unique(cdt$Clusters)))
cluster_colors <- structure(paletteDiscrete(
  values = clusters, set = "stallion"
), names = clusters)
cdt[, Clusters := factor(Clusters, levels = clusters)]
cols <- c("nFrags", "ReadsInPromoter", "ReadsInTSS")
cdt[, paste0("log10", cols) := lapply(.SD, log10), .SDcols = cols]
cdt[, cols := NULL]
mcdt <- melt.data.table(cdt, id.vars = "Clusters")[!is.na(Clusters)]



cgp <- ggplot(mcdt, aes(x = Clusters, y = value, fill = Clusters)) +
    geom_violin(draw_quantiles = c(0.25, 0.5, 0.75), width = 1.4) +
    facet_grid(variable ~ ., scales = "free_y", switch = "both") +
    scale_fill_manual(values = cluster_colors) +
    theme(
        panel.grid.major = element_line(size = 0.01, colour = "grey60"),
        panel.grid.minor = element_blank(),
        legend.position = "none", strip.text = element_text(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
    )

cpgc <- ggplot(
  mcdt[variable == "TSSEnrichment"],
  aes(x = Clusters, fill = Clusters)
) +
    geom_bar(color = "black") +
    scale_fill_manual(values = cluster_colors) +
    scale_y_continuous(expand = c(0, 50)) +
    labs(y = "cells", x = "Clusters") +
    theme(
        panel.grid.major = element_line(size = 0.01, colour = "grey60"),
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
    )

sdt <- as.data.table(as.data.frame(proj@cellColData))[
  , .(Clusters, Sample)][!is.na(Clusters)]
sdt[, Clusters := factor(Clusters, levels = names(cluster_colors))]
cgs <- ggplot(sdt, aes(x = Clusters, fill = Sample)) +
    geom_bar(color = "black") +
    scale_fill_manual(values = orig_pal) +
    scale_y_continuous(expand = c(0, 50)) +
    labs(y = "cells", x = "") +
    theme(
      panel.grid.major = element_line(size = 0.01, colour = "grey60"),
      panel.grid.minor = element_blank(),
      legend.position = "top",
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
    )


plot_w <- 8
plot_h <- 10
pdf(
  file.path(plot_dir, sprintf("Clusters-Stats-%s.pdf", rdim)),
  height = plot_h,
  width = plot_w
)
cgs / cpgc
cgp
dev.off()
```

Cluster containing small/poor cells need to be removed. After doing that, re-do LSI and clustering.

```{r}
# calculate median TSS enrichment
median_mcdt <- mcdt[
  , .(value = median(value)), .(variable, Clusters)
][order(variable, Clusters)]
kdt <- rbindlist(list(
  median_mcdt,
  sdt[, .(value = .N), Clusters][, variable := "nCells"]
), use.names = TRUE)

# clusters to remove
kdct <- dcast.data.table(kdt, Clusters ~ variable, value.var = "value")
kdct[, keep := TRUE]
kdct[TSSEnrichment < 8, keep := FALSE]
kill_clusters <- kdct[keep == FALSE]$Clusters

# remove clusters and update project
remove_clusters <- as.character(kill_clusters)
remove_cells <- rownames(proj@cellColData[
  proj@cellColData$Clusters %in% remove_clusters, 
])
keep_cells <- intersect(
  rownames(proj@embeddings$UMAP$df),
  setdiff(rownames(proj@cellColData), remove_cells)
)
length(remove_cells)
length(keep_cells)
proj <- proj[keep_cells, ]
saveArchRProject(
  ArchRProj = proj,
  outputDirectory = sprintf("ArchRProj_%s", sample),
  load = TRUE
)
proj <- loadArchRProject(sprintf("ArchRProj_%s", sample))
```

# Metadata and signal files

```{r}
proj <- loadArchRProject(sprintf("ArchRProj_%s", sample))
```

Fragments

```{r}
frag_dir <- file.path(sprintf("ArchRProj_%s", sample), "Fragments")
dir.create(frag_dir, showWarnings = FALSE)
```

Save fragments file for all samples.

```{r}
# get all fragments and add sample ids to barcodes
frag_fns <- structure(file.path(
  input_dir, samples, "alignments",
  sprintf("%s.bwa.shift.fragments.gz", samples)
), names = samples)
frag_dt <- rbindlist(lapply(names(frag_fns), function(x) {
  fdt <- fread(frag_fns[x])[, V4 := paste(x, V4, sep = "#")][, Sample := x]
  fdt[, cell := FALSE][V4 %in% rownames(proj@cellColData), cell := TRUE]
}))
setnames(frag_dt, c("chr", "start", "end", "bc", "reads", "sample", "cell"))
saveRDS(frag_dt, file.path(frag_dir, "Fragments.all.RDS"))

# subset cells only
frag_dt_cell <- frag_dt[cell == TRUE][, cell := NULL]
fwrite(
  frag_dt_cell[, 1:5],
  file.path(frag_dir, "Fragments.tsv"),
  sep = "\t", col.names = FALSE
)

# save per-sample fragment files
for (i in samples) {
  message(i)
  fwrite(
    frag_dt_cell[sample == i, 1:5],
    file.path(frag_dir, sprintf("Fragments.%s.tsv", i)),
    sep = "\t", col.names = FALSE
  )
}
```

```{bash}
for i in Fragments*tsv
do
  echo ${i}
  sort -k 1,1 -k2,2n ${i} > ${i%%tsv}sort.tsv
  bgzip ${i%%tsv}sort.tsv
  tabix -p bed ${i%%tsv}sort.tsv.gz
done
```

Count cells, fragments per sample, and save stats.

```{r}
frag_dt <- readRDS(file.path(frag_dir, "Fragments.all.RDS"))

# cells per sample
cell_counts_dt <- as.data.table(
  cbind(table(proj@cellColData$Sample)),
  keep.rownames = "sample"
)
setnames(cell_counts_dt, "V1", "nCells")

# total reads per sample in cells
reads_in_cells_dt <- frag_dt[cell == TRUE][, sum(reads), sample]
setnames(reads_in_cells_dt, "V1", "nReadsCell")
counts_dt <- merge.data.table(
  cell_counts_dt, reads_in_cells_dt, by = "sample"
)

# total fagments per sample
tot_frag_all_dt <- frag_dt[, .N, sample]
setnames(tot_frag_all_dt, "N", "nFrags")
tot_frag_cell_dt <- frag_dt[cell == TRUE][, .N, sample]
setnames(tot_frag_cell_dt, "N", "nFragsCell")
tot_frag_dt <- merge.data.table(
  tot_frag_all_dt, tot_frag_cell_dt, by = "sample"
)

# median fragments per cell/barcode and sample
frag_n_dt <- frag_dt[, .N, .(sample, bc)]
frag_median_dt <- unique(frag_n_dt[, ":="(
  medianFrags = median(N),
  meanFrags = mean(N)
), sample][, .(sample, medianFrags, meanFrags)])

frag_n_cell_dt <- frag_dt[cell == TRUE][, .N, .(sample, bc)]
frag_median_sample_cell_dt <- copy(frag_n_cell_dt)[, ":="(
  medianFragsCell = median(N),
  meanFragsCell = mean(N)
), sample]
frag_median_cell_dt <- copy(frag_n_cell_dt)[, ":="(
  medianFragsCell = median(N),
  meanFragsCell = mean(N)
)][, .(medianFragsCell, meanFragsCell)]
frag_median_cell_dt <- unique(frag_median_cell_dt)
message(sprintf(
  "Median fragments per cell: %.0f", frag_median_cell_dt$medianFragsCell
))
frag_summary_dt <- rbindlist(list(
  unique(frag_median_sample_cell_dt[
    , .(sample, medianFragsCell, meanFragsCell)
  ]),
  frag_median_cell_dt[, ":="(sample = "all")]
), use.names = TRUE)
fwrite(
  frag_summary_dt,
  file.path(frag_dir, "Frags-By-Sample.tsv"),
  sep = "\t", col.names = TRUE
)

# combine
frag_counts_dt <- merge.data.table(
  frag_median_dt, frag_median_sample_cell_dt,
  by = "sample", all = TRUE
)

# NFR fragments per sample
frag_nfr_median_dt <- unique(
  as.data.table(proj@cellColData)[, ":="(
    medianFragsCellNFR = median(nFrags),
    meanFragsCellNFR = mean(nFrags)
  ), Sample][, .(Sample, medianFragsCellNFR, meanFragsCellNFR)]
)
setnames(frag_nfr_median_dt, "Sample", "sample")

frag_counts_all_dt <- merge.data.table(
  frag_counts_dt, frag_nfr_median_dt,
  by = "sample", all = TRUE
)

# combine
dt_tot_frag <- merge.data.table(
  cell_counts_dt, tot_frag_dt,
  by = "sample", all = TRUE
)
dt_counts <- merge.data.table(
  dt_tot_frag, frag_counts_all_dt,
  by = "sample", all = TRUE
)

# reads per fragments
dt_frag <- unique(frag_dt[, ":="(
  reads_median = median(reads),
  reads_mean = mean(reads)
), sample][, .(sample, reads_mean, reads_median)])
dt_frag_cell <- unique(frag_dt[cell==TRUE][, ":="(
  reads_median_cells = median(reads),
  reads_mean_cells = mean(reads)
), sample][, .(sample, reads_mean_cells, reads_median_cells)])
dt_frags <- merge.data.table(
  dt_frag, dt_frag_cell,
  by = "sample", all = TRUE
)

# total reads per sample in cells
reads_in_cells_dt <- frag_dt[cell == TRUE][, sum(reads), sample]
setnames(reads_in_cells_dt, "V1", "nReadsCell")
dt_frags_reads <- merge.data.table(dt_frags, reads_in_cells_dt, by = "sample")

# save
dt <- merge.data.table(
  dt_counts, dt_frags_reads,
  by = "sample", all = TRUE
)
dt[, sample := factor(sample, levels = samples)]
setorder(dt, sample)
setcolorder(dt, c(
  "sample", "nCells", "nFrags", "nFragsCell", "nReadsCell",
  "medianFrags", "medianFragsCell", "medianFragsCellNFR",
  "meanFrags", "meanFragsCell", "meanFragsCellNFR",
  "reads_median", "reads_median_cells", "reads_mean", "reads_mean_cells"
))
fwrite(
  dt, file.path(frag_dir, "Stats-By-Sample.tsv"),
  sep = "\t", col.names = TRUE
)
```

Save per-sample plots

```{r}
orig <- unique(proj@cellColData$Sample)
orig_ord <- orig[order(as.integer(stringr::str_extract(orig, "\\d+")))]
orig_pal <- orig_cols[orig_ord]

frag_dt <- readRDS(file.path(frag_dir, "Fragments.all.RDS"))
setnames(frag_dt, c("chr", "start", "end", "bc", "reads", "sample", "cell"))
frag_dt[, sample := factor(sample, levels = samples)]

# reads per fragments (for cells only)
frag_dt <- frag_dt[cell == TRUE]
plot_dt <- frag_dt[, reads_q := quantile(reads, 0.99), sample][reads < reads_q]
p4 <- ggplot(
  plot_dt,
  aes(sample, reads, fill = sample)) +
  geom_boxplot(outlier.colour = NA) +
  scale_fill_manual(values = orig_pal) +
  scale_y_continuous(limits = c(0, 15), expand = c(0, 0)) +
  labs(title = "Cells", y = "Reads per fragment", x = "") +
  theme_light() +
  theme(
    legend.position = "none",
    text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

pdf(
  file.path(plot_dir, "Plot-QC-Sample-Reads-Frags.pdf"),
  height = 4, width = 3
)
p4
dev.off()

# insert size distribution
frag_dt[, insert := abs(start - end)]
p5 <- ggplot(frag_dt, aes(insert, color = sample)) +
  geom_density() +
  scale_color_manual(values = orig_pal) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 600)) +
  labs(title = "", x = "Fragment Size (bp)", y = "") +
  theme_light() +
  theme(legend.position = "none", text = element_text(size = 10))

# TSS enrichment
tss_dt <- plotTSSEnrichment(ArchRProj = proj, returnDF = TRUE)
tss_dt <- setDT(as.data.frame(tss_dt))
p6 <- ggplot(tss_dt, aes(x, smoothValue, color = group, group = group)) +
  geom_line() +
  scale_color_manual(values = orig_pal) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(title = "", x = "Distance From TSS (bp)", y = "") +
   theme_light() +
  theme(legend.position = "none", text = element_text(size = 10))

# number of cells per sample
cdt <- as.data.table(as.data.frame(proj@cellColData))[, .(Sample)]
cdt[, Sample := factor(Sample, levels = samples)]
p7 <- ggplot(cdt, aes(Sample, fill = Sample)) +
  geom_bar(color = "black") +
  scale_fill_manual(values = orig_pal) +
  scale_y_log10(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "", y = "Number of cells", x = "") + theme_light() +
  theme(
    legend.position = "none",
    text = element_text(size = 10),
    axis.text.x = element_blank(),
    axis.ticks.length.x = unit(0, "mm")
  )

# fragments per cell per sample
n_frag_dt <- frag_dt[, .(frags = .N), .(sample, bc)]
n_frag_dt[, sample := factor(sample, levels = samples)]
setorder(n_frag_dt, sample)
p8 <- ggplot(n_frag_dt, aes(sample, frags, fill = sample)) +
  geom_violin(width = 1.8, draw_quantiles = c(0.5)) +
  scale_fill_manual(values = orig_pal) +
  scale_y_log10(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "", y = "Number of fragments", x = "") +
  theme_light() +
  theme(
    legend.position = "none",
    text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

plots_stats <- p7 + p5 + p8 + p6 +
  plot_layout(widths = c(1, 2))

pdf(
  file.path(plot_dir, "Plot-QC-Sample-Statistics.pdf"),
  height = 6, width = 6
)
plots_stats
dev.off()

```

Save metadata

```{r}
meta_dt <- as.data.table(
  proj@cellColData,
  keep.rownames = "Cell"
)
fwrite(meta_dt, file.path(
  sprintf("ArchRProj_%s", sample),
  sprintf("%s-cellColData.tsv", sample)
))
```

Save bigwigs for clusters

```{r}
group_by <- "Clusters"
ct_gt_num_cells <- sort(table(proj@cellColData[[group_by]]))
ct_gt_num_cells[ct_gt_num_cells > 10]
getGroupBW(
  ArchRProj = proj,
  groupBy = group_by,
  normMethod = "ReadsInTSS",
  tileSize = 25,
  maxCells = 1000,
  ceiling = 4,
  verbose = TRUE,
  threads = getArchRThreads(),
  logFile = createLogFile("getGroupBW")
)
fwrite(
  data.table(ct_gt_num_cells),
  file.path(
    sprintf("ArchRProj_%s",sample),
    "GroupBigWigs",
    group_by,
    "group_sizes.tsv
  "),
  sep = "\t", col.names = FALSE
)
```

Combine peak-by-cell matrices

```{r}
mtxDirs <- structure(unlist(lapply(samples, function(smp)
  sprintf("~/cluster/aelek/proj/scATAC_nvec/scatacseq-pipeline/%s/filtered_matrix/EmptyDrop",smp)
)), names=samples)
cells <- rownames(proj@cellColData)
  
# load matrices for individual samples
mat_list <- lapply(samples, function(sample) {
  
  mex_dir_path <- mtxDirs[sample]
  message("Reading matrix from ", mex_dir_path)
  sample_cells <- grep(sample, cells, value=TRUE)
  message("Selected cells: ", length(sample_cells))

  feature_path <- file.path(mex_dir_path, "features.txt")
  barcode_path <- file.path(mex_dir_path, "barcodes.txt")
  mtx_path <- file.path(mex_dir_path, "matrix.mtx")
   
  features <- fread(feature_path, header = F)
  barcodes <- fread(barcode_path, header = F)
  mtx <- Matrix::readMM(mtx_path)
  
  rownames(mtx) <- features[[1]]
  colnames(mtx) <- paste(sample,barcodes[[1]],sep="#")
  
  cid <- match(sample_cells,colnames(mtx))
  if (any(is.na(cid))) {
    message(sum(is.na(cid)), " cells not found in matrix")
    cid <- cid[!is.na(cid)]
  }
  mtx[,cid]
  
})

# get missing ranges for individual samples
ranges_list <- sapply(mat_list, rownames)
ranges <- unique(unlist(ranges_list, use.names = FALSE))
range_gr <- makeGRangesFromDataFrame(
  tidyr::separate(
    data.frame(range=ranges),
    range, c("seqnames","start","end"), sep = "-", remove = TRUE)
)
mat_list_full <- lapply(1:length(mat_list), function(i) {
  message(i," out of ",length(mat_list))
  mat <- mat_list[[i]]
  sample_range <- rownames(mat)
  missing_range <- setdiff(ranges,sample_range)
  missing_mat <- matrix(0, nrow=length(missing_range), ncol=ncol(mat))
  rownames(missing_mat) <- missing_range
  colnames(missing_mat) <- colnames(mat)
  full_mat <- do.call(rbind,list(mat,missing_mat))
  full_mat[ranges,]
})
mat_full <- do.call(cbind,mat_list_full)
mat_full_sparse <- Matrix(mat_full, sparse = TRUE) 

# sort
range_gr <- makeGRangesFromDataFrame(
  tidyr::separate(
    data.frame(range=rownames(mat_full_sparse)),
    range, c("seqnames","start","end"), sep = "-", remove = TRUE)
)
sort_gr <- sort(range_gr)
rownames <- Signac::GRangesToString(sort_gr)
mat_full_sorted <- mat_full_sparse[rownames,]

# save
mat_dir <- file.path(sprintf("ArchRProj_%s",sample),"Matrices")
dir.create(mat_dir,showWarnings=FALSE)
saveRDS(mat_full_sparse, file.path(mat_dir,"Matrix-Peak-Cell.rds"))
```

Save tiles matrix

```{r}
tl_mat <- getMatrixFromProject(proj, "TileMatrix", binarize=TRUE)
# if binarize=FALSE
# Error in .getMatFromArrow(ArrowFile = ArrowFile, featureDF = featureDF,  : Sparse Matrix in Arrow is Binarized! Set binarize = TRUE to use matrix!

# Error in .aggregate_and_align_all_colnames(all_colnames, strict.colnames = strict.colnames) : the DFrame objects to combine must have the same column names

useMatrix <- "TileMatrix"
ArrowFiles <- getArrowFiles(proj)
cellNames <- proj$cellNames
seL <- lapply(seq_along(ArrowFiles), function(x){
  allCells <- ArchR:::.availableCells(ArrowFile = ArrowFiles[x], subGroup = useMatrix)
  allCells <- allCells[allCells %in% cellNames]
  if(length(allCells) != 0){
    getMatrixFromArrow(
      ArrowFile = ArrowFiles[x],
      useMatrix = useMatrix,
      useSeqnames = NULL,
      cellNames = allCells, 
      ArchRProj = proj,
      verbose = FALSE,
      binarize = TRUE
    )
  }else{
    NULL
  }
}) 
cD_list <- lapply(seq_along(seL), function(x){
  colData(seL[[x]])
})
cD <- Reduce("rbind", cD_list)

# # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # 

mat_dir <- file.path(sprintf("ArchRProj_%s",sample), "TileMatrix")
dir.create(mat_dir,showWarnings=FALSE)
Matrix::writeMM(tl_mat@assays@data$TileMatrix,file.path(mat_dir,"matrix.mtx"))
writeLines(colnames(tl_mat),file.path(mat_dir,"cells"))
row_gr <- GRanges(seqnames=rowData(tl_mat)$seqnames,ranges=IRanges(start=rowData(tl_mat)$start,width=500))
rows <- Signac::GRangesToString(row_gr)
writeLines(rows,file.path(mat_dir,"tiles"))
rownames(tl_mat) <- rows
saveRDS(tl_mat, file.path(mat_dir, "Matrix-Tiles.rds"))
```

# Peak calling

Calling peaks per cluster.

```{r}
groupBy <- "Clusters"
proj <- addGroupCoverages(
  ArchRProj = proj,
  groupBy = groupBy,
  minCells = 200,
  maxCells = 1000,
  minReplicates = 2,
  maxReplicates = 2,
  sampleRatio = 1,
  force = TRUE
)

proj <- addReproduciblePeakSet(
  ArchRProj = proj,
  groupBy = groupBy,
  genomeSize = 204927134,
  extsize = 200, shift = -100,
  method = "q",
  cutOff = 0.001,
  extendSummits = 125,
  promoterRegion = c(200, 50),
  reproducibility = "1",
  plot = FALSE,
  force = TRUE,
  pathToMacs2 = "/home/anamaria/.local/bin/macs2"
)

proj <- addPeakMatrix(proj)

# update project
saveArchRProject(
  ArchRProj = proj,
  outputDirectory = sprintf("ArchRProj_%s", sample),
  load = TRUE
)

# get all peaks
fs <- list.files(
  file.path(sprintf("ArchRProj_%s", sample), "PeakCalls"),
  pattern = "*.gr.rds",
  full.names = TRUE
)
groups <- str_extract(basename(fs), "C\\d+")
names(fs) <- groups
gl <- lapply(groups,function(grp) {
  gr <- readRDS(fs[[grp]])
  gr$Group <- grp
  gr
})
names(gl) <- groups

# save peaks per cluster
peaks_subdir <- file.path(sprintf("ArchRProj_%s",sample), "Peaks", groupBy)
dir.create(peaks_subdir, showWarnings = FALSE, recursive = TRUE)

for (cluster in names(gl)) {
  print(cluster)
  dt <- as.data.table(
    gl[[cluster]]
  )[, c("seqnames", "start", "end", "Group", "score", "strand")]
  dt[, Group := paste0("peak", .I)]
  fwrite(
    dt,
    file.path(peaks_subdir, sprintf("Peaks-%s.bed", cluster)),
    sep = "\t",
    col.names = FALSE
  )
}

# save all peaks
gL <- GRangesList(gl)
gr <- unlist(gL)
gr <- sortSeqlevels(gr)
gr <- sort(gr)
dt <- as.data.table(gr)[
  , c("seqnames", "start", "end", "Group", "score", "strand")
]
dt[, Group := sprintf("peak%s.%s", .I, Group)]
fwrite(
  dt,
  file.path(peaks_subdir, "Peaks.bed"),
  sep = "\t",
  col.names = FALSE
)

# save iteratively reduced peakset
red_gr <- proj@peakSet
red_dt <- as.data.table(red_gr)[
  , c("seqnames", "start", "end", "GroupReplicate", "score", "strand")
]
fwrite(
  red_dt,
  file.path(peaks_subdir, sprintf("Peaks-iterative-reduced.bed")),
  sep = "\t",
  col.names = FALSE
)

# filter peaks
peaks_subdir_filt <- file.path(
  sprintf("ArchRProj_%s", sample),
  "Peaks",
  sprintf("%s_filtered", groupBy)
)
dir.create(peaks_subdir_filt, showWarnings = FALSE)
bed_files <- list.files(peaks_subdir)
bed_cols <- c("seqnames", "start", "end", "peak", "score", "strand")
lapply(bed_files, function(bedf) {
  message(bedf)
  dt <- fread(file.path(peaks_subdir, bedf))
  setnames(dt, bed_cols)
  dts <- dt[score > 10]
  fwrite(dts, file.path(peaks_subdir_filt,bedf), sep = "\t", col.names = FALSE)
})
```

Combine cluster peaks with bulk and pseudo-bulk consensus peaks.

```{r}
bed_cols <- c("seqnames", "start", "end", "peak", "score", "strand")

consensus_peaks_file <- file.path(
  "..", "/consensus_peaks/Jul23/consensusSeekeR-peaks.bed"
)
consensus_peaks <- fread(consensus_peaks_file) # 250bp
setnames(consensus_peaks, bed_cols)
consensus_gr <- makeGRangesFromDataFrame(
  consensus_peaks[1:3, ], keep.extra.columns = FALSE
)

groupBy <- "Clusters"
cluster_peaks_file <- file.path(
  sprintf("ArchRProj_%s", sample),
  "Peaks",
  groupBy,
  "Peaks-iterative-reduced.bed"
)
cluster_peaks <- fread(cluster_peaks_file)[, 1:3] # 250bp
setnames(cluster_peaks, bed_cols[1:3])
cluster_gr <- makeGRangesFromDataFrame(cluster_peaks)

# select consensus peaks that don't overlap with per cluster peaks
ovl <- findOverlaps(
  query = consensus_gr, subject = cluster_gr, minoverlap = 100
)
ovl_gr <- cluster_gr[subjectHits(ovl)]
nonovl <- setdiff(seq_along(consensus_gr), queryHits(ovl))
nonovl_gr <- consensus_gr[nonovl]
all_gr <- sort(c(cluster_gr, nonovl_gr))

# save all peaks
comb_peaks <- as.data.table(all_gr)
setnames(comb_peaks, c(bed_cols[1:3], "name", "strand"))
comb_peaks[, name := paste0("peak", .I)]
comb_peaks[, score := "."]
setcolorder(comb_peaks, c("seqnames", "start", "end", "name", "score", "strand"))
comb_fn <- file.path(
  sprintf("ArchRProj_%s", sample),
  "Peaks",
  sprintf("Peaks-consensus-%s.bed", groupBy)
)
fwrite(
  comb_peaks, comb_fn,
, sep = "\t", col.names = FALSE
)

# save gff3 file
rtracklayer::export.gff3(
  comb_peaks,
  str_replace(comb_fn, "\\.bed$", ".gff3")
)

# plot some numbers
ovldt <- rbindlist(list(
  as.data.table(cluster_gr[
    unique(subjectHits(ovl))
  ])[, c("set", "ovl") := list("cluster", "overlapping")],
  as.data.table(cluster_gr[
    unique(setdiff(seq_along(cluster_gr), subjectHits(ovl)))
  ])[, c("set", "ovl") := list("cluster", "non-overlapping")],
  as.data.table(consensus_gr[
    unique(queryHits(ovl))
  ])[, c("set", "ovl") := list("consensus", "overlapping")],
  as.data.table(consensus_gr[
    unique(setdiff(seq_along(consensus_gr), queryHits(ovl)))
  ])[,c("set", "ovl") := list("consensus", "non-overlapping")]
))
ovldt[,set := factor(set,levels = c("consensus", "cluster"))]
ovldt[,ovl := factor(ovl,levels = c("non-overlapping", "overlapping"))]
ovldts <- ovldt[, .N, .(set,ovl)]
gp <- ggplot(ovldts, aes(set, N, fill = ovl,  label = N)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(position = position_stack(vjust = 0.5)) +
  scale_y_continuous( 
    expand = c(0,0),
    labels = scales::unit_format(unit = "K", scale = 1e-3, accuracy = 1)
  ) +
  scale_fill_manual("sets overlap", values = c("red", "orange")) +
  labs(y = "number of peaks")
ggsave(file.path(plot_dir, "Peak-Calls-overlap.pdf"), height = 6, width = 6)
```

Add all peaks to ArchR project.

```{r}
# integrated consensus + per cluster peaks
bed_cols <- c("seqnames", "start", "end", "peak", "score", "strand")
groupBy <- "Clusters"
peaks <- fread(file.path(
  sprintf("ArchRProj_%s", sample),
  "Peaks",
  sprintf("Peaks-consensus-%s.bed", groupBy)
))
setnames(peaks, bed_cols)
peaks$name <- peaks$peak
peaks_gr <- makeGRangesFromDataFrame(
  peaks[
    seqnames %in% BSgenome.jaNemVect1.1.DToL.Assembly@seqinfo@seqnames[1:15]
  ],
  keep.extra.columns = TRUE
)

# add peaks to ArchR project
proj <- addPeakSet(proj, peakSet = peaks_gr, force = TRUE)
proj <- addPeakMatrix(proj, force = TRUE)
saveArchRProject(
  ArchRProj = proj,
  outputDirectory = sprintf("ArchRProj_%s", sample),
  load = TRUE
)

```

Save peaks matrix

```{r}
pmat <- getMatrixFromProject(proj, "PeakMatrix", binarize = FALSE)
mat_dir <- file.path(sprintf("ArchRProj_%s", sample), "PeakMatrix")
dir.create(mat_dir, showWarnings = FALSE)
pmat_mat <- pmat@assays@data$PeakMatrix
Matrix::writeMM(pmat_mat, file.path(mat_dir, "matrix.mtx"))
writeLines(colnames(pmat), file.path(mat_dir, "cells"))
row_names <- rowData(pmat)$name
writeLines(row_names, file.path(mat_dir, "peaks"))
rownames(pmat_mat) <- row_names
saveRDS(proj@peakSet, file.path(mat_dir, "peaks.rds"))
saveRDS(pmat_mat, file.path(mat_dir, "Matrix-Peaks.rds"))
```

Peaks overlapping different genomic regions

```{r}
# load gene annotation
genes_gr <- read_gtf(file.path(ann_dir, "Nvec_v4_merged_annotation_sort.gtf"))
genes_gr$name <- str_extract(
  genes_gr$attribute, '(?<=transcript_id ").+(?="; gene_id)'
)

# define regions
promoters_gr <- promoters(genes_gr, upstream = 200, downstream = 50)

transcripts_gr <- genes_gr[genes_gr$feature == "transcript"]

exons_gr <- genes_gr[genes_gr$feature == "exon"]

cds_gr <- genes_gr[genes_gr$feature == "CDS"]

genes_dt <- as.data.table(genes_gr)
genes_dt[strand=="+",feature_id:=1:.N,.(name,feature)]
genes_dt[strand=="-",feature_id:=.N:1,.(name,feature)]
genes_dt[strand=="+",feature_id_rev:=.N:1,.(name,feature)]
genes_dt[strand=="-",feature_id_rev:=1:.N,.(name,feature)]

# 5' UTR: beginning of transcript - beginning of first CDS
utr5 <- genes_dt[feature %in% c("transcript","CDS") & feature_id==1]
utr5[strand=="-",':='(end=.SD[feature=="transcript"]$end,start=.SD[feature=="CDS"]$end),name]
utr5[strand=="+",':='(start=.SD[feature=="transcript"]$start,end=.SD[feature=="CDS"]$start),name]
utr5_gr <- makeGRangesFromDataFrame(unique(utr5[,.(seqnames,start,end,strand,name)]))

# 3' UTR: end of last CDS - end of transcript
utr3 <- genes_dt[feature %in% c("transcript","CDS") & feature_id_rev==1]
utr3[strand=="-",':='(start=.SD[feature=="transcript"]$start,end=.SD[feature=="CDS"]$start),name]
utr3[strand=="+",':='(end=.SD[feature=="transcript"]$end,start=.SD[feature=="CDS"]$end),name]
utr3_gr <- makeGRangesFromDataFrame(unique(utr3[,.(seqnames,start,end,strand,name)]))

# first intron
idt <- genes_dt[feature=="exon"]
idt_plus <- idt[strand=="+"][,.SD[1:2],name][!is.na(start)&!is.na(end)][,N:=.N,name]
idt_minus <- idt[strand=="-"][order(-start)][,.SD[1:2],name][!is.na(start)&!is.na(end)][,N:=.N,name]
idt <- rbindlist(list(idt_plus,idt_minus))
first_intron_dt <- unique(idt[N>1][,':='(start=min(end),end=max(start),feature="intron"),name][,.(seqnames,start,end,strand,name)])
first_intron_gr <- makeGRangesFromDataFrame(first_intron_dt, keep.extra.columns=TRUE)

# define overlaps
promoters_ovl <- unique(peaks_gr[subjectHits(
    findOverlaps(promoters_gr, peaks_gr, minoverlap = 100)
)]$peak)

exons_ovl <- unique(peaks_gr[subjectHits(
    findOverlaps(exons_gr, peaks_gr, minoverlap = 100)
)])

transcripts_ovl <- unique(peaks_gr[subjectHits(
    findOverlaps(transcripts_gr, peaks_gr, minoverlap = 100)
)]$peak)

cds_ovl <- unique(peaks_gr[subjectHits(
    findOverlaps(cds_gr, peaks_gr, minoverlap = 100)
)]$peak)

first_introns_ovl <- unique(peaks_gr[subjectHits(
    findOverlaps(first_intron_gr, peaks_gr, minoverlap = 100)
)]$peak)

utr5_ovl <- unique(peaks_gr[subjectHits(
    findOverlaps(utr5_gr, peaks_gr, minoverlap = 100)
)]$peak)

utr3_ovl <- unique(peaks_gr[subjectHits(
    findOverlaps(utr3_gr, peaks_gr, minoverlap = 100)
)]$peak)

other_introns_ovl <- setdiff(
    transcripts_ovl,
    c(first_introns_ovl, exons_ovl, cds_ovl, utr5_ovl, utr3_ovl)
)

intergenic_ovl <- setdiff(
    peaks_gr$peak,
    transcripts_ovl
)

# plot
all_peaks_ovl <- rbindlist(list(
  "promoter" = data.table(peak = promoters_ovl),
  "CDS" = data.table(peak = cds_ovl),
  "5'UTR" = data.table(peak = utr5_ovl),
  "3'UTR" = data.table(peak = utr3_ovl),
  "first_intron" = data.table(peak = first_introns_ovl),
  "other_introns" = data.table(peak = other_introns_ovl),
  "intergenic" = data.table(peak = intergenic_ovl)
), use.names = TRUE, idcol = "overlap")
all_peaks_ovl[, overlap := factor(overlap, levels = c(
    "promoter", "5'UTR", "CDS", "3'UTR",
    "first_intron", "other_introns", "intergenic"
))]
all_peaks_ovl <- unique(all_peaks_ovl)
ovl_cols <- c(
    "promoter" = "#4393c3",
    "CDS" = "#66bd63",
    "5'UTR" = "#1a9850",
    "3'UTR" = "#a6d96a",
    "first_intron" = "#de2d26",
    "other_introns" = "#fb6a4a",
    "intron" = "#a50f15",
    "intergenic" = "#ffda1f",
    "unannotated_gene" = "#7a9695"
)
all_peaks_ovl[, peak_occur := 1:nrow(.SD), peak]

# save
fwrite(
  all_peaks_ovl,
  file.path(
    sprintf("ArchRProj_%s", sample), "Peaks",
    "annotation_peaks_genomic_overlap.tsv.gz"
  ),
  sep = "\t", col.names = TRUE
)
```

Plot genomic regions overlap for peaks

```{r}
all_peaks_ovl <- fread(file.path(
  sprintf("ArchRProj_%s", sample), "Peaks",
  "annotation_peaks_genomic_overlap.tsv.gz"
))

# include annotated promoters # # # did dnot run
peaks_annotation <- fread(file.path(ap_dir, "annotaton_peaks.tsv"))
peaks_res <- fread(file.path(ap_dir, "annotaton_promoters_decison_tree.tsv"))
gp <- rbindlist(list(
    # annotated promoters
    unique(
        peaks_res[promoter != "NO"][, .SD[1], .(peak)][, .(peak)]
    )[, overlap := "promoter"],
    # non-promoter peaks
    unique(
        copy(all_peaks_ovl)[peak %in% peaks_annotation$peak][
            overlap != "promoter" &
            !peak %in% peaks_annotation[promoter != "NO"]$peak
        ][order(peak_occur)][, .SD[1], peak]
    )[, .(peak, overlap)]
), use.names = TRUE)
gp[, promoter := ifelse(overlap == "promoter", "YES", "NO")]

# summarize
gp <- copy(all_peaks_ovl) # fix before annotating promoters
gps <- gp[, .N, overlap][, prop := N / sum(N)]
gpp <- ggplot(gps, aes("", N, fill = overlap)) +
  geom_bar(stat = "identity", width = 1, color = "black") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = ovl_cols, limits = force) +
  geom_text(
    aes(label = scales::percent(prop, accuracy = 0.1), x=1.2),
    position = position_stack(vjust=0.5),
    color = "black"
  ) +
  theme_void() + theme(legend.title = element_blank())
ggsave(
    file.path(plot_dir, "annotaiton_overlap_all_peaks_pie.pdf"),
    gpp, width = 4, height = 3
)

gpb <- ggplot(gp, aes(overlap, fill = overlap)) +
  geom_bar(width = 1, color = "black") +
  scale_fill_manual(values = ovl_cols, limits = force) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(x = "overlapping region", y = "peaks") +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.title = element_blank(), legend.position = "none"
  )
ggsave(
    file.path(plot_dir, "annotaiton_overlap_all_peaks_bar.pdf"),
    gpb, width = 4, height = 4
)

gpb <- ggplot(gp, aes(overlap, fill=overlap)) +
  geom_bar(width=1, color="black") +
  scale_fill_manual(values = ovl_cols, limits = force) +
  scale_y_continuous(expand = expansion(mult = c(0,0.02))) +
  labs(x="overlapping region", y="peaks") +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    legend.title = element_blank(), legend.position = "none"
  )
ggsave(
  file.path(plot_dir, "annotaiton_overlap_all_peaks_bar.pdf"),
  gpb,
  width = 4, height = 4
)

# # # did dnot run
mean_peak_scores_dt <- unique(assigned_peaks_bct[,.(peak,score)][,score:=pmin(1,mean(score)),peak])[]
peaks_ovl_scores_dt <- merge.data.table(gp, mean_peak_scores_dt, by="peak", sort=FALSE)

gpv <- ggplot(peaks_ovl_scores_dt, aes(overlap, score, fill=overlap)) +
  geom_violin(scale = "width", alpha = 0.8, color = "black") +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.8, color = "black") + 
  scale_fill_manual(values = ovl_cols, limits = force) +
  scale_y_continuous(expand = expansion(mult = c(0,0.01))) +
  labs(x = "overlapping region", y = "scaled average\nchromatin accessibility") +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    legend.title = element_blank(), legend.position = "none"
  )
ggsave(
  file.path(plot_dir, "annotaiton_overlap_all_peaks_violin.pdf"),
  gpv, width = 4, height = 4
)

gpb_p <- gpb + theme(
  axis.title.x = element_blank(), 
  axis.ticks.x = element_blank(), 
  axis.text.x = element_blank()
) 
gpb_p / gpv
ggsave(file.path(plot_dir, "annotaiton_overlap_all_peaks.pdf"), gpb_p / gpv, width = 5, height = 6)

```

Check if intergenic peaks overlap splice junctions, then update plots

```{r}
# splice junctions (i.e. introns)
sj_dt <- fread(file.path(ann_dir, "portcullis_all.junctions.bed"), select = 1:6)
setnames(sj_dt, c("seqnames", "start", "end", "name", "score", "strand"))
sj_gr <- makeGRangesFromDataFrame(sj_dt, keep.extra.columns = TRUE)
rtracklayer::export.gff3(
  sj_gr,
  file.path(ann_dir, "portcullis_all.junctions.gff")
)

# 5' peaks
ap_dir <- "../../scATAC_nvec_v2/clustering/SEACells/Results/ArchRProj_Nvec_TSS4_frag200/ArchRProj/Promoters"
scrna_peaks <- fread(
  file.path(ap_dir, "5prime_reduced_peaks_cell_type_gene_assignment.tsv.gz")
)
scrna_peaks <- scrna_peaks[seqnames != ""]
scrna_peaks <- unique(scrna_peaks[, .(seqnames, start, end, strand, peak)])
scrna_gr <- makeGRangesFromDataFrame(scrna_peaks, keep.extra.columns = TRUE)

# intergenic peaks
peaks_intg <- peaks_gr[
  peaks_gr$peak %in% all_peaks_ovl[overlap == "intergenic"]$peak
]

# overlap with splice junctions
sj_ovl <- findOverlaps(
  query = sj_gr, subject = peaks_intg, ignore.strand = TRUE
)
pk_hits_sj <- unique(subjectHits(sj_ovl))
pk_nohits_sj <- setdiff(1:length(peaks_intg), pk_hits_sj)
peaks_intg_ovl_sj <- peaks_intg[pk_hits_sj]
peaks_intg_novl_sj <- peaks_intg[pk_nohits_sj]

length(pk_hits_sj) / (length(pk_hits_sj) + length(pk_nohits_sj))

sj_gr[queryHits(sj_ovl)]
peaks_intg[subjectHits(sj_ovl)]

# overlap with 5' scRNA
fp_ovl <- findOverlaps(
  query = scrna_gr, subject = peaks_intg, ignore.strand = TRUE
)
pk_hits_fp <- unique(subjectHits(fp_ovl))
pk_nohits_fp <- setdiff(1:length(peaks_intg), pk_hits_fp)
peaks_intg_ovl_fp <- peaks_intg[pk_hits_fp]
peaks_intg_novl_fp <- peaks_intg[pk_nohits_fp]

length(pk_hits_fp) / (length(pk_hits_fp) + length(pk_nohits_fp))

scrna_gr[queryHits(fp_ovl)]
peaks_intg[subjectHits(fp_ovl)]

# compare 5' ovl and SJ peaks ovl
require(eulerr)
peaks_ovls <- list(
  "intergenic" = as.character(unique(peaks_intg$peak)),
  "5prime" = as.character(unique(peaks_intg[subjectHits(fp_ovl)]$peak)),
  "SJ" = as.character(unique(peaks_intg[subjectHits(sj_ovl)]$peak))
)
efit <- euler(peaks_ovls, shape = "ellipse")
pdf(
  file.path(plot_dir, "intergenic_peaks_ovl.pdf"),
  width = 2, height = 2
)
plot(efit, quantities = TRUE)
dev.off()

# compare 5' and SJ
ug_ovl <- findOverlaps(
  query = scrna_gr, subject = sj_gr, ignore.strand = FALSE
)
ug_hits_fp <- unique(subjectHits(ug_ovl))
ug_nohits_fp <- setdiff(1:length(sj_gr), ug_hits_fp)
ug_ovl_fp <- sj_gr[ug_hits_fp]
ug_novl_fp <- sj_gr[ug_nohits_fp]

length(ug_ovl_fp) / (length(ug_ovl_fp) + length(ug_novl_fp))

# update annotations
unannot_gene_ovl <- intersect(
  unique(peaks_intg[subjectHits(sj_ovl)]$peak),
  unique(peaks_intg[subjectHits(fp_ovl)]$peak)
)

all_peaks_ovl[peak %in% unannot_gene_ovl, overlap := "unannotated_gene"]
fwrite(
  all_peaks_ovl,
  file.path(
    sprintf("ArchRProj_%s", sample), "Peaks",
    "annotation_peaks_genomic_overlap.tsv.gz"
  ),
  sep = "\t", col.names = TRUE
)
```


# Gene activity scores

Assign peaks to genes

```{r}
source("../../motif-analysis/mta_downstream_functions.R")

# peaks
peaks_gr <- proj@peakSet

# genes
genes_gr <- proj@geneAnnotation$genes
genes_gr$name <- genes_gr$symbol
genes_gr$type <- "gene"

# TSS
tss_gr <- proj@geneAnnotation$TSS

# promotes
promoters_gr <- promoters(tss_gr, upstream = 200, downstream = 50)
promoters_gr$name <- promoters_gr$symbol
promoters_gr$type <- "gene" 

# genome
chrom_sizes <- as.data.table(
  proj@genomeAnnotation$chromSizes
)[, .(seqnames,end)]
setnames(chrom_sizes, c("chr", "length"))

# map peaks to genes
max_tss_dist <- 50000
genes_peaks_table <- mta_match_peaks_to_genes(
    gff_object = genes_gr,
    peak_object = peaks_gr,
    index_object = chrom_sizes,
    list_genes = NULL,
    feature_to_match = "gene",
    feature_field = "symbol",
    exclude_genes = NULL,
    max_tss_dist = max_tss_dist,
    min_overlap = 0,
    promoter_upstream = 100,
    promoter_downstream = 50,
    promoter_object = promoters_gr
)
setDT(genes_peaks_table)
```

Add weights to peaks and create gene activity score matrix.

```{r}
# peaks matrix
peaks_mat <- readRDS(file.path(
  sprintf("ArchRProj_%s", sample),
  "PeakMatrix",
  "Matrix-Peaks.rds"
))

# cell groups (clusters)
cells <- rownames(proj@cellColData)
groups <- proj@cellColData[cells,"Clusters"]
cells_groups <- data.table(cells = cells, Clusters = groups)

# calculate gene scores
gene_scores <- mta_gene_scores(
  genes_peaks_table = genes_peaks_table,
  gff_object = genes_gr,
  peak_object = peaks_gr,
  peaks_mat = peaks_mat,
  cells_groups = cells_groups
)

# save
fwrite(
  gene_scores$genes_peaks_table, 
  file.path(
    sprintf("ArchRProj_%s", sample),
    "Peaks",
    sprintf("Peaks-to-Gene-assignment-%skb.tsv", max_tss_dist / 1000)
  ),
  sep = "\t",
  col.names = TRUE
)

# save matrix
mat_dir <- file.path(sprintf("ArchRProj_%s", sample), "GeneScoreMatrix")
dir.create(mat_dir, showWarnings = FALSE)

# gs_mat <- as("dgCMatrix", gene_scores$genes_scores_matrix)
# Matrix::writeMM(gs_mat,file.path(mat_dir,"matrix.mtx"))
# writeLines(colnames(gs_mat),file.path(mat_dir,"cells"))
# writeLines(rownames(gs_mat),file.path(mat_dir,"genes"))
# rownames(pmat_mat) <- row_names

saveRDS(
  gene_scores$genes_scores_matrix,
  file.path(
    mat_dir, sprintf(
      "Matrix-Gene-Scores-Dist-Var-Weighted-%skb.rds",
      max_tss_dist / 1000
    )
  )
)

# add to project
# not possible!

```

Inspect weights distribution.

```{r}
max_tss_dist <- 50000
DT <- fread(file.path(
  sprintf("ArchRProj_%s", sample),
  "Peaks",
  sprintf("Peaks-to-Gene-assignment-%skb.tsv", max_tss_dist / 1000)
), sep = "\t", header = TRUE)

# weights dist
gp_d <- ggplot(DT, aes(x=rel_dist_to_tss)) +
  geom_density() + 
  scale_x_continuous(
    expand = c(0,0),
    labels = scales::unit_format(unit = "kb", scale = 1e-3, accuracy = 1)
  ) +
  scale_y_continuous(
    expand = c(0,0),
    labels = scales::unit_format(unit = "", scale = 1e4, accuracy = 1.1)
  ) +
  labs(x = "distance to TSS", y = "density x 1e4")
gp_d2 <- ggplot(DT, aes(x = abs(rel_dist_to_tss))) + 
  stat_ecdf(geom = "step") +
  scale_x_continuous(
    expand = c(0,0),
    labels = scales::unit_format(unit = "kb", scale = 1e-3, accuracy = 1)
  ) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = "absolute distance to TSS", y = "cdf")
gp_p <- gp_d + gp_d2
ggsave(file.path(
  sprintf("ArchRProj_%s", sample),
  "Plots",
  sprintf("Peaks-to-Gene-weights-%skb-dist.pdf", max_tss_dist / 1000)
), plot = gp_p, width = 9, height = 4)

# weights by distance from TSS
mdt <- melt.data.table(DT, measure.vars = c("w_dist", "w_var", "weight"))
mdt[,peak_location_annotation := factor(
  peak_location_annotation,
  levels = c("promoter", "gene_body", "intergenic", "other")
)]
bin_size <- 2000
mdt[, bin := cut(
  rel_dist_to_tss,
  breaks = seq(-max_tss_dist, max_tss_dist, bin_size)
)]
gp_w <- ggplot(mdt, aes(x = bin, y = value, color = variable)) +
  geom_boxplot() + facet_grid(variable ~ ., scales = "free_y") + 
  theme(
    axis.text.x = element_text(size = 8, angle = 90, vjust = 0.5, hjust=1)
  ) +
  labs(x = sprintf("distance to TSS (%s kb bins)", bin_size / 1000))
ggsave(file.path(
  sprintf("ArchRProj_%s", sample),
  "Plots",
  sprintf("Peaks-to-Gene-weights-%skb-bin.pdf", max_tss_dist / 1000)
), width = 8, height = 8)

bin2_size <- 100
mdt[,bin2 := cut(rel_dist_to_tss, breaks = seq(-1100, 1100, 100))]
gp_w2 <- ggplot(
  mdt[abs(rel_dist_to_tss) < 1e3][variable == "w_var"],
  aes(x = bin2, y = value, color = variable)
  ) +
  geom_boxplot() +
  facet_grid(variable ~ ., scales = "free_y") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x=sprintf("distance to TSS (%s bp bins)",bin2_size))
ggsave(file.path(sprintf("ArchRProj_%s",sample),"Plots",sprintf("Peaks-to-Gene-weights-var-%skb-bin.pdf",max_tss_dist/1000)), width = 8, height = 5)

# weights by class
stat_box_data <- function(x, upper_limit = max(mdt$value)) {
  data.frame(
    y = 0.95 * upper_limit,
    label = format(
      length(x),
      big.mark = ",",
      decimal.mark = ".",
      scientific = FALSE
    )
  )
}
gp_l <- ggplot(mdt, aes(x = peak_location_annotation, y = value) ) + 
  geom_boxplot() + facet_grid(. ~ variable, scales = "free_y") + 
  stat_summary(
    fun.data = stat_box_data, 
    geom = "text"
  ) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
ggsave(
  file.path(
    sprintf("ArchRProj_%s", sample),
    "Plots",
    sprintf("Peaks-to-Gene-weights-%skb-location.pdf", max_tss_dist / 1000)
  ), width = 8, height = 8
)

```

Plot weights for peaks assigned to selected genes.

```{r}
require(ggrepel)
setorder(DT, gene, start)
gene_list <- c(
  "Elav" = "Nvec_vc1.1_XM_048724864.1"
)
gene_id <- gene_list[1]
DT[grep(gene_id, gene), .(peak, w_dist, w_var, weight, peak_location_annotation)]
sdt <- mdt[grep(gene_id, gene)]
gp_p <- ggplot(
  sdt,
  aes(x = peak_midpoint, y = value, color = variable, group = variable, label = peak)
  ) +
  geom_point() +
  geom_line() +
  geom_text(data = sdt[variable == "w_var" & value > 1.5], color = "black") +
  scale_x_continuous(
    expand = c(0,0),
    labels = scales::unit_format(suffix = "", big.mark = ",")
  )  +
  geom_vline(xintercept = sdt$gene_start)
ggsave(file.path(
  sprintf("ArchRProj_%s",
  sample),
  "Plots",
  sprintf("peaks_weights_%skb_%s.pdf", max_tss_dist / 1000, gene_id)
), width = 12, height = 4)
```

Plot gene scores on UMAP
```{r include=FALSE, eval=FALSE}
# parse markers file from steger cole paper
technau_marks <- fread(file.path(
  "..", "annotation", "StegerCole_22_Markers.tsv"
))
setnames(
  technau_marks,
  c("Seb-Pedros annotation", "gene"),
  c("ID_JGI_Vienna", "gene_name")
)
technau_marks <- tidyr::separate_longer_delim(
  technau_marks, "ID_JGI_Vienna", ","
)
setDT(technau_marks)
technau_marks[, ID_JGI_Vienna := paste0("Nvec_", ID_JGI_Vienna)]
dict <- fread(file.path(
  "..", "annotation", "DICTIONARY_Nvec_old_vs_Nvec_DTOL_v2.txt"
))[, 1:2]
setnames(dict, "ID_DToL", "gene")
technau_marks <- merge.data.table(
  technau_marks, dict, by = "ID_JGI_Vienna",
  all.x = TRUE, sort = FALSE
)
fwrite(technau_marks, file.path(
  "..", "annotation", "StegerCole_22_Markers_mapped.tsv"
), sep = "\t")
```
```{r}
# load saved gene scores
gene_scores <- readRDS(file.path(
  sprintf("ArchRProj_%s", sample),
  "GeneScoreMatrix",
  "Matrix-Gene-Scores-Dist-Var-Weighted-50kb.rds"
))

# 2d proj coords
emb <- "TSNE"
tsne_df <- proj@embeddings[[emb]]$df

# markers
mark_files <- list.files(
  file.path("..", "annotation"),
  pattern = "marks.tsv",
  full.names = TRUE
)
names(mark_files) <- str_remove(basename(mark_files), ".marks.tsv")
mark_list <- sapply(mark_files, function(x) {
  dt <- fread(x)
  structure(dt[[1]], names = dt[[2]])
}, simplify = FALSE, USE.NAMES = TRUE)

# gastrula markers
technau_marks <- fread(file.path(
  "..", "annotation", "StegerCole_22_Markers_mapped.tsv"
))
technau_marks <- technau_marks[!is.na(gene)]
technau_mark_list <- sapply(unique(technau_marks$cluster), function(x) {
  cl <- str_replace(x, ".", "_")
  dt <- technau_marks[cluster == x]
  structure(dt$gene, names = dt$gene_name)
}, simplify = FALSE, USE.NAMES = TRUE)

# plots
for (ct in names(technau_mark_list)) {

  message(ct)
  plot_list <- lapply(names(technau_mark_list[[ct]]), function(gn) {
    gene <- technau_mark_list[[ct]][[gn]]
    message(gene)
    if (gene %in% rownames(gene_scores)) {
      gene_vals <- gene_scores[gene, rownames(tsne_df)]
      plot_2d_proj_gene(
        tsne_df,
        gene_vals,
        title = sprintf("%s (%s)", gn, gene),
        xlabel = paste0(emb, 1), ylabel = paste0(emb, 2),
        legend.position = c(0.85, 0.05),
        randomize = FALSE,
        interactive = FALSE
      )
    }
  })
  pdf(file.path(plot_dir, sprintf("markers.%s.%s.pdf", emb, ct)))
  print(plot_list)
  dev.off()

}
```

# Cell type annotation

```{r}
cluster_mapping <- c(
  "C1" = "endomeso_embryonic",
  "C2" = "endomeso_embryonic",
  "C3" = "cnidocytes",
  "C4" = "gland",
  "C5" = "neuron",
  "C6" = "ecto_pharyngeal_oral",
  "C7" = "ecto_embryonic_aboral",
  "C8" = "ecto_embryonic_aboral",
  "C9" = "ecto_embryonic",
  "C10" = "ecto_embryonic",
  "C11" = "ecto_embryonic"
)
proj@cellColData$cell_type <- cluster_mapping[proj@cellColData$Clusters]
```

Colors

```{r}
# add cell type colors
cts_cols <- c(
  "endomeso_embryonic" = "#ffef10",
  "cnidocytes" = "#f391f3",
  "neuron" = "#253dc7",
  "gland" = "#f89d2d",
  "ecto_pharyngeal_oral" = "#f02f2f",
  "ecto_embryonic_aboral" = "#2fd461",
  "ecto_embryonic" = "#2fced4"
)
proj@cellColData$cell_type_color <- cts_cols[proj@cellColData$cell_type]

# also add cluster colors
clusters <- paste0("C", seq_along(unique(proj@cellColData$Clusters)))
cluster_cols <- structure(
  paletteDiscrete(values = clusters, set = "stallion"),
  names = clusters
)
proj@cellColData$Clusters_color <- cluster_cols[proj@cellColData$Clusters]
```

Update project

```{r}
saveArchRProject(
  ArchRProj = proj,
  outputDirectory = sprintf("ArchRProj_%s", sample),
  load = TRUE
)
```

# Files for SEACells

Output folder

```{r}
export_dir <- file.path(
  "..", "SEACells", "Data", sprintf("ArchRProj_%s", sample)
)
dir.create(export_dir, showWarnings = FALSE)
```

Metadata

```{r}
meta <- getCellColData(proj)
cells <- rownames(meta)
write.csv(meta, file.path(export_dir, 'cell_metadata.csv'), quote = FALSE)

svd <- getReducedDims(proj)
svd <- svd[cells, ]
write.csv(svd, file.path(export_dir, 'svd.csv'), quote = FALSE)
```

Save peaks matrix

```{r}
# export dir
mat_dir <- file.path(export_dir, "PeakMatrix")
dir.create(mat_dir, showWarnings = FALSE)

# get matrix
pmat <- getMatrixFromProject(proj, "PeakMatrix", binarize = FALSE)
pmat_mat <- pmat@assays@data$PeakMatrix
pmat_mat <- pmat_mat[, cells]
peaks <- rowData(pmat)$name

# get peaks
peaks_gr <- proj@peakSet
names(peaks_gr) <- peaks_gr$name
peaks_df <- as.data.frame(peaks_gr)[peaks, ]

 # save
Matrix::writeMM(pmat_mat, file.path(mat_dir, "matrix.mtx"))
writeLines(cells, file.path(mat_dir, "cells"))
write.csv(peaks_df, file.path(mat_dir, "peaks.csv"))
```

Save gene score matrix

```{r}
# export dir
mat_dir <- file.path(export_dir, "GeneScoreMatrix")
dir.create(mat_dir, showWarnings = FALSE)

# get matrix
genes_scores_matrix <- readRDS(file.path(
  sprintf("ArchRProj_%s", sample), "GeneScoreMatrix", "Matrix-Gene-Scores-Dist-Var-Weighted-50kb.rds")
)
gscore_mat <- as(genes_scores_matrix[, cells], "dgCMatrix")
genes <- rownames(gscore_mat)

# save
Matrix::writeMM(gscore_mat,file.path(mat_dir, "matrix.mtx"))
writeLines(cells, file.path(mat_dir, "cells"))
writeLines(genes, file.path(mat_dir, "genes"))
```
