---
title: "Integration"
output: html_notebook
---

Here we integrate adult and gastrula scATAC-seq datasets.

# Setup

```{r}
library(data.table)
library(stringr)
library(ggplot2)
library(ggnewscale)
library(ggiraph)
library(ggdendro)
library(patchwork)
library(openxlsx)
library(dendextend)
require(tidytree)
require(ggtree)
require(treeio)
library(ggraph)
library(igraph)
library(tidygraph)
library(loomR)
library(BSgenome.jaNemVect1.1.DToL.Assembly)
library(GenomicRanges)
library(ArchR)
library(metacell)
library(ComplexHeatmap)
source("../scatac_helper_functions.R")
source("../../../metacell_downstream_functions/Downstream_functions.R")
source("../../../metacell_downstream_functions/Tree_functions.R")
source("../../../motif-analysis/mta_downstream_functions.R")
theme_py <- theme_light() + theme(
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.border = element_rect(colour = "black", fill = NA),
  text = element_text(size = 20),
  strip.placement = "outside",
  strip.text = element_text(size = 20, color = "black"),
  strip.background = element_rect(fill = "white")
)
theme_set(theme_py)
blank_theme <- theme_minimal() + theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.ticks = element_blank(),
  text = element_text(size = 20)
)
```

Data directories and colors

```{r}
gastr_dir <- file.path(
  "/home/anamaria/cluster/aelek/proj/scATAC_nvec_gastrula/clustering/SEACells", 
  "Results",
  "ArchRProj_Nvec_gastrula", "ArchRProj"
)
proj_gastr <- loadArchRProject(gastr_dir)

adult_dir <- file.path(
  "/home/anamaria/cluster/aelek/proj/scATAC_nvec_v2/clustering/SEACells",
  "Results", "ArchRProj_Nvec_TSS4_frag200", "ArchRProj"
)
proj_adult <- loadArchRProject(adult_dir)

res_dir <- file.path("results")
plot_dir <- file.path("plots", "Integration")
dir.create(plot_dir, showWarnings = FALSE)
data_dir <- file.path(res_dir, "Clustering")
dir.create(data_dir, showWarnings = FALSE)

# colors
ct_cols <- c(
  cnidocyte                  = "#ff42ff",
  cnidocyte_gastrula         = "#f7abf7",
  ecto_pharynx               = "#5bc0e8",
  ectoderm                   = "#51a0be",
  ecto_aboral                = "#045170",
  EMS                        = "#bdf5bd",
  EMS_ecto_boundary          = "#93dbce",
  gastro_circular_muscle_1   = "#85c90e",
  gastro_circular_muscle_2   = "#73b009",
  gastro_parietal_muscle     = "#8ceb10",
  gastro_IRF1_2              = "#c1eb05",
  gastro_somatic_gonad       = "#bde314",
  muscle_tentacle_retractor  = "#ffd700",
  muscle_mesentery_retractor = "#f0e229",
  digestive_filaments_1      = "#e33d3d",
  digestive_filaments_2      = "#d10606",
  digestive_filaments_3      = "#ad0303",
  epidermis_1                = "#04ccd4",
  epidermis_2                = "#16bacc",
  precursors_PGC             = "#bebebe",
  precursors_endoNPC         = "#8a8686",
  precursors_NPC             = "#636363",
  NPC_1                      = "#808d91",
  NPC_2                      = "#758d92",
  neuron_GATA_Islet_1        = "#0c82f7",
  neuron_GATA_Islet_2        = "#1175f0",
  neuron_Pou4_FoxL2_1        = "#101cde",
  neuron_Pou4_FoxL2_2        = "#0b16bf",
  neuron_Pou4_FoxL2_3        = "#2e39dd",
  neuronal                   = "#063cb9",
  gland                      = "#ff6f08",
  gland_mucin                = "#ff8f12"
)
cell_types <- names(ct_cols)
adult_cts <- c(
  "cnidocyte",
  "gastro_circular_muscle_1", 
  "gastro_circular_muscle_2",
  "gastro_parietal_muscle",
  "gastro_IRF1_2",
  "gastro_somatic_gonad",
  "muscle_mesentery_retractor",
  "muscle_tentacle_retractor",
  "digestive_filaments_1",
  "digestive_filaments_2",
  "digestive_filaments_3",
  "epidermis_1",
  "epidermis_2",
  "precursors_PGC",
  "precursors_endoNPC",
  "precursors_NPC",
  "neuron_GATA_Islet_1",
  "neuron_GATA_Islet_2",
  "neuron_Pou4_FoxL2_1",
  "neuron_Pou4_FoxL2_2",
  "neuron_Pou4_FoxL2_3",
  "gland"
)
gastr_cts <- c(setdiff(names(ct_cols), adult_cts))
bct_cols <- toupper(c(
  cnidocyte                 = "#FF42FF",
  ecto                      = "#51a0be",
  EMS                       = "#bdf5bd",
  gastro_circular_muscle    = "#73b009",
  gastro_parietal_muscle    = "#8ceb10",
  gastro                    = "#85c90e",  
  muscle                    = "#FFD700",
  digestive_filaments       = "#e33d3d",
  precursors                = "#BEBEBE",
  NPC                       = "#808d91",
  epidermis                 = "#04ccd4",
  neuronal                  = "#063cb9",
  neuron_GATA_Islet         = "#1175f0",
  neuron_Pou4_FoxL2         = "#101cde",
  gland                     = "#ff6f08"
))
bct_maps <- setDT(cbind.data.frame(
  cell_type = names(ct_cols),
  broad_cell_type = str_extract(names(ct_cols), paste(names(bct_cols), collapse = "|"))
))
all_cell_types <- unique(unlist(lapply(seq_len(nrow(bct_maps)), function(i) {
  c(bct_maps[i, broad_cell_type], bct_maps[i, cell_type])
})))
cell_type_rename <- c(
  "cnidocyte_precursors" = "NPC_2",
  "^muscle_1" = "muscle_tentacle_retractor",
  "^muscle_2" = "muscle_mesentery_retractor",
  "adult_muscle_1" = "adult_muscle_tentacle_retractor",
  "adult_muscle_2" = "adult_muscle_mesentery_retractor",
  "gastro_unknown_1" = "gastro_IRF1_2",
  "gastro_unknown_2" = "gastro_somatic_gonad",
  "gland_mucous" = "gland",
  "ectoderm_embryonic_oral" = "ecto_pharynx",
  "ectoderm_embryonic$" = "ectoderm",
  "ectoderm_embryonic_aboral" = "ecto_aboral",
  "mesendoderm_embryonic" = "EMS",
  "mesendoderm_ectoderm" = "EMS_ecto_boundary",
  "neuron_GATA_Islet_3" = "neuron_Pou4_FoxL2_3",
  "NPC$" = "NPC_1",
  "precursors_1" = "precursors_PGC",
  "precursors_2" = "precursors_endoNPC",
  "precursors_3" = "precursors_NPC"
)

# adult gastrula colors
stage_cols <- c("adult" = "#fdac44", "gastrula" = "#fde7ab", "both" = "#fed38d")

# samples
samples <- c(
  # whole organism
  "25m_PFA","50m_PFA",
  "3_FixSor", "9_F6k", "10_F6k", "11_F5k", "12_F", "13_F", 
  "16_FC", 
  "20_Fn", "21_Fn", "22_Fn", "23_Fn", 
  "24_Fn", 
  "25_Fn", "26_Fn", "27_Fn",
  # elav
  "18_Elav", "19_Elav",
  "28_Elav", "29_Elav", "30_Elav",
  # multiome
  "Multiome_07563AAD",
  # gastrula
  "G26_fx", "G26_fs"
)

# assign colors to samples
sample_cols <- c(
  # whole organism
  hsv(h=c(0.08,0.1), s=1, v=1),
  hsv(h=c(0.14,0.16,0.18,seq(0.2,0.45,length.out=3)), s=1, v=1),
  hsv(h=0.5, s=1, v=1),
  hsv(h=seq(0.55,0.7,length.out=4), s=1, v=1),
  hsv(h=0.75, s=1, v=1),
  hsv(h=seq(0.8,0.85,length.out = 3), s=1, v=1),
  # elav
  hsv(h=c(0.94,0.96), s=1, v=1),
  hsv(h=c(0.98, 0.99, 1), s=1, v=1),
  # multiome
  "#4D4D4D",
  # gastrula
  "#ccff00", "#cc00ff"
)
sample_colors <- structure(sample_cols, names = samples)

```

# Mapping peaks and combining metadata

Mapping between gastrula and adult peaks

```{r}
gastr_peaks <- fread(
    file.path(gastr_dir, "Peaks", "Peaks-consensus-cell_type.bed")
)
adult_peaks <- fread(
    file.path(adult_dir, "Peaks", "Peaks-consensus-cell_type.bed")
)
bed_cols <- c("seqnames", "start", "end", "peak", "score", "strand")
setnames(gastr_peaks, bed_cols)
setnames(adult_peaks, bed_cols)
gastr_peaks_gr <- makeGRangesFromDataFrame(
    gastr_peaks, keep.extra.columns = TRUE
)
adult_peaks_gr <- makeGRangesFromDataFrame(
    adult_peaks, keep.extra.columns = TRUE
)

# overlaps between two sets of peaks
ovl <- findOverlaps(
    query = gastr_peaks_gr, subject = adult_peaks_gr,
    minoverlap = 50
)
adult_peaks[, ovl := FALSE]
adult_ovl <- adult_peaks_gr[unique(subjectHits(ovl))]$peak
adult_peaks[peak %in% adult_ovl, ovl := TRUE]

gastr_peaks[, ovl := FALSE]
gastr_ovl <- gastr_peaks_gr[unique(queryHits(ovl))]$peak
gastr_peaks[peak %in% gastr_ovl, ovl := TRUE]

gastr_peaks[, .N, ovl][, perc := N / sum(N)][]
#     ovl     N     perc
#1:  TRUE 44331 0.640464
#2: FALSE 24886 0.359536
adult_peaks[, .N, ovl][, perc := N / sum(N)][]
#     ovl     N     perc
#1: FALSE 44020 0.501127
#2:  TRUE 43822 0.498873

# overlapping peaks
ovl_pks <- adult_peaks_gr[subjectHits(ovl)]
ovl_pks$peak_adult <- ovl_pks$peak
ovl_pks$peak_gastrula <- gastr_peaks_gr[queryHits(ovl)]$peak

# adult only peaks
onl_adult <- adult_peaks_gr[
    setdiff(seq_along(adult_peaks_gr), unique(subjectHits(ovl)))
]
onl_adult$peak_adult <- onl_adult$peak

# gastrula only peaks
onl_gastr <- gastr_peaks_gr[
    setdiff(seq_along(gastr_peaks_gr), unique(queryHits(ovl)))
]
onl_gastr$peak_gastrula <- onl_gastr$peak

# combine
pks_all <- c(ovl_pks, onl_adult, onl_gastr)
seqlevels(pks_all) <- as.character(unique(seqnames(adult_peaks_gr)))
pks_all <- sort(pks_all)
pks_dt <- as.data.table(pks_all)
pks_dt[is.na(pks_dt)] <- ""
pks_dt[, id := paste(seqnames, start, end, sep = "_")]
pks_dt[, id := factor(id, levels = unique(pks_dt$id))]
pks_dt[, peak := paste0("peak", as.integer(id))]

fwrite(
    pks_dt,
    file.path(res_dir, "Peaks", "Peaks_cell_type_mapped.tsv"),
    sep = "\t"
)
```

Combine metadata

```{r}
# annotation
meta_gastr <- as.data.table(
    proj_gastr@cellColData, keep.rownames = "cell"
)[, .(cell, nFrags, FRIP, TSSEnrichment, SEACell, cell_type, color)]
meta_adult <- as.data.table(
    proj_adult@cellColData, keep.rownames = "cell"
)[, .(cell, nFrags, FRIP, TSSEnrichment, SEACell, cell_type, cell_type_color)]
setnames(meta_adult, "cell_type_color", "color")
meta_all <- rbindlist(list(
    "adult" = meta_adult, "gastrula" =  meta_gastr
), idcol = "stage")
meta_all[, cell_type := str_replace_all(cell_type, "\\.", "_")]
meta_all[, color := ct_cols[as.character(cell_type)]]
meta_all[, cell_type := factor(cell_type, levels = names(ct_cols))]
setorder(meta_all, cell_type)
fwrite(meta_all, file.path(
    data_dir,
    "Annotation_Adult_Gastrula_SEACell.tsv"
), sep = "\t")

meta_all[, cell_type := factor(
    cell_type, levels = sort(unique(meta_all$cell_type))
)]
setorder(meta_all, cell_type)
meta_melt <- melt.data.table(
    meta_all,
    id.vars = c("stage", "SEACell", "cell_type"),
    measure.vars = c("nFrags", "TSSEnrichment", "FRIP")
)
meta_melt[, cell_type := factor(
    cell_type, levels = sort(unique(meta_melt$cell_type))
)]
setorder(meta_melt, cell_type)
meta_melt[, SEACell := factor(
    SEACell, levels = unique(meta_melt$SEACell)
)]
setorder(meta_melt, SEACell)
gp_qc_gastr <- ggplot(
      meta_melt[stage == "gastrula"],
      aes(SEACell, value, fill = cell_type)
    ) +
    geom_violin(draw_quantiles = c(0.25, 0.5, 0.75), width = 1.4) +
    facet_grid(variable ~ ., scales = "free_y", switch = "both") +
    scale_fill_manual(values = ct_cols) +
    theme(
        legend.position = "bottom",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank()
    )
gp_qc_adult <- ggplot(
      meta_melt[stage == "adult"],
      aes(SEACell, value, fill = cell_type)
    ) +
    geom_violin(draw_quantiles = c(0.25, 0.5, 0.75), width = 1.4) +
    facet_grid(variable ~ ., scales = "free_y", switch = "both") +
    scale_fill_manual(values = ct_cols) +
    theme(
        legend.position = "bottom",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(
            angle = 90, vjust = 0.5, hjust = 1, size = 3
        ),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank()
    )

pdf(
    file.path(plot_dir, "meta_qc.pdf"),
    width = 45, height = 12
)
gp_qc_adult
gp_qc_gastr
dev.off()
```

Peak counts with unified set of peaks

```{r}
# peaks counts
gastr_counts <- readRDS(
    file.path(gastr_dir, "PeakMatrix", "Matrix-Peaks.rds")
)
adult_counts <- readRDS(
    file.path(adult_dir, "PeakMatrix", "Matrix-Peaks.rds")
)

# mapped peaks
pks_dt <- fread(file.path(res_dir, "Peaks", "Peaks_cell_type_mapped.tsv"))

# metadata
meta_all <- fread(file.path(
    data_dir,
    "Annotation_Adult_Gastrula_SEACell.tsv"
))

# map gastrula counts
gastr_pks <- lapply(
    rownames(gastr_counts),
    function(p) pks_dt[peak_gastrula == p]$peak
)
gastr_pks_new <- unlist(gastr_pks)
gastr_pks_old <- rep(
    rownames(gastr_counts),
    times = sapply(gastr_pks, length)
)
gastr_counts_map <- gastr_counts[gastr_pks_old, ]
rownames(gastr_counts_map) <- gastr_pks_new

# map adult counts
adult_pks <- lapply(
    rownames(adult_counts),
    function(p) pks_dt[peak_adult == p]$peak
)
adult_pks_new <- unlist(adult_pks)
adult_pks_old <- rep(
    rownames(adult_counts),
    times = sapply(adult_pks, length)
)
adult_counts_map <- adult_counts[adult_pks_old, ]
rownames(adult_counts_map) <- adult_pks_new

# save files
saveRDS(
    gastr_counts_map,
    file.path(gastr_dir, "PeakMatrix", "Matrix-Peaks-mapped.rds")
)
saveRDS(
    adult_counts_map,
    file.path(radult_dir, "PeakMatrix", "Matrix-Peaks-mapped.rds")
)
```

Combine peaks

```{r}
# metadata
meta_all <- fread(file.path(
    data_dir,
    "Annotation_Adult_Gastrula_SEACell.tsv"
))

# counts
adult_counts <- readRDS(file.path(
    adult_dir, "PeakMatrix", "Matrix-Peaks-mapped.rds"
))[, meta_all[stage == "adult"]$cell]
gastr_counts <- readRDS(file.path(
    gastr_dir, "PeakMatrix", "Matrix-Peaks-mapped.rds"
))[, meta_all[stage == "gastrula"]$cell]

# combine
# adult_dt <- as.data.table(as.matrix(adult_counts), keep.rownames = "peak")
# gastr_dt <- as.data.table(as.matrix(gastr_counts), keep.rownames = "peak")
# count_dt <- merge.data.table(adult_dt, gastr_dt, by = "peak", all = TRUE)
# count_dt[is.na(count_dt)] <- 0
# peaks_mat <- as.matrix(count_dt[, -1])
# peaks_mat <- as(peaks_mat, "dgCMatrix")
# rownames(peaks_mat) <- count_dt[[1]]

# combine
adult_miss_pks <- setdiff(rownames(gastr_counts), rownames(adult_counts))
adult_miss_mat <- matrix(0, nrow = length(adult_miss_pks), ncol = ncol(adult_counts), dimnames = list(adult_miss_pks, colnames(adult_counts)))
adult_miss_mat <- as(adult_miss_mat, "dgCMatrix")
adult_cont_mat <- rbind(adult_counts, adult_miss_mat)

gastr_miss_pks <- setdiff(rownames(adult_counts), rownames(gastr_counts))
gastr_miss_mat <- matrix(0, nrow = length(gastr_miss_pks), ncol = ncol(gastr_counts), dimnames = list(gastr_miss_pks, colnames(gastr_counts)))
gastr_miss_mat <- as(gastr_miss_mat, "dgCMatrix")
gastr_cont_mat <- rbind(gastr_counts, gastr_miss_mat)

all_pks <- sort(intersect(rownames(adult_cont_mat), rownames(gastr_cont_mat)))
uni_pks <- sort(unique(c(rownames(adult_cont_mat), rownames(gastr_cont_mat))))
all.equal(all_pks, uni_pks)
peaks_mat <- cbind(
    adult_cont_mat[all_pks, ],
    gastr_cont_mat[all_pks, ]
)

# save combined peaks matrix
saveRDS(peaks_mat, file.path(
    res_dir, "PeakMatrix", "Matrix-Adult-Gastrula-Peaks-mapped.rds"
))

rm(adult_dt)
rm(gastr_dt)
rm(count_dt)
```

```{r}
# peaks counts
gastr_counts <- readRDS(
    file.path(gastr_dir, "PeakMatrix", "Matrix-Peaks.rds")
)
adult_counts <- readRDS(
    file.path(adult_dir, "PeakMatrix", "Matrix-Peaks.rds")
)

# mapped counts
adult_counts_map <- readRDS(file.path(
    adult_dir, "PeakMatrix", "Matrix-Peaks-mapped.rds"
))
gastr_counts_map <- readRDS(file.path(
    gastr_dir, "PeakMatrix", "Matrix-Peaks-mapped.rds"
))

# combined counts
peaks_mat <- readRDS(file.path(
    res_dir, "PeakMatrix", "Matrix-Adult-Gastrula-Peaks-mapped.rds"
))

# plot counts for individual peak
pid <- "peak109"
pid_a <- pks_dt[peak == pid]$peak_adult
pid_g <- pks_dt[peak == pid]$peak_gastrula
pdt_a <- data.table(
    cell = colnames(adult_counts),
    count = adult_counts[match(pid_a, rownames(adult_counts)), ]
)
pdt_a <- merge.data.table(
    pdt_a, unique(meta_all[, .(cell, SEACell, cell_type, stage)]),
    by = "cell"
)
pdt_g <- data.table(
    cell = colnames(gastr_counts),
    count = gastr_counts[match(pid_g, rownames(gastr_counts)), ]
)
pdt_g <- merge.data.table(
    pdt_g, unique(meta_all[, .(cell, SEACell, cell_type, stage)]),
    by = "cell"
)
pdt_c <- rbindlist(list(pdt_a, pdt_g))
pdt_c[, cell_type := factor(cell_type, levels = names(ct_cols))]

gp_pc <- ggplot(pdt_c[count > 0], aes(cell_type, count)) +
    geom_boxplot(aes(color = cell_type), alpha = 0.5, outlier.color = NA) +
    geom_jitter(aes(fill = cell_type, shape = stage), width = 0.2, height = 0.1) +
    scale_color_manual(values = ct_cols) +
    scale_fill_manual(values = ct_cols) +
    scale_shape_manual(values = c("adult" = 21, "gastrula" = 24)) +
    theme(
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank()
    ) +
    labs(title = sprintf("adult %s; gastrula %s", pid_a, pid_g))


# plot mapped counts for individual peak
pdt_a <- data.table(
    cell = colnames(adult_counts_map),
    count = adult_counts_map[match(pid, rownames(adult_counts_map)), ]
)
pdt_a <- merge.data.table(
    pdt_a, unique(meta_all[, .(cell, SEACell, cell_type, stage)]),
    by = "cell"
)
pdt_g <- data.table(
    cell = colnames(gastr_counts_map),
    count = gastr_counts_map[match(pid, rownames(gastr_counts_map)), ]
)
pdt_g <- merge.data.table(
    pdt_g, unique(meta_all[, .(cell, SEACell, cell_type, stage)]),
    by = "cell"
)
pdt_c <- rbindlist(list(pdt_a, pdt_g))
pdt_c[, cell_type := factor(cell_type, levels = names(ct_cols))]

gp_pc_m <- ggplot(pdt_c[count > 0], aes(cell_type, count)) +
    geom_boxplot(aes(color = cell_type), alpha = 0.5, outlier.color = NA) +
    geom_jitter(aes(fill = cell_type, shape = stage), width = 0.2, height = 0.1) +
    scale_color_manual(values = ct_cols) +
    scale_fill_manual(values = ct_cols) +
    scale_shape_manual(values = c("adult" = 21, "gastrula" = 24)) +
    theme(
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank()
    ) +
    labs(title = sprintf("%s", pid))
    

# plot combined counts for individual peak
pdt_m <- data.table(
    cell = colnames(peaks_mat),
    count = peaks_mat[match(pid, rownames(peaks_mat)), ]
)
pdt_m <- merge.data.table(
    pdt_m, unique(meta_all[, .(cell, SEACell, cell_type, stage)]),
    by = "cell"
)
pdt_m[, cell_type := factor(cell_type, levels = names(ct_cols))]

gp_pm <- ggplot(pdt_m[count > 0], aes(cell_type, count)) +
    geom_boxplot(aes(color = cell_type), alpha = 0.5, outlier.color = NA) +
    geom_jitter(aes(fill = cell_type, shape = stage), width = 0.2, height = 0.1) +
    scale_color_manual(values = ct_cols) +
    scale_fill_manual(values = ct_cols) +
    scale_shape_manual(values = c("adult" = 21, "gastrula" = 24)) +
    theme(
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank()
    ) +
    labs(title = sprintf("%s", pid))

# save plot
pdf(
    file.path(plot_dir, sprintf("cell_counts_%s.pdf", pid)),
    width = 8, height = 5
)
gp_pc
gp_pc_m
gp_pm
dev.off()

```

Inspect coverage correlation 

```{r}
peaks_mat <- readRDS(file.path(
    res_dir, "PeakMatrix", "Matrix-Adult-Gastrula-Peaks-mapped.rds"
))
idx <- sapply(colnames(peaks_mat), function(cb) {
    if (cb %in% rownames(proj_adult@cellColData)) {
        "adult"
    } else if (cb %in% rownames(proj_gastr@cellColData)) {
        "gastrula"
    }
})
idx <- unlist(idx)
cov_list <- tapply(names(idx), idx, function(x) {
    data.table(
        peak = rownames(peaks_mat),
        count = Matrix::rowMeans(peaks_mat[, x])
    )
})
names(cov_list) <- unique(idx)
cov_dt <- rbindlist(cov_list, idcol = "dataset")
cov_dt <- cov_dt[, count := mean(count), .(peak, dataset)]
cov_dt <- unique(cov_dt)
cov_dc <- dcast.data.table(cov_dt, peak ~ dataset, value.var = "count")
gp_cov <- ggplot(cov_dc, aes(adult, gastrula)) +
    geom_point(alpha = 0.2) +
    geom_abline() +
    coord_fixed() +
    scale_x_continuous(trans = "log10", breaks = c(0, 0.001, 0.01, 0.1, 1)) +
    scale_y_continuous(trans = "log10", breaks = c(0, 0.001, 0.01, 0.1, 1))
ggsave(
    file.path(plot_dir, "peaks_mean_counts_correlation.pdf"),
    width = 6, height = 6
)
```

Quantile normalize peaks matrix

```{r}
require(preprocessCore)
peaks_mat <- readRDS(file.path(
    res_dir, "PeakMatrix", "Matrix-Adult-Gastrula-Peaks-mapped.rds"
))
peaks_norm_mat <- normalize.quantiles(as.matrix(peaks_mat))
peaks_norm_mat <- as(peaks_norm_mat, "dgCMatrix")
rownames(peaks_norm_mat) <- rownames(peaks_mat)
colnames(peaks_norm_mat) <- colnames(peaks_mat)
saveRDS(peaks_norm_mat, file.path(
    res_dir, "PeakMatrix",
    "Matrix-Adult-Gastrula-Peaks-mapped-qnormalized.rds"
))
```

Inspect coverage correlation after normalization

```{r}
peaks_norm_mat <- readRDS(file.path(
    res_dir, "PeakMatrix",
    "Matrix-Adult-Gastrula-Peaks-mapped-qnormalized.rds"
))
idx <- sapply(colnames(peaks_norm_mat), function(cb) {
    if (cb %in% rownames(proj_adult@cellColData)) {
        "adult"
    } else if (cb %in% rownames(proj_gastr@cellColData)) {
        "gastrula"
    }
})
idx <- unlist(idx)
cov_list <- tapply(names(idx), idx, function(x) {
    data.table(
        peak = rownames(peaks_norm_mat),
        count = Matrix::rowMeans(peaks_norm_mat[, x])
    )
})
names(cov_list) <- unique(idx)
cov_dt <- rbindlist(cov_list, idcol = "dataset")
cov_dt <- cov_dt[, count := mean(count), .(peak, dataset)]
cov_dt <- unique(cov_dt)
cov_dc <- dcast.data.table(cov_dt, peak ~ dataset, value.var = "count")
gp_cov <- ggplot(cov_dc, aes(adult, gastrula)) +
    geom_point(alpha = 0.2) +
    geom_abline() +
    coord_fixed() +
    scale_x_continuous(
        trans = "log10",
        breaks = c(0.0001, 0.001, 0.01, 0.1, 1)
    ) +
    scale_y_continuous(
        trans = "log10",
        breaks = c(0.0001, 0.001, 0.01, 0.1, 1)
    )
ggsave(
    file.path(plot_dir, "peaks_mean_counts_qnormalized_correlation.pdf"),
    width = 6, height = 6
)
```

Aggregate per SEACell

```{r}
# peaks counts
peaks_mat <- readRDS(file.path(
    adult_dir, "PeakMatrix", "Matrix-Adult-Gastrula-Peaks-mapped.rds"
))
peaks_norm_mat <- readRDS(file.path(
    res_dir, "PeakMatrix",
    "Matrix-Adult-Gastrula-Peaks-mapped-qnormalized.rds"
))

# metadata
meta_all <- fread(file.path(
    data_dir,
    "Annotation_Adult_Gastrula_SEACell.tsv"
))

# aggregate
clusters_vector <- structure(meta_all$SEACell, names = meta_all$cell)
clusters_vector <- clusters_vector[colnames(peaks_norm_mat)]
fp_clusters <- accessibility_footprint(peaks_norm_mat, clusters_vector)
saveRDS(
    fp_clusters,
    file.path(
        data_dir,
        "Footprint_Adult_Gastrula_Peaks_SEACell_qnorm.rds"
    )
)

# sum
sum_clusters <- accessibility_summary(peaks_norm_mat, clusters_vector)
sum_clusters <- unique(sum_clusters)
saveRDS(
    sum_clusters,
    file.path(
        data_dir,
        "Sum_Adult_Gastrula_Peaks_SEACell_qnorm.rds"
    )
)


# adult peak counts
adult_counts <- readRDS(file.path(
    adult_dir, "PeakMatrix", "Matrix-Peaks-mapped.rds"
))[, meta_all[stage == "adult"]$cell]
clusters_vector <- structure(meta_all$SEACell, names = meta_all$cell)
clusters_vector <- clusters_vector[colnames(adult_counts)]
fp_clusters <- accessibility_footprint(adult_counts, clusters_vector)
saveRDS(
    fp_clusters,
    file.path(
        data_dir,
        "Footprint_Adult_Peaks_SEACell.rds"
    )
)

# gastrula peak counts
gastr_counts <- readRDS(file.path(
    gastr_dir, "PeakMatrix", "Matrix-Peaks-mapped.rds"
))[, meta_all[stage == "gastrula"]$cell]
clusters_vector <- structure(meta_all$SEACell, names = meta_all$cell)
clusters_vector <- clusters_vector[colnames(gastr_counts)]
fp_clusters <- accessibility_footprint(gastr_counts, clusters_vector)
saveRDS(
    fp_clusters,
    file.path(
        data_dir,
        "Footprint_Gastrula_Peaks_SEACell.rds"
    )
)

```

Aggregate per cell type

```{r}
# peaks counts
peaks_mat <- readRDS(file.path(
    res_dir, "PeakMatrix", "Matrix-Adult-Gastrula-Peaks-mapped.rds"
))
peaks_norm_mat <- readRDS(file.path(
    res_dir, "PeakMatrix",
    "Matrix-Adult-Gastrula-Peaks-mapped-qnormalized.rds"
))

# metadata
meta_all <- fread(file.path(
    data_dir,
    "Annotation_Adult_Gastrula_SEACell.tsv"
))

# aggregate
clusters_vector <- structure(
    paste(meta_all$stage, meta_all$cell_type, sep = "__"),
    names = meta_all$cell
)
clusters_vector <- clusters_vector[colnames(peaks_norm_mat)]
fp_clusters <- accessibility_footprint(peaks_norm_mat, clusters_vector)
saveRDS(
    fp_clusters,
    file.path(
        data_dir,
        "Footprint_Adult_Gastrula_Peaks_cell_type_qnorm.rds"
    )
)

sum_clusters <- accessibility_summary(peaks_norm_mat, clusters_vector)
sum_clusters <- unique(sum_clusters)
saveRDS(
    sum_clusters,
    file.path(
        data_dir,
        "Sum_Adult_Gastrula_Peaks_cell_type_qnorm.rds"
    )
)


# adult peak counts
adult_counts <- readRDS(file.path(
    adult_dir, "PeakMatrix", "Matrix-Peaks-mapped.rds"
))[, meta_all[stage == "adult"]$cell]
clusters_vector <- structure(
    meta_all$cell_type,
    names = meta_all$cell
)
clusters_vector <- clusters_vector[colnames(adult_counts)]
fp_clusters <- accessibility_footprint(adult_counts, clusters_vector)
saveRDS(
    fp_clusters,
    file.path(
        data_dir,
        "Footprint_Adult_Peaks_cell_type.rds"
    )
)

# gastrula peak counts
gastr_counts <- readRDS(file.path(
    gastr_dir, "PeakMatrix", "Matrix-Peaks-mapped.rds"
))[, meta_all[stage == "gastrula"]$cell]
clusters_vector <- structure(
    meta_all$cell_type,
    names = meta_all$cell
)
clusters_vector <- clusters_vector[colnames(gastr_counts)]
fp_clusters <- accessibility_footprint(gastr_counts, clusters_vector)
saveRDS(
    fp_clusters,
    file.path(
        data_dir,
        "Footprint_Gastrula_Peaks_cell_type.rds"
    )
)
```

# Clustering and integrated 2D projection

Load data for clustering

```{r}
# aggregated footprint
qnorm <- TRUE
fp_clusters <- readRDS(file.path(
    data_dir, "Footprint_Adult_Gastrula_Peaks_SEACell_qnorm.rds"
))

# metadata
meta_all <- fread(file.path(
    data_dir, "Annotation_Adult_Gastrula_SEACell.tsv"
))

```

Peaks max FC

```{r}
# max peak fotprints distribution
top_fc_adult <- apply(
    fp_clusters[, colnames(fp_clusters) %in% proj_adult$SEACell], 1, max
)
top_fc_gastr <- apply(
    fp_clusters[, colnames(fp_clusters) %in% proj_gastr$SEACell], 1, max
)
top_fc <- apply(fp_clusters, 1, max)
top_fc_dt <- data.table(
    peak = rownames(fp_clusters),
    adult = top_fc_adult,
    gastrula = top_fc_gastr,
    top_fc
)
top_fc_md <- melt.data.table(
    top_fc_dt[, .(peak, adult, gastrula)],
    id.vars = "peak"
)
gp_top_fc <- ggplot(top_fc_md, aes(value)) +
    geom_histogram(color = "white") +
    scale_x_continuous() +
    labs(x = "max FC per peak") +
    facet_grid(. ~ variable, scales = "free_x")
ggsave(
    file.path(plot_dir, "max_fc_norm_peaks.pdf"),
    width = 8, height = 4
)
```

Select peaks based on max FC

```{r}
# top peak footprints
fc_thrs <- NA
adult_ids <- which(colnames(fp_clusters) %in% proj_adult@cellColData$SEACell)
gastr_ids <- which(colnames(fp_clusters) %in% proj_gastr@cellColData$SEACell)
fc_thrs_adult <- round(
    quantile(apply(fp_clusters[, adult_ids], 1, max), 0.95)
)
# 6 for intersect peaks; 5 for union
fc_thrs_gastr <- round(
    quantile(apply(fp_clusters[, gastr_ids], 1, max), 0.95)
)
# 4 for intersect peaks; 3 for union
thrs <- sprintf("FC%s_gastrula_FC%s_adult", fc_thrs_gastr, fc_thrs_adult)
if (qnorm) thrs <- sprintf("%s_qnorm", thrs)
feats <- names(which(apply(
    fp_clusters, 1, function(x) {
        (max(x[adult_ids]) > fc_thrs_adult) |
        (max(x[gastr_ids]) > fc_thrs_gastr)
    }
)))
sum(rownames(fp_clusters) %in% feats)
mean(rownames(fp_clusters) %in% feats)

gp_top_fc_cor <- ggplot(
        top_fc_dt, aes(
            adult, gastrula,
            color = peak %in% feats
        )
    ) +
    geom_point(alpha = 0.2) +
    geom_abline() +
    coord_fixed() +
    scale_x_continuous(trans = "log10") +
    scale_y_continuous(trans = "log10") +
    scale_color_manual(
        values = c("FALSE" = "grey", "TRUE" = "red"),
        name = thrs
    ) +
    theme(legend.position = "bottom") +
    labs(title = sprintf("%s peaks", nrow(top_fc_dt[peak %in% feats])))
ggsave(
    file.path(
        plot_dir,
        sprintf("max_fc_peaks_correlation_%s.pdf", thrs)
    ),
    width = 5, height = 5
)
```

Select peaks based on variability

```{r eval = FALSE}
# peaks footprints variance
var_adult <- apply(
    fp_clusters[, colnames(fp_clusters) %in% proj_adult$SEACell], 1, var
)
var_gastr <- apply(
    fp_clusters[, colnames(fp_clusters) %in% proj_gastr$SEACell], 1, var
)
var_fc <- apply(fp_clusters, 1, var)
var_fc_dt <- data.table(
    peak = rownames(fp_clusters),
    adult = var_adult,
    gastrula = var_gastr,
    var_fc
)
var_fc_md <- melt.data.table(
    var_fc_dt[, .(peak, adult, gastrula)],
    id.vars = "peak"
)
gp_var_fc <- ggplot(var_fc_md, aes(value)) +
    geom_histogram(color = "white") +
    scale_x_continuous(trans = "log10") +
    labs(x = "variance per peak") +
    facet_grid(. ~ variable, scales = "free_x")
ggsave(
    file.path(plot_dir, "var_fc_peaks.pdf"),
    width = 8, height = 4
)

# variable peak footprints
var_thrs <- 0.1
thrs <- sprintf("var_%s", var_thrs)
feats <- names(which(apply(fp_clusters, 1, var) > var_thrs))
sum(rownames(fp_clusters) %in% feats)
mean(rownames(fp_clusters) %in% feats)

gp_top_fc_cor <- ggplot(
        var_fc_dt,
        aes(adult, gastrula, color = var_fc > var_thrs)
    ) +
    geom_point(alpha = 0.2) +
    geom_abline() +
    coord_fixed() +
    scale_x_continuous(trans = "log10") +
    scale_y_continuous(trans = "log10") +
    scale_color_manual(
        values = c("FALSE" = "grey", "TRUE" = "red"),
        name = sprintf("var > %s", var_thrs)
    ) +
    theme(legend.position = "bottom") +
    labs(title = sprintf("%s peaks", nrow(var_fc_dt[var_fc > var_thrs])))
ggsave(
    file.path(
        plot_dir,
        sprintf("var_fc_peaks_correlation_var%s.pdf", var_thrs)
    ),
    width = 5, height = 5
)
```

Select peaks using DESeq2 LTR test

```{r eval = FALSE}
require(DESeq2)

# mildly preselect peaks to make things easier
sum_cl <- sum_clusters[
    apply(sum_clusters, 1, function(x) sort(x, decreasing = FALSE)[500]) > 0,
]

# prepare metadata
sum_dt <- unique(meta_all[, .(SEACell, stage, cell_type, color)])
setnames(sum_dt, "cell_type", "celltype")
sum_dt[grep("ectoderm", celltype), celltype := "ectoderm"]
sum_dt[grep("mesendoderm", celltype), celltype := "mesendoderm"]
sum_dt[grep("cnidocyte", celltype), celltype := "cnidocyte"]
sum_dt[grep("muscle", celltype), celltype := "muscle"]
sum_dt[grep("gastro", celltype), celltype := "gastroderm"]
sum_dt[grep("epidermis", celltype), celltype := "epidermis"]
sum_dt[grep("digestive_filaments", celltype), celltype := "digestive_filaments"]
sum_dt[grep("neuron", celltype), celltype := "neuron"]
sum_dt[grep("gland", celltype), celltype := "gland"]
sum_dt[grep("precursors", celltype), celltype := "precursors"]
sum_dt[, .N, celltype][order(celltype)]
sum_df <- sum_dt[, -1]
class(sum_df) <- "data.frame"
rownames(sum_df) <- sum_dt[[1]]
sum_df <- sum_df[colnames(sum_clusters), ]

# DESeq2
dds <- DESeqDataSetFromMatrix(
  countData = sum_cl + 1,
  colData = sum_df,
  design = ~ stage + celltype
)
dds <- DESeq(dds, test = "LRT", reduced = ~ 1)
saveRDS(dds, file.path(
    data_dir, "dds_peaks_mat_peaks_adult_gastrula_broad.rds"
))
dds_res <- results(dds)
dds_qval <- p.adjust(dds_res$padj, method = "fdr")
names(dds_qval) <- rownames(dds_res)
qval_thr <- 1e-5
feats <- names(which(dds_qval < qval_thr))
thrs <- sprintf("dds_qval_%s", qval_thr)

gp_top_fc_cor <- ggplot(
        top_fc_dt,
        aes(adult, gastrula, color = peak %in% feats)
    ) +
    geom_point(alpha = 0.2) +
    geom_abline() +
    coord_fixed() +
    scale_x_continuous(trans = "log10") +
    scale_y_continuous(trans = "log10") +
    scale_color_manual(
        values = c("FALSE" = "grey", "TRUE" = "red"),
        name = sprintf("qval < %s", qval_thr)
    ) +
    theme(legend.position = "bottom") +
    labs(title = sprintf("%s peaks", nrow(top_fc_dt[peak %in% feats])))
ggsave(
    file.path(
        plot_dir,
        sprintf("max_fc_peaks_correlation_dds_broad.pdf")
    ),
    width = 5, height = 5
)
```

```{r}
# selected features
feats_select <- intersect(feats, rownames(fp_clusters))

# subset matrices
table_fp <- fp_clusters[feats_select, ]
```

## Tree based clustering

```{r}
# dmatrix <- dist(t(table_fp))
dmatrix <- tgstat::tgs_dist(t(table_fp))
# saveRDS(
#     dmatrix,
#     file.path(
#         data_dir,
#         sprintf("dist_mat_peaks_adult_gastrula_%s.RDS", thrs)
#     )
# )

# hierarchical clustering
hc <- hclust(dmatrix)
# saveRDS(
#     hc,
#     file.path(
#         data_dir,
#         sprintf("hclust_peaks_adult_gastrula_tree_%s.RDS", thrs)
#     )
# )

# NJ clustering
require(ape)
njoin <- nj(dmatrix)
# saveRDS(
#     njoin,
#     file.path(
#         data_dir,
#         sprintf("NJ_peaks_adult_gastrula_tree_%s.RDS", thrs)
#     )
# )
```

```{r}
# dmatrix <- readRDS(
#     file.path(
#         data_dir,
#         sprintf("dist_mat_peaks_adult_gastrula_%s.RDS", thrs)
#     )
# )
# hc <- readRDS(
#     file.path(
#         data_dir,
#         sprintf("hclust_peaks_adult_gastrula_tree_%s.RDS", thrs)
#     )
# )
# njoin <- readRDS(
#     file.path(
#         data_dir,
#         sprintf("NJ_peaks_adult_gastrula_tree_%s.RDS", thrs)
#     )
# )
```

Plot NJ cell type tree

```{r}
# NJ tree
# tree <- readRDS(
#     file.path(
#         data_dir,
#         sprintf("NJ_peaks_adult_gastrula_tree_%s.RDS", thrs)
#     )
# )
tree <- njoin
tips <- tree$tip.label
tips <- paste(meta_all[match(tips, SEACell)]$cell_type, tips)
tree$tip.label <- tips
meta_all[, label := paste(cell_type, SEACell)]
tree <- groupOTU(
    tree,
    sapply(
        unique(meta_all$cell_type),
        function(ct) meta_all[cell_type == ct]$label,
        USE.NAMES = TRUE,
        simplify = FALSE
    )
)
gt <- ggtree(tree, aes(color = group), ladderize = FALSE) +
  scale_color_manual(values = ct_cols) +
  theme_tree2() +
  theme(
    axis.text = element_blank(),
    legend.position = "none"
  )
gt_lin <- gt +
    geom_tiplab(size = 2) +
    theme_tree2() +
    theme(legend.position = "none")
ggsave(file.path(
    plot_dir,
    sprintf("NJ_peaks_adult_gastrula_tree_%s.pdf", thrs)
), height = 46, width = 12)
gt_cir <- gt + layout_inward_circular(xlim = 80)
ggsave(file.path(
    plot_dir,
    sprintf(
        "NJ_peaks_adult_gastrula_tree_%s_circular.pdf",
        thrs
    )
), height = 24, width = 24)
```

Plot hclust cell type tree

```{r eval=FALSE}
# hclust tree
tree <- readRDS(file.path(
    data_dir, sprintf("hclust_peaks_adult_gastrula_tree_%s.RDS", thrs)
))
tree <- as.phylo(tree)

tips <- tree$tip.label
tips <- paste(meta_all[match(tips, SEACell)]$cell_type, tips)
tree$tip.label <- tips

meta_all[, label := paste(cell_type, SEACell)]
tree <- groupOTU(
    tree,
    sapply(
        unique(meta_all$cell_type),
        function(ct) meta_all[cell_type == ct]$label,
        USE.NAMES = TRUE,
        simplify = FALSE
    )
)

gt <- ggtree(tree, aes(color = group), ladderize = FALSE) +
  scale_color_manual(values = ct_cols) +
  theme_tree2() +
  theme(
    axis.text = element_blank(),
    legend.position = "bottom"
  )
gt_lin <- gt +
    geom_tiplab(size = 2) +
    theme_tree2() +
    scale_x_continuous(expand = c(0, 100)) +
    theme(legend.position = "none")
ggsave(file.path(plot_dir, sprintf(
    "hclust_peaks_adult_gastrula_tree_%s.pdf", thrs
)), height = 46, width = 12)
gt_cir <- gt + layout_inward_circular(xlim = 200)
ggsave(file.path(plot_dir, sprintf(
    "hclust_peaks_adult_gastrula_tree_%s_circular.pdf", thrs
)), height = 24, width = 24)
```

## 2D projection

UMAP

```{r}
require(uwot)
set.seed(1950)

# UAMP
sca_umap <- uwot::umap(
  t(table_fp),
  n_neighbors = 35,
  metric = "cosine",
  spread = 1,
  min_dist = 0.4,
  local_connectivity = 1,
  repulsion_strength = 1
)

sca_umap_dt <- as.data.table(sca_umap, keep.rownames = "SEACell")
setnames(sca_umap_dt, c("SEACell", "UMAP1", "UMAP2"))
sca_umap_dt <- merge.data.table(
  sca_umap_dt,
  unique(meta_all[, .(stage, SEACell, cell_type, color)]),
  by = "SEACell", all.x = TRUE, sort = FALSE
)
sca_umap_dt[cell_type == "gland_mucous", cell_type := "gland"]
sca_umap_dt[cell_type == "mesendoderm_ectoderm", cell_type := "mesendoderm_oral"]
fwrite(sca_umap_dt, file.path(
    data_dir, sprintf("SEACells_adult_gastrula_UMAP_%s.tsv", thrs)
))

sca_umap_dt <- fread( file.path(data_dir, sprintf("SEACells_adult_gastrula_UMAP_%s.tsv", thrs)))

# rename cell types
sca_umap_dt[, cell_type := str_replace_all(cell_type, cell_type_rename)]

# plot
gp_umap <- ggplot(
        sca_umap_dt,
        aes(UMAP1, UMAP2, shape = stage, color = stage, fill = cell_type)
    ) +
    geom_point(
        data = sca_umap_dt[stage == "adult"], inherit.aes = TRUE, alpha = 0.9
    ) +
    geom_point(
        data = sca_umap_dt[stage == "gastrula"], inherit.aes = TRUE
    ) +
    scale_fill_manual(
        values = ct_cols
    ) +
    scale_shape_manual(
        values = c("adult" = 21, "gastrula" = 24)
    ) +
    scale_color_manual(
        values = c("adult" = "#383838", "gastrula" = "#000000")
    ) +
    coord_fixed() +
    guides(
        fill = guide_legend(override.aes = list(size = 5, shape = 21)),
        shape = guide_legend(override.aes = list(size = 5))
    )
ggsave(
    file.path(
        plot_dir, sprintf("UMAP_SEACells_adult_gastrula_%s.pdf", thrs)
    ),
    width = 12, height = 6
)
```

Calculate relative coordinates for single cells 

```{r}
# dim reductions single cells unintegrated
umap_adult_sc <- fread(file.path(dirname(adult_dir), "X_umap.cells.csv"), header = TRUE)
setnames(umap_adult_sc, c("Cell", "UMAP1_sc", "UMAP2_sc"))
umap_gastr_sc <- fread(file.path(dirname(gastr_dir), "X_umap.cells.csv"), header = TRUE)
setnames(umap_gastr_sc, c("Cell", "UMAP1_sc", "UMAP2_sc"))

# dim reductions SEACells unintegrated
umap_adult_mc <- fread(file.path(dirname(adult_dir), "X_umap.seacells.csv"), header = TRUE)[,1:3]
setnames(umap_adult_mc, c("SEACell", "UMAP1_mc", "UMAP2_mc"))
umap_gastr_mc <- fread(file.path(dirname(gastr_dir), "X_umap.seacells.csv"), header = TRUE)[,2:4]
setnames(umap_gastr_mc, c("SEACell", "UMAP1_mc", "UMAP2_mc"))

# dim reductions SEACells integrated
thrs <- "FC3_gastrula_FC5_adult_qnorm"
fn <- sprintf("SEACells_adult_gastrula_UMAP_%s.tsv", thrs)
umap_mc <- fread(file.path(data_dir, fn))

# combine uintegrated coordinates for single cells and seacells
umap_sc <- merge.data.table(
  rbindlist(list(umap_adult_sc, umap_gastr_sc)),
  annc[, .(Cell, SEACell)],
  by = "Cell"
)
umap_sc_mc <- merge.data.table(
  rbindlist(list(umap_adult_mc, umap_gastr_mc)),
  umap_sc,
  by = "SEACell"
)
# calculate coordinates shift of single cells relative to SEACells
umap_sc_mc[, UMAP1_diff := UMAP1_sc - UMAP1_mc]
umap_sc_mc[, UMAP2_diff := UMAP2_sc - UMAP2_mc]

# add coordinates shift of single cells to integrated SEACells coordinates
umap_sc_mc <- merge.data.table(umap_sc_mc, umap_mc, by = "SEACell")
setnames(umap_sc_mc, c("UMAP1", "UMAP2"), c("UMAP1_mc_int", "UMAP2_mc_int"))
umap_sc_mc[, UMAP1 := UMAP1_mc_int + UMAP1_diff]
umap_sc_mc[, UMAP2 := UMAP2_mc_int + UMAP2_diff]
umap_sc <- umap_sc_mc[, .(Cell, UMAP1, UMAP2, stage, cell_type, color)]

# rename cell types
umap_sc[, cell_type := str_replace_all(cell_type, cell_type_rename)]

# save
fwrite(
  umap_sc,
  file.path(data_dir, str_replace(fn, "SEACells", "Single cells")),
  sep = "\t"
)

# plot
gp_umap <- ggplot(
        umap_sc,
        aes(UMAP1, UMAP2, shape = stage, color = cell_type)
    ) +
    geom_point(
        data = umap_sc[stage == "adult"], inherit.aes = TRUE, alpha = 0.9
    ) +
    geom_point(
        data = umap_sc[stage == "gastrula"], inherit.aes = TRUE
    ) +
    scale_color_manual(
        values = ct_cols
    ) +
    scale_shape_manual(
        values = c("adult" = 16, "gastrula" = 17)
    ) +
    coord_fixed() +
    guides(
        fill = guide_legend(override.aes = list(size = 5, shape = 21)),
        shape = guide_legend(override.aes = list(size = 5))
    )
ggsave(
    file.path(
        plot_dir, sprintf("UMAP_single_cell_adult_gastrula_%s.pdf", thrs)
    ),
    width = 12, height = 6
)
```

# Data for shiny app

Results directories.

```{r}
require(loomR)
require(ArchR)
source("../scatac_helper_functions.R")
shiny_dir <- "../../../scATAC_nvec_v2/nvec-scatac-shiny/Data/"
```

Load data for single cell and SEACells.

```{r}
# single cell  matrix
adult_mat_sc <- readRDS(file.path(adult_dir, "GeneScoreMatrix", "Matrix-Gene-Scores-Dist-Var-Weighted-50kb.rds"))
gastr_mat_sc <- readRDS(file.path(gastr_dir, "GeneScoreMatrix", "Matrix-Gene-Scores-Dist-Var-Weighted-50kb.rds"))

# SEACell matrix
adult_mat_mc <- readRDS(file.path(dirname(adult_dir), "Data_filtered", "Footprint_SEACell.rds"))
gastr_mat_mc <- readRDS(file.path(dirname(gastr_dir), "Data", "Footprint_SEACell.rds"))

# subset to the same set of genes
gns_sc <- intersect(rownames(adult_mat_sc), rownames(gastr_mat_sc))
gns_mc <- intersect(rownames(adult_mat_mc), rownames(gastr_mat_mc))
gns <- intersect(gns_mc, gns_sc) 
gs_mat_sc <- cbind(adult_mat_sc[gns, ], gastr_mat_sc[gns, ])
gs_mat_mc <- cbind(adult_mat_mc[gns, ], gastr_mat_mc[gns, ])

# SEACell annotation
adult_annc <- as.data.table(proj_adult@cellColData)
gastr_annc <- as.data.table(proj_gastr@cellColData)
cols <- intersect(colnames(adult_annc), colnames(gastr_annc))
annc <- rbindlist(list(
  adult = adult_annc[, ..cols], 
  gastrula = gastr_annc[, ..cols]
), use.names = TRUE, idcol = "stage")

# order cell types
annc[, cell_type := str_replace_all(cell_type, cell_type_rename)]
annc[cell_type=="cnidocyte" & stage=="gastrula", cell_type:="cnidocyte_gastrula"]
annc[cell_type=="gland" & stage=="gastrula", cell_type:="gland_mucin"]
annc[, cell_type := factor(cell_type, levels = cell_types)]
setorder(annc, "cell_type", "SEACell")

# add colors and broad cell types
annc[, c("color", "cell_type_color", "broad_cell_type", "broad_cell_type_color") := NULL]
annc[, cell_type_color := ct_cols[cell_type]]
annc <- merge.data.table(annc, bct_maps, by="cell_type", all.x = TRUE, sort = FALSE)
annc[, broad_cell_type_color := bct_cols[broad_cell_type]]
annc[, Sample_color := sample_colors[Sample]]

# SEACell annotations
mcols <- c("SEACell", "cell_type", "cell_type_color", "broad_cell_type", "broad_cell_type_color")
ann <- unique(annc[, ..mcols])
SEACellColData <- copy(ann)
class(SEACellColData) <- "data.frame"
rownames(SEACellColData) <- SEACellColData$SEACell

# single cell annotations
mcols <- c("Cell","SEACell", "cell_type", "cell_type_color", "broad_cell_type", "broad_cell_type_color","stage","Sample","Sample_color","TSSEnrichment","ReadsInTSS","ReadsInPromoter","PromoterRatio","PassQC","NucleosomeRatio","nMultiFrags","nMonoFrags","nFrags","nDiFrags","DoubletScore","DoubletEnrichment")
cellColData <- annc[, ..mcols]
class(cellColData) <- "data.frame"
rownames(cellColData) <- cellColData$Cell

# subset for common cells and SEACells, and order 
cells <- intersect(rownames(cellColData),colnames(gs_mat_sc))
if (length(cells) != ncol(gs_mat_sc) | length(cells) != nrow(cellColData))
  message("I will subset cells (", length(cells), " of ", max(ncol(gs_mat_sc), nrow(cellColData)),")")
gs_mat_sc <- gs_mat_sc[,cells]
cellColData <- cellColData[cells,]
SEACells <- intersect(rownames(SEACellColData),colnames(gs_mat_mc))
if (length(SEACells) != ncol(gs_mat_mc) | length(SEACells) != nrow(SEACellColData))
  message("I will subset SEACells")
gs_mat_mc <- gs_mat_mc[,SEACells]
SEACellColData <- SEACellColData[SEACells,]

# save
saveRDS(gs_mat_mc, file.path(shiny_dir, sprintf("%s_gene_scores.rds", id)))
fwrite(SEACellColData, file.path(shiny_dir, sprintf("%s_mc.tsv", id)), sep = "\t")
fwrite(cellColData, file.path(shiny_dir, sprintf("%s_sc.tsv", id)), sep = "\t")

# UMAP
umap_sc <- fread(file.path(data_dir, "Single cells_adult_gastrula_UMAP_FC3_gastrula_FC5_adult_qnorm.tsv"))[,1:3]
umap_mc <- fread(file.path(data_dir, "SEACells_adult_gastrula_UMAP_FC3_gastrula_FC5_adult_qnorm.tsv"))[,1:3]
fwrite(umap_mc, file.path(shiny_dir, sprintf("%s_UMAP.tsv", id), sep = "\t"))

# convert
class(umap_sc) <- "data.frame"
rownames(umap_sc) <- umap_sc$Cell
umap_sc <- umap_sc[cells,2:3]
class(umap_mc) <- "data.frame"
rownames(umap_mc) <- umap_mc$SEACell
umap_mc <- umap_mc[SEACells,2:3]

# gene annotations
tfann <- fread("../annotation/tfs.Nvec.tsv",header=TRUE)
gene_ann <- fread("../annotation/Nvec_annotation_v3_2020-10-23_DToL_names",header=TRUE)
setnames(gene_ann,colnames(gene_ann)[1:3],c("gene","pfam","name"))
gene_ann <- rbindlist(list(
    gene_ann[,1:3],
    data.table(gene = rownames(gs_mat_sc)[!rownames(gs_mat_sc) %in% gene_ann$gene])[,pfam:=""][,name:=""]
))
class(gene_ann) <- "data.frame"
rownames(gene_ann) <- gene_ann$gene
geneData <- gene_ann[rownames(gs_mat_sc),]
```

Save loom files.

```{r}
id <- "integrated"

# save loom file for single cells
loom_file <- file.path(shiny_dir, sprintf("%s_sc.loom",id))
loomR::create(
    filename = loom_file, 
    data = t(gs_mat_sc), 
    cell.attrs = as.list(cellColData), 
    gene.attrs = as.list(geneData), 
    do.transpose = FALSE
)

lfile <- loomR::connect(loom_file, mode="r+")
lfile$add.graph.matrix(mat = as.matrix(umap_sc), MARGIN = 1, name = "UMAP")
lfile$close_all()

fwrite(as.data.table(cellColData), str_replace(loom_file, "loom", "tsv"), sep = "\t")

# save loom file for SEACells
loom_file <- file.path(shiny_dir, sprintf("%s_mc.loom",id))
loomR::create(
    filename = loom_file, 
    data = t(gs_mat_mc), 
    cell.attrs = as.list(SEACellColData), 
    gene.attrs = as.list(geneData), 
    do.transpose = FALSE
)
lfile <- loomR::connect(loom_file, mode="r+")
lfile$add.graph.matrix(mat=as.matrix(umap_mc), MARGIN = 1, name = "UMAP")
lfile$close_all()
```

Save metacell-like footprints and fraction of accessible cells.

```{r}
id <- "integrated"
for (x in c("cell_type","broad_cell_type")) {
  clusters_vector <- structure(cellColData[,x], names=cellColData$Cell)
  clusters_vector <- clusters_vector[colnames(gs_mat_sc)]
  fp_clusters <- accessibility_footprint(gs_mat_sc, clusters_vector)
  saveRDS(fp_clusters, file.path(shiny_dir, sprintf("Footprint_%s_%s.rds",id,x)))
  sum_clusters <- accessibility_summary(gs_mat_sc, clusters_vector, sum)
  saveRDS(sum_clusters, file.path(shiny_dir, sprintf("Sum_%s_%s.rds",id,x)))
}
```

# Marker peaks

Load data for clustering

```{r}
# metadata
meta_all <- fread(file.path(
    data_dir, "Annotation_Adult_Gastrula_SEACell.tsv"
))

# peaks based seacells clustering
# fc_thrs_gastr <- 4
# fc_thrs_adult <- 6
```

Peaks in SEACells heatmap

```{r}
# seacell aggregated footprint
fp_clusters <- readRDS(file.path(
    data_dir, "Footprint_Adult_Gastrula_Peaks_SEACell_qnorm.rds"
))

# subset peaks
adult_ids <- which(colnames(fp_clusters) %in% proj_adult@cellColData$SEACell)
gastr_ids <- which(colnames(fp_clusters) %in% proj_gastr@cellColData$SEACell)
feats <- names(which(apply(
    fp_clusters, 1, function(x) {
        (max(x[adult_ids]) > fc_thrs_adult) | # 8
        (max(x[gastr_ids]) > fc_thrs_gastr)   # 5
    }
)))
pks_mat <- fp_clusters[feats, ]

# order seacells
pks_mat <- pks_mat[, unique(meta_all$SEACell)]

# order peaks
pks_mat <- pks_mat[
    order(apply(pks_mat, 1, which.max)),
]

# plot
require(ComplexHeatmap)

col_vec_pks <- c("#ffffff", "#fddbc7", "#ef8a62", "#b2182b", "#520c52")
col_vec_pks <- c("#ffffff", "#e2ece2", "#7bccc4", "#43a2ca", "#0868ac")
col_fun_pks <- circlize::colorRamp2(
    seq(0, quantile(pks_mat, 0.99), length.out = length(col_vec_pks)),
    col_vec_pks
)

meta_ann <- unique(meta_all[, .(SEACell, cell_type, stage)])
col_ann <- HeatmapAnnotation(
  which = "column",
  "cell type" = as.character(
    meta_ann[
      match(colnames(pks_mat), SEACell)
    ]$cell_type
  ),
  "stage" = as.character(
    meta_ann[
      match(colnames(pks_mat), SEACell)
    ]$stage
  ),
  col = list(
    "cell type" = ct_cols,
    "stage" = c("adult" = "#f1cc65", "gastrula" = "#b229d4")
  )
)

row_ann <- HeatmapAnnotation(
  which = "row",
  "cell type" = as.character(
    meta_ann[
      match(colnames(pks_mat)[apply(pks_mat, 1, which.max)], SEACell)
    ]$cell_type
  ),
  col = list("cell type" = ct_cols),
  show_legend = FALSE
)

hm_pks_fp <- Heatmap(
  pks_mat, name = "FC",
  col = col_fun_pks,
  border = TRUE, use_raster = TRUE,
  cluster_rows = FALSE, cluster_columns = FALSE,
  show_row_dend = FALSE, show_column_dend = FALSE,
  show_row_names = FALSE, show_column_names = FALSE,
  bottom_annotation = col_ann, left_annotation = row_ann,
  row_title = sprintf("%i marker peaks", nrow(pks_mat))
)

pdf(
    file.path(plot_dir, "Adult_gastrula_SEACell_marker_peaks_fp.pdf"),
    width = 8, height = 10
)
print(hm_pks_fp)
dev.off()
```

Peaks in cell types heatmap

```{r}
# cell type aggregated footprint
fp_clusters <- readRDS(file.path(
    data_dir, "Footprint_Adult_Gastrula_Peaks_cell_type_qnorm.rds"
))

# subset peaks
adult_ids <- grep("adult", colnames(fp_clusters))
gastr_ids <- grep("gastrula", colnames(fp_clusters))
feats <- names(which(apply(
    fp_clusters, 1, function(x) {
        (max(x[adult_ids]) > 2) |
        (max(x[gastr_ids]) > 2)
    }
)))
pks_mat <- fp_clusters[feats, ]

# order cell types
ct_order <- unique(paste(meta_all$stage, meta_all$cell_type, sep = "__"))
all(ct_order %in% colnames(pks_mat))
all(colnames(pks_mat) %in% ct_order)
pks_mat <- pks_mat[, ct_order]

# order peaks
pks_mat <- pks_mat[
    order(apply(pks_mat, 1, which.max)),
]

# plot
require(ComplexHeatmap)
# pks_mat <- pks_mat^2

col_vec_pks <- c("#ffffff", "#fddbc7", "#ef8a62", "#b2182b", "#520c52")
col_vec_pks <- c("#ffffff", "#e2ece2", "#7bccc4", "#43a2ca", "#0868ac")
col_fun_pks <- circlize::colorRamp2(
    seq(0, quantile(pks_mat, 0.99), length.out = length(col_vec_pks)),
    col_vec_pks
)

meta_ann <- unique(meta_all[, .(cell_type, stage)])
col_ann <- HeatmapAnnotation(
  which = "column",
  "cell type" = str_remove(colnames(pks_mat), "(adult|gastrula)__"),
  "stage" = str_extract(colnames(pks_mat), "(adult|gastrula)"),
  col = list(
    "cell type" = ct_cols,
    "stage" = c("adult" = "#f1cc65", "gastrula" = "#b229d4")
  )
)

row_ann <- HeatmapAnnotation(
  which = "row",
  "cell type" = as.character(
      str_remove(colnames(pks_mat), "(adult|gastrula)__")[
        apply(pks_mat, 1, which.max)
      ]
  ),
  col = list("cell type" = ct_cols),
  show_legend = FALSE
)

hm_pks_fp <- Heatmap(
  pks_mat, name = "FC^2",
  col = col_fun_pks,
  border = TRUE, use_raster = TRUE,
  cluster_rows = FALSE, cluster_columns = FALSE,
  show_row_dend = FALSE, show_column_dend = FALSE,
  show_row_names = FALSE, show_column_names = FALSE,
  bottom_annotation = col_ann, left_annotation = row_ann,
  row_title = sprintf("%i marker peaks", nrow(pks_mat))
)

pdf(
    file.path(plot_dir, "Adult_gastrula_cell_type_marker_peaks_fp.pdf"),
    width = 8, height = 10
)
print(hm_pks_fp)
dev.off()
```
