---
title: "gkmSVM cell type classifiers"
execute:
  eval: false
  message: false
  warning: false
output: html_notebook
---

# Setup

```{r}
#| label: setup
#| warning: FALSE
#| message: FALSE

# load in-house functions
source("metacell_downstream_functions/Downstream_functions.R")
source("metacell_downstream_functions/utils.R")
source("motif-analysis/mta_downstream_functions.R")
source("motif-analysis/mta_gkmsvm_functions.R")
source("scripts/scatac_helper_functions.R")

# global settings
options(stringsAsFactors = FALSE)

# load packages
library(data.table)
library(stringr)
library(ggplot2)
library(patchwork)
library(BSgenome.jaNemVect1.1.DToL.Assembly)
library(GenomicRanges)
library(ArchR)
library(metacell)
library(gkmSVM)

# set ggplot2 theme
theme_py <- theme_light() + theme(
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.border = element_rect(colour = "black", fill = NA),
  text = element_text(size = 20),
  strip.placement = "outside",
  strip.text = element_text(size = 20, color = "black"),
  strip.background = element_rect(fill = "white")
)
theme_set(theme_py)
blank_theme <- theme_minimal() + theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.ticks = element_blank(),
  text = element_text(size = 20)
)

```

Data

```{r}
#| label: dirs
gastr_dir <- "ArchRProj_Nvec_gastrula"
proj_gastr <- loadArchRProject(gastr_dir)
adult_dir <- "ArchRProj_Nvec_TSS4_frag200"
proj_adult <- loadArchRProject(adult_dir)
pks_dir <- "results/Peaks"
res_dir <- "results/gkmSVM"
dir.create(res_dir, showWarnings = FALSE)
fig_dir <- "plots/gkmSVM"
dir.create(fig_dir, showWarnings = FALSE)
ann_dir <- "annotation"
```

Load gene annotations.

```{r}
# gene annotation
gnan <- fread(file.path(
  ann_dir, "Nematostella_DToL_FINAL.tsv"
))

# TF annotation
tfan <- fread(file.path(
  ann_dir, "Nematostella_DToL_TFs_FINAL.tsv"
))

# golden markers
gold <- fread(file.path(
  ann_dir, "golden-marks-231124.tsv"
), header = FALSE)
setnames(gold, c("name", "gene", "cells"))
marker_genes <- structure(
  gold$gene,
  names = gold$name
)
```

Load genome data.

```{r}
# sequence
dna_fn <- file.path("genome", "Nvec_vc1.1_gDNA.fasta")
dna <- Biostrings::readDNAStringSet(dna_fn)
```

Project-specific variables:

```{r}
ct_cols <- c(
  "cnidocyte"                  = "#ff42ff",
  "cnidocyte_gastrula"         = "#f7abf7",
  "ecto_pharynx"               = "#5bc0e8",
  "ectoderm"                   = "#51a0be",
  "ecto_aboral"                = "#045170",
  "EMS"                        = "#bdf5bd",
  "EMS_ecto_boundary"          = "#93dbce",
  "gastro_circular_muscle_1"   = "#85c90e",
  "gastro_circular_muscle_2"   = "#73b009",
  "gastro_parietal_muscle"     = "#8ceb10",
  "gastro_IRF1_2"              = "#c1eb05",
  "gastro_somatic_gonad"       = "#bde314",
  "muscle_tentacle_retractor"  = "#ffd700",
  "muscle_mesentery_retractor" = "#f0e229",
  "digestive_filaments_1"      = "#e33d3d",
  "digestive_filaments_2"      = "#d10606",
  "digestive_filaments_3"      = "#ad0303",
  "epidermis_1"                = "#04ccd4",
  "epidermis_2"                = "#16bacc",
  "precursors_PGC"             = "#bebebe",
  "precursors_endoNPC"         = "#8a8686",
  "precursors_NPC"             = "#636363",
  "NPC_1"                      = "#808d91",
  "NPC_2"                      = "#758d92",
  "neuron_GATA_Islet_1"        = "#0c82f7",
  "neuron_GATA_Islet_2"        = "#1175f0",
  "neuron_Pou4_FoxL2_1"        = "#101cde",
  "neuron_Pou4_FoxL2_2"        = "#0b16bf",
  "neuron_Pou4_FoxL2_3"        = "#2e39dd",
  "neuronal"                   = "#063cb9",
  "gland"                      = "#ff6f08",
  "gland_mucin"                = "#ff8f12"
)
cell_types <- names(ct_cols)
adult_cell_types <- c(
  "cnidocyte",
  "gastro_circular_muscle_1", 
  "gastro_circular_muscle_2",
  "gastro_parietal_muscle",
  "gastro_IRF1_2",
  "gastro_somatic_gonad",
  "muscle_mesentery_retractor",
  "muscle_tentacle_retractor",
  "digestive_filaments_1",
  "digestive_filaments_2",
  "digestive_filaments_3",
  "epidermis_1",
  "epidermis_2",
  "precursors_PGC",
  "precursors_endoNPC",
  "precursors_NPC",
  "neuron_GATA_Islet_1",
  "neuron_GATA_Islet_2",
  "neuron_Pou4_FoxL2_1",
  "neuron_Pou4_FoxL2_2",
  "neuron_Pou4_FoxL2_3",
  "gland"
)
gastr_cell_types <- c(setdiff(cell_types, adult_cell_types))
bct_cols <- toupper(c(
  "cnidocyte"                 = "#FF42FF",
  "ecto"                      = "#51a0be",
  "EMS"                       = "#bdf5bd",
  "gastro_circular_muscle"    = "#73b009",
  "gastro_parietal_muscle"    = "#8ceb10",
  "gastro"                    = "#85c90e",  
  "muscle"                    = "#FFD700",
  "digestive_filaments"       = "#e33d3d",
  "precursors"                = "#BEBEBE",
  "NPC"                       = "#808d91",
  "epidermis"                 = "#04ccd4",
  "neuron_GATA_Islet"         = "#1175f0",
  "neuron_Pou4_FoxL2"         = "#101cde",
  "neuronal"                  = "#063cb9",
  "gland"                     = "#ff6f08"
))
bct_maps <- setDT(cbind.data.frame(
  cell_type = cell_types,
  broad_cell_type = str_extract(cell_types, paste(names(bct_cols), collapse = "|"))
))
```

# gkmSVM

## Prepare data

Upon inspection of different filtering thresholds, we settled for itr `03`, i.e. relaxed filtering of peaks with Log2FC > 1 and Pval = 0.1 - this is because we want to train classifer on all accessible peaks in cell type (and not just specifically, differentially accessible peaks).

```{r}
thrs <- c(Log2FC = 1, Pval = 0.1, FDR = Inf)
lvl <- "cell_type"

for (stage_dir in c(adult_dir, gastr_dir)) {

    # all peaks
    pks_dir <- file.path(
        stage_dir, "ArchRProj", "Peaks", sprintf("%s_filtered", lvl)
    )
    pk_fn <- list.files(pks_dir, full.names = TRUE)
    pk_nm <- basename(pk_fn)
    pk_nm <- str_remove(pk_nm, "Peaks-")
    pk_nm <- str_remove(pk_nm, ".bed")
    names(pk_fn) <- pk_nm

    # diff peaks 
    pkd_dir <- file.path(
        stage_dir, "ArchRProj", "PeakDifferential", sprintf("%s_filtered", lvl)
    )
    pkd_fn <- list.files(pkd_dir, full.names = TRUE, pattern = ".tsv$")
    pk_nm <- basename(pkd_fn)
    pk_nm <- str_remove(pk_nm, "Peaks-")
    pk_nm <- str_remove(pk_nm, "-vs-others.tsv")
    names(pkd_fn) <- pk_nm

    # filter diff peaks 
    filt_dir <- file.path(stage_dir, "ArchRProj", "PeakDifferential", sprintf("%s_filtered_gkmSVM", lvl))
    dir.create(filt_dir, showWarnings = FALSE)
    for (nm in pk_nm) {

        bed_inp <- fread(pkd_fn[nm])
        bed_flt <- bed_inp[Log2FC > thrs["Log2FC"] & FDR < thrs["FDR"] & Pval < thrs["Pval"]]
        if (nrow(bed_flt) < 1e3) {
            bed_flt <- bed_inp[Log2FC > 1][order(Pval)][1:1e3, ]
            if (nrow(bed_flt) < 1e3) {
                bed_flt <- bed_inp[Log2FC > 0][order(Pval)][1:1e3, ]
                if (nrow(bed_flt) < 1e3) {
                    bed_flt <- bed_inp[order(Log2FC)][1:1e3, ]
                }
            }
        }
        bed_flt <- bed_flt[!is.na(start), 1:6]

        message(nm, ": ", length(unique(bed_flt$peak)), " peaks")

        out_fn <- basename(pkd_fn[nm])
        fwrite(bed_flt, file.path(filt_dir, out_fn), sep = "\t", col.names = FALSE)

    }
}
```

Load both all peaks (to select background) and filtered peaks (for model training).

```{r}
# all peaks
peaks_all <- rbindlist(list(
    adult = fread(file.path(
        adult_dir, "ArchRProj", "Promoters",
        "scATAC_reduced_peaks_cell_type_assignment_filtered.tsv.gz"
    )),
    gastrula = fread(file.path(
        gastr_dir, "ArchRProj", "Peaks",
        "scATAC_reduced_peaks_cell_type_assignment.tsv.gz"
    ))
), idcol = "stage", use.names=TRUE, fill = TRUE)
stopifnot(all(peaks_all$cell_type %in% cell_types))
peaks_all[, peak := paste(stage, peak, sep = "_")]
peaks_all[, seqlength := seqlengths(dna)[seqnames]]
bed_cols <- c("seqnames", "start", "end", "peak", "score", "strand")
peaks_all <- peaks_all[, ..bed_cols]
fwrite(peaks_all, file.path(res_dir, "all_peaks.tsv"), sep = "\t", col.names = TRUE, compress = "gzip")

# cell type peaks
pks_dir <- sapply(c(adult_dir, gastr_dir), function(stage_dir)
    file.path(stage_dir, "ArchRProj", "PeakDifferential", "cell_type_filtered_gkmSVM")
)
names(pks_dir) <- c("adult", "gastrula")
pk_fn <- unlist(lapply(names(pks_dir), function(x) {
    pk_fn <- list.files(pks_dir[x], full.names = TRUE)
    pk_nm <- basename(pk_fn)
    pk_nm <- str_remove(pk_nm, "Peaks-")
    pk_nm <- str_remove(pk_nm, "-vs-others.tsv")
    # pk_nm <- paste(x, pk_nm, sep = "_")
    names(pk_fn) <- pk_nm
    pk_fn
}))
pk_nm <- names(pk_fn)
stopifnot(all(pk_nm %in% cell_types))
```

Save input peaks bed files for training (70%) and testing (30%).

```{r}
# where to save inputs
dir.create(file.path(res_dir, "input"), showWarnings = FALSE)

# loop over cell types
setseed(123)
for (nm in pk_nm) {

    message(nm)
    stage <- ifelse(nm %in% adult_cell_types, "adult", "gastrula")
    bed <- fread(pk_fn[nm], select = 1:6)
    if (nrow(bed) > 0) {
        setnames(bed, c("seqnames", "start", "end", "peak", "score", "strand"))
        bed[, peak := paste(stage, peak, sep = "_")]

        # Save test peaks (~30%)
        #test_chr <- c("NC_064041.1", "NC_064042.1", "NC_064043.1")
        #test_bed <- bed[seqnames %in% test_chr]
        test_peaks <- bed[sample(seq_len(nrow(bed)), 0.3 * nrow(bed))]$peak
        test_bed <- bed[peak %in% test_peaks]
        fwrite(
            test_bed,
            file.path(res_dir, "input", sprintf("%s.test.bed", nm)),
            sep = "\t",
            col.names = FALSE
        )

        # Save train peaks (~70%)
        # train_bed <- bed[!seqnames %in% test_chr]
        train_bed <- bed[!peak %in% test_peaks]
        fwrite(
            train_bed,
            file.path(res_dir, "input", sprintf("%s.fg.bed", nm)),
            sep = "\t",
            col.names = FALSE
        )

        # Generate GC, length, and repeat matched negative set
        # and extract fasta sequence files
        gkm_mod_genNullSeqs(
            inputBedFN = file.path(res_dir, "input", sprintf("%s.fg.bed", nm)),
            genome = dna_fn,
            #select_background_from_inputBedFN = file.path(
            #    res_dir, sprintf("%s.unmatched.bg.bed", nm)
            #),
            outputBedFN = file.path(res_dir, "input", sprintf("%s.bg.bed", nm)),
            outputPosFastaFN = file.path(res_dir, "input", sprintf("%s.fg.fasta", nm)),
            outputNegFastaFN = file.path(res_dir, "input", sprintf("%s.bg.fasta", nm)),
            xfold = 1, GC_match_tol = 0.02,
            length_match_tol = 0.02, batchsize = 5000, nMaxTrials = 20
        )

        # Extract fasta sequence files for test set
        test_bed <- fread(file.path(res_dir, sprintf("%s.test.bed", nm)))
        setnames(
            test_bed,
            c("seqnames", "start", "end", "peak", "score", "strand")
        )
        ts_gr <- makeGRangesFromDataFrame(test_bed, keep.extra.columns = TRUE)
        ts_fa <- Biostrings::getSeq(dna, unique(ts_gr))
        names(ts_fa) <- unique(ts_gr)$peak
        ts_fn <- file.path(res_dir, "input", sprintf("%s.test.fasta", nm))
        Biostrings::writeXStringSet(ts_fa, ts_fn)

    }
}
```

## Train models

Train gkm-SVM using libSVM (https://github.com/kundajelab/lsgkm). Fast but cannot be used to score all possible kmers to find enriched motifs.

```{bash}
cd Scripts

mkdir -p logs_gkmsvm/train

gmkdir=results/gkmSVM/output

for ct in cnidocyte_gastrula cnidocyte digestive_filaments_1 digestive_filaments_2 digestive_filaments_3 ecto_aboral ecto_pharynx ectoderm EMS_ecto_boundary EMS epidermis_1 epidermis_2 gastro_circular_muscle_1 gastro_circular_muscle_2 gastro_IRF1_2 gastro_parietal_muscle gastro_somatic_gonad gland_mucin gland muscle_mesentery_retractor muscle_tentacle_retractor neuron_GATA_Islet_1 neuron_GATA_Islet_2 neuron_Pou4_FoxL2_1 neuron_Pou4_FoxL2_2 neuron_Pou4_FoxL2_3 neuronal NPC_1 NPC_2 precursors_endoNPC precursors_NPC precursors_PGC
do
    nohup bash gkmSVM_train.sh ${ct} ${gmkdir} > logs_gkmsvm/train_/gkmSVM_train_${ct}.out 2>&1 &
done

```

Alternatively, build models with kmers and PWMs. This approach is slower but it can be used to score kmers later.

```{bash}
cd Scripts

mkdir logs_gkmsvm/train_kmers

for ct in cnidocyte_gastrula cnidocyte digestive_filaments_1 digestive_filaments_2 digestive_filaments_3 ecto_aboral ecto_pharynx ectoderm EMS_ecto_boundary EMS epidermis_1 epidermis_2 gastro_circular_muscle_1 gastro_circular_muscle_2 gastro_IRF1_2 gastro_parietal_muscle gastro_somatic_gonad gland_mucin gland muscle_mesentery_retractor muscle_tentacle_retractor neuron_GATA_Islet_1 neuron_GATA_Islet_2 neuron_Pou4_FoxL2_1 neuron_Pou4_FoxL2_2 neuron_Pou4_FoxL2_3 neuronal NPC_1 NPC_2 precursors_endoNPC precursors_NPC precursors_PGC
do
    nohup bash gkmSVM_train_kmers.sh ${ct} > logs_gkmsvm/train_kmers/gkmSVM_train_kmers_${ct}.out 2>&1 &
done
```

## Inspect models

CV scores distribution

```{r}
out_id <- "PeakDifferential_cell_type"
itr <- "03"
res_dir <- file.path(res_dir, out_id, itr)
fig_dir <- file.path(fig_dir, out_id, itr)

nm <- "neuronal"
cvp_fn <- file.path(res_dir, sprintf("%s.cvpred.txt", nm))
cvp <- fread(cvp_fn)
setnames(cvp, c("region", "score", "label", "cvfold"))

gp_all <- ggplot(cvp, aes(score)) +
    geom_density() +
    coord_flip() +
    labs(title = "all")

gp_cvs <- ggplot(cvp, aes(score, group = cvfold)) +
    geom_density() +
    coord_flip() +
    labs(title = "CV")

ggsave(
    file.path(fig_dir, sprintf("LS-GKM-cv-scores-%s.pdf", out_id)),
    gp_all + gp_cvs,
    height = 5, width = 4
)
```

ROC for models CV

```{r}
out_id <- "PeakDifferential_cell_type"
itr <- "03"
res_dir <- file.path(res_dir, out_id, itr)
fig_dir <- file.path(fig_dir, out_id, itr)

# for LS-GKM models (CL)
pdf(
    file.path(fig_dir, sprintf("LS-GKM-cv-roc-%s.pdf", out_id)),
    height = 6, width = 5
)
roc_dt_all <- data.table()
pk_nm <- str_remove(
    list.files(res_dir, pattern = ".cvpred.txt$"),
    ".cvpred.txt"
)
for (nm in pk_nm) {
    cvp_fn <- file.path(res_dir, sprintf("%s.cvpred.txt", nm))
    tryCatch({
        cvp <- fread(cvp_fn)
        setnames(cvp, c("region", "score", "label", "cvfold"))
        message(nm)

        # number of regions used in training
        nr <- nrow(fread(file.path(
            res_dir, sprintf("%s.fg.bed", nm)
        )))

        # cross validation iterations
        cvs <- sort(unique(cvp$cvfold))

        # ROC per cross vlaidation iterations
        roc_dt <- rbindlist(sapply(cvs, function(x) {
            vpo <- cvp[cvfold == x & label == +1, score]
            vne <- cvp[cvfold == x & label == -1, score]
            roc <- pROC::roc(
                controls = vpo,
                cases = vne,
                direction = ">",
                ci = TRUE
            )
            data.table(
                specificities = roc$specificities,
                sensitivities = roc$sensitivities
            )[, auc := roc$auc][, training_peaks := nr]
        }, simplify = FALSE, USE.NAMES = TRUE), idcol = "cvfold")

        # overall ROC
        roc <- pROC::roc(
            controls = cvp[label == +1, score],
            cases = cvp[label == -1, score],
            direction = ">",
            ci = TRUE
        )

        # plot
        gp_roc <- ggplot(
                roc_dt, aes(
                    x = specificities, y = sensitivities, 
                    group = cvfold, color = str_remove(nm, "(adult|gastrula)_")
                )
            ) +
            geom_line(size = 0.5, alpha = 0.5) +
            scale_x_continuous(expand = c(0, 0)) +
            scale_y_continuous(expand = c(0, 0)) +
            scale_color_manual(values = c(ct_cols, bct_cols)) +
            coord_fixed() +
            geom_abline(slope = -1, intercept = 1, linetype = 2) +
            labs(
                x = "TNR", y = "TPR",
                title = sprintf("%s\ncross-validation", nm),
                subtitle = sprintf(
                    "AUC=%.3f [CI95=%.3f-%.3f]",
                    roc$auc, roc$ci[1], roc$ci[3]
                )
            ) +
            annotate(
                "text", x = 0.1, y = 0.1,
                label = sprintf(
                    "n = %s", format(nr, nsmall = 1, big.mark = ",")
                ),
                size = 6, hjust = 0
            ) +
            theme(legend.position = "none")
        print(gp_roc)

        roc_dt_all <- rbindlist(list(roc_dt_all, roc_dt[, model := nm]))

    }, error = function(e) NULL)
}
dev.off()

# order
roc_dt_all[, stage := str_extract(model, "(adult|gastrula)")]
roc_dt_all[, stage := factor(stage, levels = c("adult", "gastrula"))]
roc_dt_all[, cell_type := str_remove(model, "(adult|gastrula)_")]
roc_dt_all[, cell_type := factor(cell_type, levels = unique(names(ct_cols), names(bct_cols)))]
roc_dt_all[, model := factor(model, levels = unique(roc_dt_all[order(cell_type, stage)]$model))]
setorder(roc_dt_all, model)

# save
fwrite(
    roc_dt_all,
    file.path(res_dir, sprintf("%s_cv_roc.tsv.gz", out_id)),
    sep = "\t"
)

# AUC dist
gp_auc <- ggplot(
        unique(roc_dt_all[, .(model, auc, training_peaks)]),
        aes(model, auc, fill = str_remove(model, "(adult|gastrula)_"))
    ) +
    #geom_boxplot() +
    ggbeeswarm::geom_beeswarm(shape = 21) +
    # geom_text(
    #     data = unique(roc_dt_all[, .(model, training_peaks)]),
    #     aes(
    #         x = model, y = 0.95,
    #         label = format(training_peaks, nsmall = 1, big.mark = ",")
    #     ),
    #     angle = 90, size = 4
    # ) +
    guides(x.sec = ggh4x::guide_axis_manual(
        breaks = unique(roc_dt_all[, .(model, training_peaks)])$model,
        labels = format(
            unique(roc_dt_all[, .(model, training_peaks)])$training_peaks,
            nsmall = 1, big.mark = ","
        )
    )) +
    theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        panel.grid.major = element_line(linewidth = 0.2, color = "grey"),
        legend.position = "none"
    ) +
    labs(y = "AUC") +
    scale_y_continuous(limits = c(0.5, 1)) +
    scale_fill_manual(values = c(ct_cols, bct_cols)) +
    scale_color_manual(values = c(ct_cols, bct_cols))

ggsave(
    file.path(fig_dir, sprintf("LS-GKM-cv-auc-dist-%s.pdf", out_id)),
    gp_auc, width = 6, height = 6
)
```

## Predict

Prepare test fastq file with test peakset from all cell type models.

```{r}
fasta_fns <- list.files(res_dir, pattern = "test.fasta", full.names = TRUE)
fasta_seq <- Biostrings::readDNAStringSet(fasta_fns)
names(fasta_seq) <- str_remove(names(fasta_seq), "_pos_\\d+")
fasta_unx <- unique(fasta_seq)
writeXStringSet(
    fasta_unx,
    file.path(res_dir, "all.fasta")
)
```

We use `{cell_type}.model.txt` files to predict on all test sequences.

```{bash}
cd scripts

mkdir logs_gkmsvm/test

moddir="../results/gkmSVM/output"
fastafile="all.fasta"
outprefix=${fastafile%%.fasta}

for modelprefix in cnidocyte_gastrula cnidocyte digestive_filaments_1 digestive_filaments_2 digestive_filaments_3 ecto_aboral ecto_pharynx ectoderm EMS_ecto_boundary EMS epidermis_1 epidermis_2 gastro_circular_muscle_1 gastro_circular_muscle_2 gastro_IRF1_2 gastro_parietal_muscle gastro_somatic_gonad gland_mucin gland muscle_mesentery_retractor muscle_tentacle_retractor neuron_GATA_Islet_1 neuron_GATA_Islet_2 neuron_Pou4_FoxL2_1 neuron_Pou4_FoxL2_2 neuron_Pou4_FoxL2_3 neuronal NPC_1 NPC_2 precursors_endoNPC precursors_NPC precursors_PGC
do
    nohup bash gkmSVM_predict.sh ${moddir} ${modelprefix} ${fastafile} > "logs_gkmsvm/test/gkmSVM_predict_${modelprefix}_${outprefix}.out" 2>&1 &
done
```

## Inspect test results

Get all model scores in test peaks from all cell types, while keeping the metadata for peaks 
(i.e. in which cell types they were selected).

```{r}
# test peaks per cell type
fasta_fns <- list.files(file.path(res_dir, "input"), pattern = "test.fasta", full.names = TRUE)
names(fasta_fns) <- str_remove(basename(fasta_fns), "\\.test\\.fasta")
peak_dt <- rbindlist(sapply(fasta_fns, function(fasta_fn) {
    dna <- Biostrings::readDNAStringSet(fasta_fn)
    data.table(peak = names(dna))
}, USE.NAMES = TRUE, simplify = FALSE), idcol = "testset")
stopifnot(all(peak_dt$testset %in% cell_types))

# models
pk_nm <- unique(str_remove(
    list.files(file.path(res_dir, "output"), pattern = ".cvpred.txt$"),
    ".cvpred.txt"
))
stopifnot(all(pk_nm %in% cell_types))

# test scores on all peaks
test_dt <- rbindlist(sapply(pk_nm, function(model_nm) {
    tryCatch({
        out_dt <- fread(file.path(
            res_dir, "output", sprintf("%s.all.gkmpredict.txt", model_nm)
        ))
        message("\nModel: ", model_nm)
        setnames(out_dt, c("peak", "score"))
        out_dt[, peak := str_remove(peak, "_pos_\\d+")]
    }, error = function(e) NULL)
}, USE.NAMES = TRUE, simplify = FALSE), idcol = "model")
test_dt <- test_dt[!is.na(peak)]

# we will remove peaks that model saw in training
train_dt <- rbindlist(sapply(pk_nm, function(model_nm) {
    tryCatch({
        message("\nModel: ", model_nm)
        fread(file.path(
            res_dir, "input", sprintf("%s.fg.bed", model_nm)
        ))
    }, error = function(e) NULL)
}, USE.NAMES = TRUE, simplify = FALSE), idcol = "model")
setnames(train_dt, c("model", "seqnames", "start", "end", "peak", "score", "strand"))

mean(peak_dt$peak %in% test_dt$peak)
mean(test_dt$peak %in% peak_dt$peak)
mean(test_dt$peak %in% train_dt$peak)

# merge models scores on all peaks with 
# peaks test metadata (i.e. which cell type testset peaks are in)
test_dt <- merge.data.table(
    test_dt, peak_dt, by = "peak",
    allow.cartesian = TRUE, sort = FALSE
)
# inverse merge scores in test peaks with training peaks
# i.e. remove scores in peaks that model saw in training
test_dt <- merge.data.table(
    test_dt, train_dt[, .(trainset=model, peak)], by = "peak",
    allow.cartesian = TRUE, all.x = TRUE, sort = FALSE
)
test_dt[model != trainset, trainset := NA]
test_dt <- unique(test_dt[is.na(trainset)][, trainset := NULL])
setcolorder(test_dt, c("model","peak","score","testset"))
stopifnot(!any(test_dt[, .N,.(model,peak,testset)]$N>1))

# order
test_dt[, score := as.numeric(score)]
stopifnot(all(test_dt$model %in% cell_types))
test_dt[, model := factor(model, levels = cell_types)]
stopifnot(all(test_dt$testset %in% cell_types))
test_dt[, testset := factor(testset,levels = cell_types)]
setorder(test_dt, model, testset, -score)

# save
fwrite(
    test_dt,
    file.path(res_dir, "models.test.scores.tsv.gz"),
    sep = "\t", compress = "gzip"
)

```

Plot ROC and AUC

```{r}
# test results
test_dt <- fread(file.path(res_dir, "models.test.scores.tsv.gz"))

# test cell types
cts <- as.character(unique(test_dt$testset))

# models
mds <- as.character(unique(test_dt$model))

# ROC per model and per cell type
pdf(
    file.path(fig_dir, sprintf("LS-GKM-test-roc.pdf")),
    height = 6, width = 5
)
roc_dt <- rbindlist(sapply(mds, function(nm) {

    message(nm)
    cvp <- test_dt[model == nm]

    # ROC
    roc_dt <- rbindlist(sapply(cts, function(x) {
        cfg <- cvp[testset == x]
        cbg <- cvp[!peak %in% cfg$peak]
        vpo <- cfg$score
        vne <- cbg$score
        pnr <- length(unique(cfg$peak))
        nnr <- length(unique(cbg$peak))
        roc <- pROC::roc(
            controls = vpo,
            cases = vne,
            direction = ">",
            ci = TRUE
        )
        data.table(
            specificities = roc$specificities,
            sensitivities = roc$sensitivities
        )[, auc := roc$auc][, ":="(test_peaks_pos = pnr, test_peaks_neg = nnr)]
    }, simplify = FALSE, USE.NAMES = TRUE), idcol = "testset", use.names = TRUE, fill = TRUE)

    # plot
    gp_roc <- ggplot(
            roc_dt, aes(
                x = specificities, y = sensitivities, 
                group = testset,
                color = str_remove(testset, "(adult|gastrula)_"),
                label = sprintf("%s (AUC=%.2f)", testset, auc)
            )
        ) +
        geom_line(size = 0.5, alpha = 0.5) +
        scale_y_continuous(expand = c(0, 0)) +
        scale_x_continuous(expand = expansion(mult = c(0, 0.45))) +
        scale_color_manual(values = c(ct_cols, bct_cols)) +
        geom_text_repel(
            data = roc_dt[order(-auc, -specificities * sensitivities)][, .SD[1], testset][1:4],
            nudge_x = 0.1, direction = "y", hjust = "left"
        ) +
        coord_fixed() +
        geom_abline(slope = -1, intercept = 1, linetype = 2) +
        labs(
            x = "TNR", y = "TPR",
            title = sprintf("%s\ntest ROC", nm)
        ) +
        theme(legend.position = "none")
    print(gp_roc)
    
    roc_dt

}, USE.NAMES = TRUE, simplify = FALSE), idcol = "model", use.names = TRUE, fill = TRUE)
dev.off()

roc_dt[, testset := factor(testset, levels = cell_types)]
roc_dt[, model := factor(model, levels = cell_types)]
setorder(roc_dt, model, testset)

# save
fwrite(
    roc_dt,
    file.path(res_dir, "models.roc.tsv.gz"),
    sep = "\t", compress = "gzip"
)

# plor AUC for models
gp_auc <- ggplot(
        unique(roc_dt[, .(model, testset, auc)]),
        aes(
            model, auc, fill = testset, color = testset
        )
    ) +
    ggbeeswarm::geom_beeswarm(shape = 21, size = 3) +
    scale_y_continuous(limits = c(0.5, 1)) +
    scale_fill_manual(values = c(ct_cols, bct_cols)) +
    scale_color_manual(values = structure(
      colorspace::darken(c(ct_cols, bct_cols), 0.5),
        names = c(names(ct_cols), names(bct_cols))
    )) +
    theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        panel.grid.major = element_line(linewidth = 0.2, color = "grey")
    ) +
    guides(
        shape = guide_legend(override.aes = list(size = 5, fill = "grey")),
        fill = "none", color = "none"
    ) +
    labs(x = "", y = "AUC") 

# plot number of peaks
gp_pks <- ggplot(
        unique(roc_dt[, .(model, testset, test_peaks_pos)]),
        aes(
            model, test_peaks_pos,
            fill = testset, color = testset
        )
    ) +
    geom_jitter(shape = 21, size = 3) +
    scale_y_continuous(limits = c(0, NA), expand = expansion(c(0, 0.1))) +
    scale_fill_manual(values = c(ct_cols, bct_cols)) +
    scale_color_manual(values = structure(
        colorspace::darken(c(ct_cols, bct_cols), 0.5),
        names = c(names(ct_cols), names(bct_cols))
    )) +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major = element_line(linewidth = 0.2, color = "grey")
    ) +
    guides(
        shape = guide_legend(override.aes = list(size = 5, fill = "grey")),
        fill = "none", color = "none"
    ) +
    labs(x = "", y = "test peaks")

gp_gkm_auc <- gp_pks / gp_auc + plot_layout(guides = "collect")
ggsave(
    file.path(fig_dir, "LS-GKM-test-auc.pdf"),
    gp_gkm_auc, height = 8, width = 10
)
```

Cell type similarities based on AUC

```{r}
roc_dt <- fread(file.path(res_dir, "models.roc.tsv.gz"))
auc_dt <- unique(roc_dt[, .(model, testset, auc)])
auc_dt[, model := factor(model, levels = cell_types)]
auc_dt[, testset := factor(testset, levels = cell_types)]
setorder(auc_dt, model, testset)
auc_dc <- dcast.data.table(auc_dt, model ~ testset, value.var = "auc")
auc_mt <- as.matrix(auc_dc[, -1])
rownames(auc_mt) <- auc_dc[[1]]
auc_mt <- auc_mt[rownames(auc_mt), rownames(auc_mt)]
all.equal(rownames(auc_mt), colnames(auc_mt))
```

Cell type tree based on AUC

```{r}
# NJ clustering
require(ape)
require(ggtree)
tree <- nj(dist(auc_mt))
tree_nm <- sprintf("LS-GKM-test-auc-NJ-adult-gastrula-tree")
tree_fn <- file.path(res_dir, paste0(tree_nm, ".txt"))

# tree visuals
gt <- ggtree(tree, aes(color = label), ladderize = FALSE) +
  scale_color_manual(values = ct_cols) +
  theme_tree2() +
  geom_tiplab() +
  scale_x_continuous(expand = expansion(mult = c(0, 1))) +
  theme(legend.position = "none")
ggsave(file.path(
    fig_dir,
    paste0(tree_nm, ".pdf")
), gt, height = 6, width = 7)
```

Tree with bootstrap in `90_Figures.qmd`.

Order AUC matrix by tree.

```{r}
tips <- tree$tip.label
hc_mt <- auc_mt[tips, tips]
diag(hc_mt) <- NA

row_ann <- HeatmapAnnotation(
  which = "row", 
  "cell type" = rownames(hc_mt), 
  col = list("cell type" = c(ct_cols, bct_cols))
)
col_ann <- HeatmapAnnotation(
  which = "column",
  "cell type" = colnames(hc_mt), 
  col = list("cell type" = c(ct_cols, bct_cols)), show_legend = FALSE
)
col_vec <- viridis::plasma(10)
col_fun <- circlize::colorRamp2(seq(0, 1, length.out = length(col_vec)), col_vec)
hm_cor <- Heatmap(
  hc_mt, name = "AUC",
  col = col_vec,
  border = TRUE, use_raster = TRUE,
  cluster_rows = FALSE, cluster_columns = FALSE, 
  show_row_dend = FALSE, show_column_dend = FALSE,
  show_row_names = TRUE, show_column_names = FALSE, 
  top_annotation = col_ann, left_annotation = row_ann
)
pdf(file.path(
    fig_dir,
    sprintf("LS-GKM-test-auc-heatmap_%s.pdf", out_id)
), width = 10.5, height = 6)
print(hm_cor)
dev.off()
```

## Explain

Top scoring sequences for each model

```{r}
out_id <- "PeakDifferential_cell_type"
itr <- "03"
res_dir <- file.path(res_dir, out_id, itr)
fig_dir <- file.path(fig_dir, out_id, itr)

test_dt <- fread(file.path(res_dir, "models.test.scores.tsv"))
test_fq <- readDNAStringSet(file.path(res_dir, "all.fasta"))

# plot scores distributions
setorder(test_dt, model, -score)
test_dt[, rank := 1:.N, model]
pdf(
    file.path(fig_dir, "gkmexplain_input_peak_model_scores.pdf"),
    gpe, width = 8, height = 8
)
for (mn in unique(test_dt$model)) {
  message(mn)
  gpe <- ggplot(test_dt[model == mn], aes(
          rank, score, group = model
      )) +
      geom_line() +
      geom_rug(aes(color = test_cell_type)) +
      scale_color_manual(values = ct_cols) +
      scale_fill_manual(values = structure(
          colorspace::lighten(ct_cols, 0.8),
          names = names(ct_cols)
      )) +
      theme(
        panel.grid.major = element_line(color = "grey60", linewidth = 0.25),
        panel.grid.minor = element_line(color = "grey80", linewidth = 0.05),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "none"
      ) +
    labs(title = sprintf("model: %s", mn))
  print(gpe)
}
dev.off()

# get peaks with best scores per model
setorder(test_dt, model, -score)
best_dt <- test_dt[, .SD[score > quantile(.SD$score, 0.95)], model]
best_dt <- test_dt[, unique(.SD[, .(model, peak, score)])[1:4000], model]
best_dt <- test_dt[score > 0]
best_dt[, .N, model]

# plot best scores distributions
gpe <- ggplot(best_dt, aes(model)) +
    geom_bar(
        aes(color = model_cell_type, fill = model_cell_type)
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    scale_fill_manual(values = ct_cols) +
    scale_color_manual(values = structure(
        colorspace::darken(ct_cols, 0.5),
        names = names(ct_cols)
    )) +
    theme(
      panel.grid.major = element_line(color = "grey60", linewidth = 0.25),
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      legend.position = "none"
    ) +
  labs(x = "model", y = "top scored peaks")
ggsave(
    file.path(fig_dir, "gkmexplain_peaks.pdf"),
    gpe, width = 10, height = 6
)

# save sequences
for (x in unique(best_dt$model)) {
    expl_fq <- test_fq[best_dt[model == x]$peak]
    writeXStringSet(
        expl_fq,
        file.path(res_dir, sprintf("%s.explain.fasta", x))
    )
}
```

```{bash}
cd Scripts
itr="03"
mkdir logs_gkmsvm/explain_${itr}
wdir="/home/anamaria/cluster/aelek/proj/scATAC_nvec_v2/clustering/SEACells/Results/ArchRProj_Nvec_TSS4_frag200/ArchRProj/gkmSVM/PeakDifferential_cell_type"
moddir=${wdir}/${itr} 
modelprefix="adult_cnidocyte"
for modelprefix in adult_digestive_filaments_1 adult_digestive_filaments_2 adult_digestive_filaments_3 adult_epidermis_1 adult_epidermis_2 adult_gastro_circular_muscle_1 adult_gastro_circular_muscle_2 adult_gastro_parietal_muscle adult_gastro_unknown_1 adult_gastro_unknown_2 adult_gland adult_muscle_1 adult_muscle_2 adult_neuron_GATA_Islet_1 adult_neuron_GATA_Islet_2 adult_neuron_GATA_Islet_3 adult_neuron_Pou4_FoxL2_1 adult_neuron_Pou4_FoxL2_2 adult_precursors_1 adult_precursors_2 adult_precursors_3 cnidocyte cnidocyte_precursors ectoderm_embryonic ectoderm_embryonic_aboral ectoderm_embryonic_oral gland mesendoderm_ectoderm mesendoderm_embryonic neuronal NPC
do
    fastafile=${modelprefix}".explain.fasta"
    fastafile="all.fasta"
    nohup bash 05_gkmSVM_explain.sh ${moddir} ${modelprefix} ${fastafile} > "logs_gkmsvm/explain_"${itr}"/02_gkmSVM_explain_PeakDifferential_cell_type_${modelprefix}.out" 2>&1 &
done

# create dnshuff.fasta file in 08_gkmSVM.ipynb

for modelprefix in adult_cnidocyte 
do
    fastafile=${modelprefix}".dnshuff.fasta"
    nohup bash 05_gkmSVM_explain_dnshuff.sh ${moddir} ${modelprefix} ${fastafile} > "logs_gkmsvm/explain_"${itr}"/02_gkmSVM_explain_dnshuff_PeakDifferential_cell_type_${modelprefix}.out" 2>&1 &
done
```

### Inspect explanantions

Load importance scores

```{r}
out_id <- "PeakDifferential_cell_type"
itr <- "03"
res_dir <- file.path(res_dir, out_id, itr)
fig_dir <- file.path(fig_dir, out_id, itr)

sct <- "adult_cnidocyte"
imp_score_dir <- file.path(res_dir, sprintf("%s_norm_impscores", sct))

# load normalized importance scores
imp_score_fns <- list.files(imp_score_dir, pattern = "norm_impscores", full.names = TRUE)
names(imp_score_fns) <- str_remove(str_remove(basename(imp_score_fns), ".txt"), "norm_impscores_")
imp_dt <- rbindlist(sapply(
    names(imp_score_fns),
    function(x) fread(imp_score_fns[x]),
    simplify = FALSE, USE.NAMES = TRUE
), idcol = "peak")
setnames(imp_dt, paste0("V", 1:4), c("A", "C", "G", "T"))


# load normalized hypothetical importance scores
imp_hyp_score_fns <- list.files(imp_score_dir, pattern = "norm_hyp_impscores", full.names = TRUE)
names(imp_hyp_score_fns) <- str_remove(str_remove(basename(imp_hyp_score_fns), ".txt"), "norm_hyp_impscores_")
imp_hyp_dt <- rbindlist(sapply(
    names(imp_hyp_score_fns),
    function(x) fread(imp_hyp_score_fns[x]),
    simplify = FALSE, USE.NAMES = TRUE
), idcol = "peak")
setnames(imp_hyp_dt, paste0("V", 1:4), c("A", "C", "G", "T"))

```

Plot importance scores

```{r}
require(ggseqlogo)

# plot normalized importance scores
pdf(
    file.path(fig_dir, sprintf("gkm_explain_score_%s.pdf", x)),
    width = 14, height = 2
)
for (i in unique(imp_dt$peak)) {
    imp <- imp_dt[peak == i][, peak := NULL]
    sco <- sum(apply(imp, 1, sum))
    pwm <- t(as.matrix(imp))
    rownames(pwm) <- c("A", "C", "G", "T")

    # plot
    gg_pwm <- ggseqlogo(pwm, method = "custom", seq_type = "dna") +
        theme(
            axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 5)
        ) +
        labs(title = sprintf(
            "peak idx %s (score: %.4f)", i, sco
        ))
    print(gg_pwm)

}
dev.off()

# plot normalized hypothetical importance scores
pdf(
    file.path(fig_dir, sprintf("gkm_explain_hyp_score_%s.pdf", sct)),
    width = 36, height = 3
)
for (i in unique(imp_hyp_dt$peak)) {
    imp <- imp_hyp_dt[peak == i][, peak := NULL]
    pwm <- t(as.matrix(imp))
    rownames(pwm) <- c("A", "C", "G", "T")

    # plot
    gg_pwm <- ggseqlogo(pwm, method = "custom", seq_type = "dna") +
        theme(
            axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
        )
    print(gg_pwm)

}
dev.off()
```

Do importance scores drop?

```{r}
imp_sco_dt <- rbindlist(lapply(unique(imp_dt$peak), function(i) {
    imp <- imp_dt[peak == i][, peak := NULL]
    sco <- sum(apply(imp, 1, sum))
    data.table(peak = i, score = sco)
}))
imp_sco_dt[, peak := as.integer(peak)]
gpe <- ggplot(imp_sco_dt, aes(
        peak, score
    )) +
    geom_line() +
    theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "none"
    )
ggsave(
    file.path(fig_dir, sprintf("gkmexplain_imp_scores_%s.pdf", sct)),
    gpe, width = 10, height = 6
)

```

Search for motif hits in high importance scores subsequences

```{r}
# convert values below 0.1 to N
seq_dt <- copy(imp_dt)
seq_dt[A < 0.01, A := 0]
seq_dt[C < 0.01, C := 0]
seq_dt[G < 0.01, G := 0]
seq_dt[T < 0.01, T := 0]

# plot "filtered" importance scores
pdf(
    file.path(fig_dir, sprintf("gkm_explain_filt_score_%s.pdf", sct)),
    width = 36, height = 3
)
for (i in unique(seq_dt$peak)) {
    imp <- seq_dt[peak == i][, peak := NULL]
    sco <- sum(apply(imp, 1, sum))
    pwm <- t(as.matrix(imp))
    rownames(pwm) <- c("A", "C", "G", "T")

    # plot
    gg_pwm <- ggseqlogo(pwm, method = "custom", seq_type = "dna") +
        theme(
            axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
        ) +
        labs(title = sprintf(
            "peak idx %s (score: %.4f)", i, sco
        ))
    print(gg_pwm)

}
dev.off()

# get sequence for filtered importance scores
nucs <- apply(seq_dt[ ,-1], 1, function(x) {
    if (any(x > 0)) {
        c("A", "C", "G", "T", "N")[which(x > 0)]
    } else {
        "N"
    }
})
seq_dt[, nuc := nucs]

# get chunks of important nucleotides
seq_dt[, pos := 1:.N, peak][nuc == "N", pos := 0]
seq_dt <- seq_dt[pos != 0]
for(pk in unique(seq_dt$peak)) {
  seq_pk <- seq_dt[peak == pk] 
  for (i in seq_len(nrow(seq_pk))) {
    ps <- seq_pk[i]$pos
    if (i > 1) {
      if (ps - seq_pk[i-1]$pos > 1) region <- region + 1
    } else {
      region <- 1
    }
    seq_dt[peak == pk & pos == ps, chunk := region]
  }
}

# filter out small chunks
seq_dt[, chunk_length := nrow(.SD), .(peak, chunk)]
seq_dt <- seq_dt[chunk_length > 4]

# sequences
seq_dt[, peak_chunk := paste(peak, chunk, sep = "_")]
seq_st <- sapply(unique(seq_dt$peak_chunk), function(x) {
    sstr <- paste0(seq_dt[peak_chunk == x]$nuc, collapse = "")
    if (nchar(sstr) < 40) {
      w <- (40 - nchar(sstr)) / 2
      ns <- paste0(rep("N", w), collapse = "")
      sstr <- paste0(ns, sstr, ns)
    }
    sstr
})
seq_str <- DNAStringSet(seq_st)

# among motifs that were assigned to genes with chromVAR
# select best motif for each gene based on correlation
exp_chromvar_dt <- fread(
    file.path(
        adult_dir, "insilicoChIP", "metacell", "03",
        "gene_expression_fc_chromVAR_genes_exp_FC2_acc_FC4_spearman.tsv.gz"
    )
)
exp_chromvar_dt[, cor := cor(.SD$expression, .SD$zscore), .(
    gene, og, pfam, motif, id
)]
exp_chromvar_cor_dt <- unique(
    exp_chromvar_dt[, .(gene, og, pfam, id, cor)]
)
exp_chromvar_cor_top <- exp_chromvar_cor_dt[
    order(cor, decreasing = TRUE)
][, .SD[1], gene]
tf_dt <- exp_chromvar_dt[id %in% exp_chromvar_cor_top$id]
tf_dt <- unique(tf_dt[, .(gene, motif)])

# match motifs
mots_lst <- readRDS(file.path(adult_dir, "Archetypes", "motif-archetypes-PPM-PCCnorm-0.8-IC0.5-8bp-pwms.rds")) # 1649
mots_lst <- mots_mta_lst[names(mots_mta_lst) %in% tf_dt$motif] # 276
mot_pwm  <- universalmotif::convert_type(mots_lst, "PWM", pseudocount = 1)
mot_pwme <- universalmotif::convert_motifs(mot_pwm, "PWMEnrich-PWM")
mot_lens <- sapply(mot_pwme, length, USE.NAMES=TRUE)

# scanning archetypes
scan_dt <- rbindlist(lapply(seq_along(seq_str), function(i) {
    
    seq <- seq_str[[i]]
    sn <- names(seq_str)[i]
    message(sprintf("%s (%s/%s)", sn, i, length(seq_str)))
    seq_dt <- rbindlist(lapply(seq_along(mot_pwme), function(j) {

        mot <- mot_pwme[[j]]
        mn <- names(mot_pwme)[j]
        gw_scan_c <- PWMEnrich::scanWithPWM(
            mot, seq, 
            both.strands = FALSE, strand.fun = "max"
        )
        gw_scan_dt <- data.table(motif_score = gw_scan_c)
        gw_scan_dt[, motif := mn][, position := 1:.N]
        gw_scan_dt
    }))
    seq_dt[, peak := sn]
    setcolorder(seq_dt, c("peak", "motif", "position", "motif_score"))
    seq_dt

}))
scan_dt[is.na(motif_score), motif_score := 0]
scan_dt <- scan_dt[motif_score > 0]

setnames(scan_dt, "peak", "peak_chunk")
scan_dt[, peak := str_remove(peak_chunk, "_\\d+")]


# nomr by motif length
scan_dt[, motif_len := mot_lens[motif]]
scan_dt[, motif_score_norm := motif_score / motif_len ]

# distribution of motf scores
gp_mot_sco <- ggplot(scan_dt[motif_score_norm > 0], aes(motif_score_norm, color = peak)) + 
    geom_density() +
    theme(legend.position = "none")
ggsave(
    file.path(fig_dir, sprintf("imp_scores_mot_score_%s.pdf", sct)),
    width = 6, height = 5
)

# add gene info
scan_dt <- merge.data.table(scan_dt, tf_dt, by = "motif", all.x = TRUE, sort = FALSE, allow.cartesian = TRUE)
scan_dt <- merge.data.table(scan_dt, tfan[,.SD[1], gene], by = "gene", all.x = TRUE, sort = FALSE, allow.cartesian = FALSE)
scan_dt <- merge.data.table(scan_dt, marks, by = "gene", all.x = TRUE, sort = FALSE, allow.cartesian = FALSE)
scan_dt[!is.na(marker_name), og := marker_name]
scan_dt[, arch := str_extract(motif, "ARCH\\d+")]
scan_dt[, label := sprintf(
    "%s | %s| %s", arch, str_remove(gene, "Nvec_((vc1.1_)|(v1g)_)*"), substr(og, 1, 30)
)]

# save
fwrite(
    scan_dt,
    file.path(res_dir, sprintf("motif_scores_imp_scores_%s.tsv.gz", sct)),
    sep = "\t"
)
```

Inspect hits for all TFs

```{r}
scan_dt <- fread(
    file.path(res_dir, sprintf("motif_scores_imp_scores_%s.tsv.gz", sct)),
)

# take top motifs per peak
top_dt <- scan_dt[motif_score_norm > 0.5]

# take one entry per motif
top_dt <- top_dt[order(-motif_score)][, .SD[1], .(arch, peak, position)]

# make a table
top_dc <- dcast.data.table(top_dt, peak ~ label, value.var = "motif_score_norm", fun.aggregate = length)
top_mt <- t(as.matrix(top_dc[, -1]))
top_mt <- pmin(top_mt, 10)
colnames(top_mt) <- top_dc[[1]]

# cluster motifs
hc <- hclust(dist(top_mt, method = "maximum"), method = "ward.D2")
top_mt <- top_mt[hc$order, ]

# cluster peaks
hc <- hclust(dist(t(top_mt), method = "maximum"), method = "ward.D2")
top_mt <- top_mt[, hc$order]

# plot
# RColorBrewer::brewer.pal(8, "Blues")[-1]
hm_col <- c("white", rev(viridis::viridis(10)))
top_hm <- Heatmap(
    top_mt, name = "motif hits", col = hm_col,
    border = TRUE,
    rect_gp = gpar(col = "white", lwd = 0.1),
    show_column_dend = FALSE, show_row_dend = FALSE,
    cluster_columns = FALSE, cluster_rows = FALSE,
    column_names_gp = gpar(fontsize = 5),
    row_names_gp = gpar(fontsize = 5)
)
pdf(
    file.path(fig_dir, sprintf("imp_scores_mot_score_all_heatmap_%s.pdf", sct)),
    width = 14, height = 16
)
print(top_hm)
dev.off()

```

Hits for motifs of TFs that are in GRN

```{r}
# subset for TFs active in cell type
grn_ct_dt <- fread(file.path(
    adult_dir, "insilicoChIP", "metacell", "03", sprintf(
        "grn_tf_target_genes_expression_%s_%s_accessibility_%s_%s_chromvar_%s.tsv.gz",
        "fc", "1.4", "access", "0.4", "4")
))
ct_st <- paste(
    str_remove(sct, "(adult|gastrula)_"),
    str_extract(sct, "(adult|gastrula)"),
    sep = "_"
)
ct_st_grn <- grn_ct_dt[cell_type_stage == ct_st]
top_dt <- scan_dt[gene %in% ct_st_grn$gene]

# take top motifs per peak
top_dt <- top_dt[motif_score_norm > 0]
# top_dt <- scan_dt[order(-motif_score)][, .SD[1:10], motif]
top_dt <- top_dt[motif_score_norm > 0.5]

# take only peaks with good overall importance score
top_dt <- top_dt[peak %in% imp_sco_dt[score > 1.2]$peak]

# one entry per motif per chunk
top_dt <- top_dt[order(-motif_score)][, .SD[1], .(motif, gene, peak, peak_chunk)]

# take one entry per motif (multiple genes)
top_dt <- top_dt[order(marker_name == "", -motif_score)][, .SD[1], .(arch, peak, peak_chunk, position)]

# TBA do this systematically, for now manual inspection of other genes
# top_dt[, annot_genes := paste(
#   substr(og, 1, 10), str_remove(gene, "Nvec_((vc1.1_)|(v1g)_)*"), collapse = " "
# ), gene]
# top_dt[, arch_genes := paste(
#   annot_genes, sep = " | "
# ), arch]
# top_dt[, label_unique := sprintf(
#     "%s | %s", arch, arch_genes
# )]

# make a table
maxs <- function(x) max(x[!is.infinite(x)], na.rm = TRUE)
top_dc <- dcast.data.table(top_dt, peak ~ label, value.var = "motif_score_norm", fun.aggregate = length, fill = 0)
top_mt <- t(as.matrix(top_dc[, -1]))
top_mt <- pmin(top_mt, 10)
colnames(top_mt) <- top_dc[[1]]

# cluster motifs
hc <- hclust(dist(top_mt, method = "euclidean"), method = "ward.D2")
top_mt <- top_mt[hc$order, ]

# cluster peaks
hc <- hclust(dist(t(top_mt), method = "euclidean"), method = "ward.D2")
top_mt <- top_mt[, hc$order]

# plot
# RColorBrewer::brewer.pal(8, "Blues")[-1]
hm_col <- c("white", rev(viridis::viridis(10)))
top_hm <- Heatmap(
    top_mt, name = "motif hits", col = hm_col,
    border = TRUE,
    rect_gp = gpar(col = "white", lwd = 0.1),
    show_column_dend = FALSE, show_row_dend = FALSE,
    cluster_columns = FALSE, cluster_rows = FALSE,
    column_names_gp = gpar(fontsize = 5),
    row_names_gp = gpar(fontsize = 5)
)
pdf(
    file.path(fig_dir, sprintf("imp_scores_mot_score_grn_heatmap_%s.pdf", sct)),
    width = 8, height = 1.2
)
print(top_hm)
dev.off()
```

Quantify

```{r}
md <- rbindlist(lapply(colnames(top_mt), function(x) {
  data.table(code=paste0(top_mt[,x], collapse = ""))[, peak := x]
}))
md[,occur := .N, code]
md_top <- md[occur > 1][order(-occur)]
top_pks <- md[order(-occur)]$peak
tot <- length(unique(md$peak))

# plot normalized importance scores for top peaks
require(ggseqlogo)
pdf(
    file.path(fig_dir, sprintf("gkm_explain_score_top_%s.pdf", x)),
    width = 14, height = 2
)
for (i in top_pks) {
    imp <- imp_dt[peak == i][, peak := NULL]
    sco <- sum(apply(imp, 1, sum))
    occ <- md[peak==i]$occur[1]
    pwm <- t(as.matrix(imp))
    rownames(pwm) <- c("A", "C", "G", "T")

    gg_pwm <- ggseqlogo(pwm, method = "custom", seq_type = "dna") +
        theme(
            axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 5)
        ) +
        labs(title = sprintf(
            "peak idx %s (score: %.4f, occurrences: %s/%s, %.2f%%)", i, sco, occ, tot, occ/tot*100
        ))
    print(gg_pwm)

}
dev.off()

# get motifs for each peak
md_str <- md_top$code[1]
md_spl <- strsplit(md_str, "")[[1]]
md_arc <- rownames(top_mt)[which(md_spl>0)]
md_rep <- md_spl[md_spl>0]
md_mot <- unname(unlist(sapply(seq_along(md_arc), function(i) rep(md_arc[i], md_rep[i]))))
```

```{r}
# exp_fns <- list.files(res_dir, pattern = "explain.hyp.txt")
# names(exp_fns) <- str_remove(exp_fns, ".explain.hyp.txt")
#
# exp_1 <- fread(
#     file.path(res_dir, exp_fns[x]),
#     header = FALSE, sep = "\t", select = 1:2
# )
# setnames(exp_1, c("region id", "score"))
# exp_2 <- fread(
#     file.path(res_dir, exp_fns[x]),
#     header = FALSE, sep = ";", fill = TRUE
# )
# exp_2 <- unique(exp_2)
# exp_2[, peak := str_extract(V1, "((adult|gastrula)_)peak\\d+")]
# exp_2[, V1 := str_remove(V1, "((adult|gastrula)_)peak\\d+\\t")]
# exp_2[, score := str_extract(V1, "\\d+\\.\\d+")]
# exp_2[, score := as.numeric(score)]
# exp_2[, V1 := str_remove(V1, "\\d+\\.\\d+\\t")]
# setcolorder(exp_2, c("peak", "score"))
# setorder(exp_2, -score)
#
# require(ggseqlogo)
#
# lapply(
#     seq_len(nrow(exp_2)),
#     function(i) {
#         pk <- exp_2[i, ]$peak
#         mt_dt <- melt.data.table(exp_2[i, -c(1, 2)], measure.vars = colnames(exp_2)[-c(1:2)])[, variable := NULL]
#         mt_dt <- tidyr::separate(mt_dt, col = "value", into = c("A", "C", "G", "T"), sep = ",")
#         setDT(mt_dt)
#         mt_dt <- mt_dt[, lapply(.SD, as.numeric), ,.SDcols = colnames(mt_dt)]
        
#         #normalize the hyp scores such that, at each position, hypothetical importance
#         # scores that have the same sign as the original importance score all sum
#         # up to the original importance score value.

#         pwm <- t(as.matrix(mt_dt))
#         rownames(pwm) <- c("A", "C", "G", "T")

#         # plot
#         gg_pwm <- ggseqlogo(pwm, method = "prob") +
#             theme(
#                 axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
#             )
#         ggsave(
#             file.path(fig_dir, "gkm_explain_hyp_score.pdf"),
#             gg_pwm, width = 28, height = 3
#         )

#         # # save matrix
#         # write_matrix(t(pwm))
#     }
# )

```


## Compare predictions with models built using different peaksets

Compare models built using different iterations of filtering cell type specific peaks

```{r}
itrs <- c("00", "01", "02", "03", "04")
roc_dt_all <- rbindlist(sapply(itrs, function(itr) {
    
    message(itr)

    # load data
    itr_dir <- file.path(res_dir, out_id, itr)
    roc_dt <- fread(
        file.path(itr_dir, "models.roc.tsv.gz"),
        sep = "\t"
    )
    roc_dt
    
}, USE.NAMES = TRUE, simplify = FALSE), idcol = "itr", use.names = TRUE, fill = TRUE)

roc_dt <- unique(roc_dt_all[, .(itr, model, testset, auc, test_peaks_pos, test_peaks_neg, test_cell_type, stage)])
roc_dt[, filt := filts[itr]]
roc_dt[, filt := factor(filt, levels = filts)]
roc_dt[, stage := factor(stage, levels = c("adult", "gastrula"))]
roc_dt[, test_cell_type := factor(test_cell_type, levels = unique(names(ct_cols), names(bct_cols)))]
roc_dt[, model := factor(model, levels = unique(roc_dt[order(test_cell_type)]$model))]
setorder(roc_dt, itr, model)

# plor AUC for CVs
gp_auc <- ggplot(
        unique(roc_dt[, .(model, test_cell_type, filt, auc)]),
        aes(
            model, auc, shape = filt, fill = test_cell_type, color = test_cell_type
        )
    ) +
    ggbeeswarm::geom_beeswarm() +
    scale_y_continuous(limits = c(0.5, 1)) +
    scale_shape_manual(values = structure(
        rep(21:25, 2)[seq_along(itrs)],
        names = filts[itrs]
    )) +
    scale_fill_manual(values = c(ct_cols, bct_cols)) +
    scale_color_manual(values = structure(
        colorspace::darken(c(ct_cols, bct_cols), 0.5),
        names = c(names(ct_cols), names(bct_cols))
    )) +
    theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        panel.grid.major = element_line(linewidth = 0.2, color = "grey")
    ) +
    guides(
        shape = guide_legend(override.aes = list(size = 5, fill = "grey")),
        fill = "none", color = "none"
    ) +
    labs(x = "", y = "test AUC") 

# plot number of peaks
gp_pks <- ggplot(
        unique(roc_dt[, .(filt, model, testset, test_cell_type, test_peaks_pos)][model == testset]),
        aes(
            model, test_peaks_pos, shape = filt,
            fill = test_cell_type, color = test_cell_type
        )
    ) +
    ggbeeswarm::geom_beeswarm(size = 3) +
    scale_shape_manual(values = structure(
        rep(21:25, 2)[seq_along(itrs)],
        names = filts[itrs]
    )) +
    scale_y_continuous(limits = c(0, NA)) +
    scale_fill_manual(values = c(ct_cols, bct_cols)) +
    scale_color_manual(values = structure(
        colorspace::darken(c(ct_cols, bct_cols), 0.5),
        names = c(names(ct_cols), names(bct_cols))
    )) +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major = element_line(linewidth = 0.2, color = "grey")
    ) +
    guides(
        shape = guide_legend(override.aes = list(size = 5, fill = "grey")),
        fill = "none", color = "none"
    ) +
    labs(x = "", y = "test peaks")

gp_gkm_auc <- gp_pks / gp_auc + plot_layout(guides = "collect")
ggsave(
    file.path(dirname(fig_dir), "LS-GKM-test-roc.pdf"),
    gp_gkm_auc, height = 8, width = 10
)

# plot together
gp_gkm_auc_wrap <- ggplot(
  unique(roc_dt[, .(filt, model, test_cell_type, auc, test_peaks_pos)]),
  aes(test_peaks_pos, auc, shape = filt, fill = test_cell_type, color = test_cell_type)
  ) +
  geom_point(position = position_jitter(width = 0.1, height = 0), size = 4) +
  scale_shape_manual(values = structure(
      rep(21:25, 2)[seq_along(itrs)],
      names = filts[itrs]
  )) +
  scale_fill_manual(values = c(ct_cols, bct_cols)) +
  scale_color_manual(values = structure(
      colorspace::darken(c(ct_cols, bct_cols), 0.5),
      names = c(names(ct_cols), names(bct_cols))
  )) +
  guides(
      shape = guide_legend(override.aes = list(size = 5, fill = "grey")),
      fill = "none", color = "none"
  ) +
  facet_wrap("model") +
  theme(
      panel.grid.major = element_line(linewidth = 0.2, color = "grey")
  ) +
  labs(x = "test peaks", y = "test AUC")
ggsave(
    file.path(dirname(fig_dir), "LS-GKM-test-roc-wrap.pdf"),
    gp_gkm_auc_wrap, height = 14, width = 24
)
```

## Cross species

Use Nematostella cell type models for predictions across species

Mouse, cell type level

```{bash eval=FALSE, include=FALSE}
for modelprefix in cnidocyte digestive_filaments_1 digestive_filaments_2 digestive_filaments_3 epidermis_1 epidermis_2 gastro_circular_muscle_1 gastro_circular_muscle_2 gastro_parietal_muscle gastro_unknown_1 gastro_unknown_2 gland muscle_1 muscle_2 neuron_GATA_Islet_1 neuron_GATA_Islet_2 neuron_GATA_Islet_3 neuron_Pou4_FoxL2_1 neuron_Pou4_FoxL2_2 precursors_1 precursors_2 precursors_3
do
    qsub -N Peaks_${modelprefix} -q long-sl7 -l virtual_free=60G,h_vmem=60G,h_rt=96:00:00 -M anamaria.elek@crg.eu -m abe -o log/Peaks_${modelprefix}.out -e log/Peaks_${modelprefix}.err 07_gkmSVM_ant_predict_mmus_Peaks_cell_type.sh ${modelprefix}
done
```

```{bash}
mkdir -p logs_cell_type
for modelprefix in cnidocyte muscle_1 epidermis_1 neuron_GATA_Islet_1 neuron_Pou4_FoxL2_1 gland precursors_1
do
    # other outprefix: Unknown Immature_B_cells Activated_B_cells Regulatory_T_cells Type_II_pneumocytes Cerebellar_granule_cells Collisions Distal_convoluted_tubule Loop_of_henle Collecting_duct DCT_CD Hematopoietic_progenitors Proximal_tubule Proximal_tubule_S3 Podocytes Purkinje_cells SOM_Interneurons Endothelial_I_glomerular
    for outprefix in Ex_neurons_CThPN Ex_neurons_CPN Ex_neurons_SCPN Inhibitory_neurons Enterocytes Hepatocytes Cardiomyocytes Erythroblasts Sperm T_cells B_cells NK_cells Macrophages Alveolar_macrophages Monocytes Microglia Dendritic_cells Astrocytes Oligodendrocytes Type_I_pneumocytes Endothelial_I_cells Endothelial_II_cells
    do
        nohup bash 03_gkmSVM_predict_mmus_Peaks_cell_type.sh ${modelprefix} ${outprefix} > logs_cell_type/03_gkmSVM_predict_mmus_Peaks_cell_type_bg_${modelprefix}_${outprefix}.out 2>&1 &
        nohup bash 03_gkmSVM_predict_mmus_PeakDifferential_cell_type.sh ${modelprefix} ${outprefix} > logs_cell_type/03_gkmSVM_predict_mmus_PeakDifferential_cell_type_bg_${modelprefix}_${outprefix}.out 2>&1 &
    done
done
```

Mouse, broad cell type level

```{bash}
mkdir -p logs_broad_cell_type
for modelprefix in cnidocyte muscle gastro_circular_muscle gastro_parietal_muscle gastro_unknown digestive_filaments epidermis precursors neuron_GATA_Islet neuron_Pou4_FoxL2 gland
do
    for outprefix in T_cells Hepatocytes B_cells Activated_B_cells Ex_neurons_CThPN Ex_neurons_CPN Ex_neurons_SCPN Inhibitory_neurons Enterocytes Cardiomyocytes Cerebellar_granule_cells Endothelial_II_cells Hematopoietic_progenitors Proximal_tubule Proximal_tubule_S3 NK_cells Regulatory_T_cells Erythroblasts Sperm SOM_Interneurons Macrophages Microglia Dendritic_cells Alveolar_macrophages Collisions Distal_convoluted_tubule Loop_of_henle Collecting_duct DCT_CD Astrocytes Type_I_pneumocytes Oligodendrocytes Endothelial_I_cells Endothelial_I_glomerular Monocytes Podocytes Purkinje_cells Immature_B_cells Type_II_pneumocytes Unknown
    do
        nohup bash 07_gkmSVM_predict_mmus_Peaks_broad_cell_type.sh ${modelprefix} ${outprefix} > logs_broad_cell_type/07_gkmSVM_predict_mmus_Peaks_broad_cell_type_${modelprefix}_${outprefix}.out 2>&1 &
        nohup bash 07_gkmSVM_predict_mmus_PeakDifferential_broad_cell_type.sh ${modelprefix} ${outprefix} > logs_broad_cell_type/07_gkmSVM_predict_mmus_PeakDifferential_broad_cell_type_${modelprefix}_${outprefix}.out 2>&1 &
    done
done
```

Drosophila, cell type level

```{bash}
mkdir -p logs_cell_type
for modelprefix in cnidocyte muscle_1 epidermis_1 neuron_GATA_Islet_1 neuron_Pou4_FoxL2_1 gland precursors_1
do
    for outprefix in Neural Somatic_muscle Pharnyx Epidermis Head_ectoderm Ventral_nerve_cord Malpighian_tubule Midgut Yolk
    do
        nohup bash 04_gkmSVM_predict_dmel_Peaks_cell_type.sh ${modelprefix} ${outprefix} > logs_cell_type/04_gkmSVM_predict_dmel_Peaks_cell_type_bg_${modelprefix}_${outprefix}.out 2>&1 &
        nohup bash 04_gkmSVM_predict_dmel_PeakDifferential_cell_type.sh ${modelprefix} ${outprefix} > logs_cell_type/04_gkmSVM_predict_dmel_PeakDifferential_cell_type_bg_${modelprefix}_${outprefix}.out 2>&1 &
    done
done
```

Params

```{r}
sp <- "mmus"
sps_dir <- file.path(res_dir, sp, "specific_peaks", "250bp", out_id)
# test cell types
if (sp == "mmus") {
    sps_nm <- c(
        "T_cells", "Hepatocytes", "B_cells", "Activated_B_cells",
        "Ex_neurons_CThPN", "Ex_neurons_CPN", "Ex_neurons_SCPN",
        "Inhibitory_neurons", "Enterocytes", "Cardiomyocytes",
        "Cerebellar_granule_cells", "Endothelial_II_cells",
        "Hematopoietic_progenitors", "Proximal_tubule", "Proximal_tubule_S3",
        "NK_cells", "Regulatory_T_cells", "Erythroblasts", "Sperm",
        "SOM_Interneurons", "Macrophages", "Microglia", "Dendritic_cells",
        "Alveolar_macrophages", "Collisions", "Distal_convoluted_tubule",
        "Loop_of_henle", "Collecting_duct", "DCT_CD", "Astrocytes",
        "Type_I_pneumocytes", "Oligodendrocytes", "Endothelial_I_cells",
        "Endothelial_I_glomerular", "Monocytes", "Podocytes", "Purkinje_cells",
        "Immature_B_cells", "Type_II_pneumocytes", "Unknown"
    )
} else if (sp == "dmel") {
     sps_nm <- c(
        "Neural", "Somatic_muscle", "Pharnyx", "Epidermis", "Head_ectoderm",
        "Ventral_nerve_cord", "Malpighian_tubule", "Midgut", "Yolk"
    )
}
```

Inspect cross-species prediction results

```{r}
# load model predictions
test_dt <- rbindlist(sapply(pk_nm, function(model_nm) {
    message(sprintf("\nLoading %s predictions:", model_nm))
    test_dt <- rbindlist(sapply(sps_nm, function(nm) {
        message(nm)
        tryCatch({
            fread(file.path(
                sps_dir, sprintf("%s.%s.gkmpredict.txt", model_nm, nm)
            ))
        }, error = function(e) NULL)
    }, USE.NAMES = TRUE, simplify = FALSE), idcol = "testset")
}, USE.NAMES = TRUE, simplify = FALSE), idcol = "model")
setnames(test_dt, c("model", "testset", "peak", "score"))
test_dt[, score := as.numeric(score)]
test_dt <- unique(test_dt)

# save
fwrite(
    test_dt,
    file.path(sps_dir, "models.test.scores.tsv.gz"),
    sep = "\t"
)

# sample data for plotting
set.seed(1950)
test_dt_plot <- test_dt[
    , .SD[sample(seq_len(nrow(.SD)), pmin(1e2, nrow(.SD)))],
    .(testset, model)
]
test_dt_plot[, testset_median := median(score), .(testset, model)]
setorder(test_dt_plot, -testset_median)
# plots
mods <- unique(c(names(ct_cols), names(bc_cols))[
    c(names(ct_cols), names(bc_cols)) %in% unique(test_dt_plot$model)
])
gp_t_list <- lapply(mods, function(m) {
    test_dt_plot_m <- test_dt_plot[model == m]
    test_dt_plot_m[, testset := factor(
        testset, levels = unique(test_dt_plot_m$testset)
    )]
    gp_t <- ggplot(
            test_dt_plot_m,
            aes(testset, score, fill = model)
        ) +
        geom_boxplot() +
        #facet_wrap(vars(model), scales = "free_y") +
        scale_fill_manual(values = c(ct_cols, bc_cols)) +
        labs(title = m) +
        theme(
            axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
            legend.position = "none"
        )
    gp_t
})
pdf(
    file.path(fig_dir, sprintf("models.%s.%s.test.scores.pdf", sp, out_id)),
    width = 18, height = 6
)
print(gp_t_list)
dev.off()
```

Inspect cross-species prediction results for specific or non-specific peaks

```{r}
# specificity scores
data_dir <- switch(
    sp,
    "mmus" = "../../cross_species/mmus/adult/",
    "dmel" = "../../cross_species/dmel/data"
)
spks_dt <- fread(
    file.path(data_dir, "peak_specificity_scores.tsv"),
    sep = "\t"
)
setnames(spks_dt, c("peak", "testset", "specificity_score"))
tdt <- merge.data.table(
    test_dt, spks_dt, by = c("peak", "testset"),
    all = FALSE, sort = FALSE
)
# label specific peaks
tdt[, specificity_score_thr := quantile(specificity_score, 0.9), testset]
tdt[, differential_peak := ifelse(
    specificity_score > specificity_score_thr, TRUE, FALSE
)]
tdt[, differential_peak := factor(differential_peak, levels = c(TRUE, FALSE))]
tdt[, model_alpha := paste0(
    model,
    ifelse(differential_peak == TRUE, "", "_alpha")
)]
# colors
all_cols <- c(
    bc_cols, ct_cols,
    structure(
        alpha(bc_cols, 0.3),
        names = paste0(names(bc_cols), "_alpha")
    ),
    structure(
        alpha(ct_cols, 0.3),
        names = paste0(names(ct_cols), "_alpha")
    ),
    "TRUE" = "#000000", "FALSE" = alpha("#000000", 0.3)
)
# order
tdt[, testset_median := median(score), .(testset, model_alpha)]
setorder(tdt, model_alpha, -testset_median)
# plots
mods <- unique(c(names(ct_cols), names(bc_cols))[
    c(names(ct_cols), names(bc_cols)) %in% unique(tdt$model)
])
gp_t_list <- lapply(mods, function(m) {
    tdt_m <- tdt[model == m]
    tdt_m[, testset := factor(
        testset, levels = unique(tdt_m$testset)
    )]
    gp_t <- ggplot(
            tdt[model == m],
            aes(testset, score, fill = model_alpha, color = differential_peak)
        ) +
        geom_boxplot(
            aes(color = model_alpha), outlier.colour = NULL
        ) +
        geom_boxplot(
            aes(fill = model_alpha, color = differential_peak),
            outlier.colour = NA
        ) +
        #facet_wrap(vars(model), scales = "free_y") +
        theme(
            axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
            legend.position = "none"
        ) +
        labs(title = m) +
        scale_fill_manual(values = all_cols) +
        scale_color_manual(values = all_cols) +
        guides(fill = "none")
    gp_t
})

pdf(file.path(
    fig_dir, sprintf("models.%s.%s.test.scores.groups.pdf", sp, out_id)
    ), width = 8, height = 6
)
print(gp_t_list)
dev.off()
```

```{r}
# specificity scores
data_dir <- switch(
    sp,
    "mmus" = "../../cross_species/mmus/adult/",
    "dmel" = "../../cross_species/dmel/data"
)
spks_dt <- fread(
    file.path(data_dir, "peak_specificity_scores.tsv")
)

setnames(spks_dt, c("peak", "testset", "specificity_score"))
test_dt <- fread(
    file.path(sps_dir, "models.test.scores.tsv.gz")
)
tssp_dt <- merge.data.table(
    test_dt, spks_dt, by = c("peak", "testset"),
    all = FALSE, sort = FALSE
)

score_dt <- copy(tssp_dt)
score_dt[, specificity_score_thr := quantile(specificity_score, 0.99), testset]
score_dt <- score_dt[specificity_score > specificity_score_thr]
score_dt[model=="cnidocyte"][, .N, .(testset)]
gp_t_dens <- ggplot(score_dt, aes(specificity_score, score, color = model)) +
    #geom_point() +
    #geom_hex(bins = 100) +
    #scale_color_continuous(type = "viridis") +
    geom_hline(yintercept = 0) +
    stat_density_2d() +
    facet_grid(model ~ testset) +
    scale_x_log10() +
    scale_color_manual(values = ct_cols) +
    theme(
        legend.position = "none",
        text = element_text(size = 12),
        strip.text = element_text(size = 10)
  )

pdf(file.path(
    fig_dir, sprintf("models.%s.%s.test.scores.density.top1.pdf", sp, out_id)
    ), width = 15, height = 10
)
print(gp_t_dens)
dev.off()
```

## Motifs

```{r}
mot_fn <- list.files(
    file.path(res_dir, out_id),
    pattern = "*meme", full.names = TRUE
)
names(mot_fn) <- str_remove(basename(mot_fn), ".pwm_models.meme")
mot_ls <- sapply(names(mot_fn), function(x) {
    ms <- universalmotif::read_meme(mot_fn[x])
    names(ms) <- paste0("GKM_", x, "_motif_", seq_along(ms))
    sapply(names(ms), function(y) {
        m <- ms[[y]]
        m@name <- y
        m
    }, simplify = FALSE, USE.NAMES = FALSE)
}, simplify = FALSE, USE.NAMES = TRUE)

saveRDS(mot_ls, file.path(res_dir, out_id, "pwm_models.RDS"))
```