---
title: "ChromBPNet"
format: html
---

## Setup

Load enviroment

```{bash}
# source /staging/leuven/stg_00002/lcb/nkemp/software/anaconda3/etc/profile.d/conda.sh
conda activate chrombpnet
# module load cuDNN/8.7.0.84-CUDA-11.8.0 
module load cuDNN/8.2.1.32-CUDA-11.3.1
```

Set variables

```{bash}
wdir=/staging/leuven/stg_00002/lcb/aelek/chrombpnet_nvec/
genome=/staging/leuven/stg_00002/lcb/aelek/data_nvec/Nvec_vc1.1_gDNA.fasta
peaks=/staging/leuven/stg_00002/lcb/aelek/data_nvec/Peaks_cell_type_mapped_250bp.bed
chromsizes=/staging/leuven/stg_00002/lcb/aelek/data_nvec/Nvec_vc1.1_gDNA.chrom.sizes
```

### Prepare fragment files

Prepare per-cell type unshifted files from input bam files in scATAC clustering analysis, `09_Figures.Rmd`, section `Group files`

### Generate splits (train, test, validate)

```{bash}
mkdir -p ${wdir}/data/splits
chrombpnet prep splits \
    -c ${chromsizes} \
    -tcr NC_064035.1 \
    -vcr NC_064034.1 \
    -op ${wdir}/data/splits
```

Inspect peaks

```{r}
library(data.table)
library(GenomicRanges)
library(rjson)

# files
peaks_fn <- "data/Peaks_cell_type_mapped_250bp.bed"
splits_fn <- "data/splits.json"

# load peaks
peaks <- fread(peaks_fn, select = c(1:4, 10))
setnames(peaks, c("seqnames", "start", "end", "peak", "summit"))

# load splits
splits <- fromJSON(file=splits_fn)

# split peaks
splits_peaks <- rbindlist(sapply(names(splits), function(st) {
    chr = splits[[st]]
    peaks[seqnames %in% chr]
}, simplify=FALSE, USE.NAMES=TRUE), idcol="split")

splits_peaks[,.N,split]
setcolorder(splits_peaks, c("seqnames", "start", "end", "peak", "summit", "split"))

# check how many peaks will be outside chr ranges when extended from summit

```

### Select GC-matched non peak regions

Since the genome we dealt with are much smaller, I adjusted the stride when creating non-peak (negatives) for training both bias model and chrombpnet model. In my case, I have used 50 or 100, but here when you subsample for using in the model, you may want to make sure that they are well-represented genome-wide.

```{bash}
inlen=500

chrombpnet prep nonpeaks \
    -g ${genome} \
    -p ${peaks} \
    -c ${chromsizes} \
    -st 100 \
    -il ${inlen} \
    -npr 2 \
    -fl ${wdir}/data/splits.json \
    -o ${wdir}/data/nvec_${inlen}

```

### Train bias model

First make sure that using default parameters works

```{bash}
bias_name=neuron_Pou4_FoxL2_2
chrombpnet bias pipeline \
    -ibam /staging/leuven/stg_00002/lcb/aelek/data_nvec/GroupBam/${bias_name}.bam \
    -d "ATAC" \
    -g ${genome} \
    -c ${chromsizes} \
    -p ${peaks} \
    -n ${wdir}/data/nvec_negatives.bed \
    -fl ${wdir}/data/splits.json \
    -b 0.5 \
    -o ${wdir}/bias_model/${bias_name}_default \
    > logs/chrombpnet_bias_pipeline_${bias_name}_default.log 2>&1 &
```

You will encounter a AT-rich bias instead of GC Tn5 bias you will see from tutorial. And while training bias model, you want to use the population that is most abundant, and use that bias model to regress across all cell types. 

```{bash}
bias_name=neuron_Pou4_FoxL2_2

inlen=1000
outlen=800

chrombpnet bias pipeline \
    -ibam /staging/leuven/stg_00002/lcb/aelek/data_nvec/GroupBam/${bias_name}.bam \
    -d "ATAC" \
    -g ${genome} \
    -c ${chromsizes} \
    -p ${peaks} \
    -n ${wdir}/data/nvec_1000_negatives.bed \
    -fl ${wdir}/data/splits.json \
    -b 0.5 \
    --inputlen ${inlen} \
    --outputlen ${outlen} \
    -o ${wdir}/bias_model/${bias_name}_inputlen_${inlen}_outputlen_${outlen} \
    > logs/chrombpnet_bias_pipeline_${bias_name}_inputlen_${inlen}_outputlen_${outlen}.log 2>&1 &

```

### Inspect bias model

```{python}
import pandas as pd
import matplotlib.pyplot as plt

dir='/staging/leuven/stg_00002/lcb/aelek/chrombpnet_nvec/'
bias_name='neuron_Pou4_FoxL2_2_default'
bias_name='neuron_Pou4_FoxL2_2_inputlen_1000_outputlen_500'
bias_dir=dir+'bias_model/'+bias_name+'/'

# Load train and validation loss per epoch
loss_dt=pd.read_csv(bias_dir+'logs/bias.log')

# Plot loss per epoch
fig, axes = plt.subplots(1, 3, figsize=(12, 4))

# Plotting logcount_predictions_loss and val_logcount_predictions_loss
axes[0].plot(loss_dt["epoch"], loss_dt["logcount_predictions_loss"], label="Training")
axes[0].plot(loss_dt["epoch"], loss_dt["val_logcount_predictions_loss"], label="Validation")
axes[0].set_title("Log Count Predictions Loss")
axes[0].set_xlabel("Epoch")
axes[0].set_ylabel("Loss")
axes[0].legend()
axes[0].grid(True)

# Plotting logits_profile_predictions_loss and val_logits_profile_predictions_loss
axes[1].plot(loss_dt["epoch"], loss_dt["logits_profile_predictions_loss"], label="Training")
axes[1].plot(loss_dt["epoch"], loss_dt["val_logits_profile_predictions_loss"], label="Validation")
axes[1].set_title("Logits Profile Predictions Loss")
axes[1].set_xlabel("Epoch")
axes[1].set_ylabel("Loss")
axes[1].legend()
axes[1].grid(True)

# Plotting loss and val_loss
axes[2].plot(loss_dt["epoch"], loss_dt["loss"], label="Training")
axes[2].plot(loss_dt["epoch"], loss_dt["val_loss"], label="Validation")
axes[2].set_title("Total Loss")
axes[2].set_xlabel("Epoch")
axes[2].set_ylabel("Loss")
axes[2].legend()
axes[2].grid(True)

plt.tight_layout()
plt.savefig(bias_dir+'logs/bias.pdf')

plt.show()
```

### Train a bias-factorized ChromBPNet model

First again try with default parameters.

```{bash}
# cell type to train the model for
name=neuron_Pou4_FoxL2_2

# bias model to use
bias_name=neuron_Pou4_FoxL2_2

# model training
chrombpnet pipeline \
    -ibam /staging/leuven/stg_00002/lcb/aelek/data_nvec/GroupBam/${name}.bam \
    -d "ATAC" \
    -g ${genome} \
    -c ${chromsizes} \
    -p ${peaks} \
    -n ${wdir}/data/nvec_negatives.bed \
    -fl ${wdir}/data/splits.json \
    -b ${wdir}/bias_model/${bias_name}_default/models/bias.h5 \
    -o ${wdir}/chrombpnet_model/${name}_default \
    > logs/chrombpnet_pipeline_${name}_default.log 2>&1 &

```

Changing the size of the filter length for the input convolution layer helps a lot with small genome for chrombpnet model (I have tried 9 bp) 

```{bash}
# cell type to train the model for
name=neuron_Pou4_FoxL2_2

# bias model to use
bias_name=neuron_Pou4_FoxL2_2
inlen=1000
outlen=800

# model training
chrombpnet pipeline \
    -ibam /staging/leuven/stg_00002/lcb/aelek/data_nvec/GroupBam/${name}.bam \
    -d "ATAC" \
    -g ${genome} \
    -c ${chromsizes} \
    -p ${peaks} \
    -n ${wdir}/data/nvec_1000_negatives.bed \
    -fl ${wdir}/data/splits.json \
    -b ${wdir}/bias_model/${bias_name}/models/bias.h5 \
    --inputlen ${inlen} \
    --outputlen ${outlen} \
    -o ${wdir}/chrombpnet_model/${name}_inputlen_${inlen}_outputlen_${outlen} \
    > logs/chrombpnet_pipeline_${name}_inputlen_${inlen}_outputlen_${outlen}.log 2>&1 &
    
```

Run from CL

```{bash}
cd scripts
bias_name=neuron_Pou4_FoxL2_2
# running
name=digestive_filaments_1
name="gastro_circular_muscle_1"
name="gastro_parietal_muscle"
name="gastro_unknown_2"
name="gland"
name="precursors_1"
name="precursors_2"
name="precursors_3"
name="gastro_circular_muscle_2"
input_len=500
output_len=250
sbatch chrombpnet_train.sh ${name} ${bias_name} -i ${input_len} -o ${output_len}
```

## Predictions

```{bash}
cd scripts

name=cnidocyte

input_len=500
output_len=250
model_file=/staging/leuven/stg_00002/lcb/aelek/chrombpnet_nvec/chrombpnet_model/${name}_inputlen_${input_len}_outputlen_${output_len}/models/chrombpnet_nobias.h5
bigwig_file=/staging/leuven/stg_00002/lcb/aelek/data_nvec/GroupBigWigs/cell_type/${name}.bw
sbatch chrombpnet_predict.sh ${model_file} ${bigwig_file}

input_len=1000
output_len=800
model_file=/staging/leuven/stg_00002/lcb/aelek/chrombpnet_nvec/chrombpnet_model/${name}_inputlen_${input_len}_outputlen_${output_len}/models/chrombpnet_nobias.h5
bigwig_file=/staging/leuven/stg_00002/lcb/aelek/data_nvec/GroupBigWigs/cell_type/${name}.bw
sbatch chrombpnet_predict.sh ${model_file} ${bigwig_file}
```

Inspect predictions accurracy

```{python}
import h5py
import numpy as np

name='cnidocyte'
input_len=500
output_len=250


```

## TF Modisco
```{bash}
cd scripts

#NAME=neuron_GATA_Islet_1
#NAME=muscle_1
#NAME=neuron_Pou4_FoxL2_2
#NAME=epidermis_1
#INPUT_LEN=1000
#OUTPUT_LEN=800
#MODEL_DIR=/staging/leuven/stg_00002/lcb/aelek/chrombpnet_nvec/chrombpnet_model/${NAME}_inputlen_${INPUT_LEN}_outputlen_${OUTPUT_LEN}
#MODEL_HEAD=profile
#
#sbatch chrombpnet_tfmodisco.sh ${NAME} ${MODEL_DIR}/motifs/${NAME}_${MODEL_HEAD} \
#    -s ${MODEL_DIR}/predictions/${NAME}.${MODEL_HEAD}_sequences.npz \
#    -a ${MODEL_DIR}/predictions/${NAME}.${MODEL_HEAD}_attributions.npz \
#    -w 250 -z 10


NAME=neuron_GATA_Islet_1
NAME=muscle_1
NAME=neuron_Pou4_FoxL2_2
NAME=epidermis_1
NAME="cnidocyte"
INPUT_LEN=500
OUTPUT_LEN=250
MODEL_DIR=/staging/leuven/stg_00002/lcb/aelek/chrombpnet_nvec/chrombpnet_model/${NAME}_inputlen_${INPUT_LEN}_outputlen_${OUTPUT_LEN}
MODEL_HEAD=profile
sbatch chrombpnet_tfmodisco.sh ${NAME} ${MODEL_DIR}/motifs/${NAME}_${MODEL_HEAD} \
    -i ${MODEL_DIR}/predictions/${NAME}.${MODEL_HEAD}_scores.h5 \
    -w 250 -z 10
	
```

## TF Modisco results

Parse TF Modisco outputs to save trimmed patterns.

```{bash}
source /staging/leuven/stg_00002/lcb/sdewin/Programs/anaconda3/etc/profile.d/conda.sh
conda activate tfmodsico-lite

wdir=/staging/leuven/stg_00002/lcb/aelek/chrombpnet_nvec/
cell_types_adult="cnidocyte gastro_circular_muscle_1 gastro_circular_muscle_2 gastro_parietal_muscle gastro_unknown_1 gastro_unknown_2 muscle_1 muscle_2 digestive_filaments_1 digestive_filaments_2 digestive_filaments_3 epidermis_1 epidermis_2 precursors_1 precursors_2 precursors_3 neuron_GATA_Islet_1 neuron_GATA_Islet_2 neuron_Pou4_FoxL2_1 neuron_Pou4_FoxL2_2 neuron_Pou4_FoxL2_3 gland"
cell_types_gastrula="ectoderm EMS ecto_aboral ecto_pharynx EMS_ecto_boundary cnidocyte_gastrula gland_mucin neuronal NPC"

# iterate over cell types
for name in $cell_types_gastrula;
do
    echo
    echo $name
    echo
    infile=${wdir}/chrombpnet_model/${name}_inputlen_500_outputlen_250/motifs/${name}_profile.modisco_results.h5
    outfile=${wdir}/chrombpnet_model/${name}_inputlen_500_outputlen_250/motifs/${name}_profile_pos_patterns.txt
    # check that input file exists
    if [ ! -f ${infile} ]; then
        echo "Input file "${infile}" not found"
    # check that output file doesn't exist
    elif [ -f ${outfile} ]; then
        echo "Output file "${outfile}" already exists"
    else
      python parse_tfmodisco_results.py $name
    fi
done
```

Below is the code we run with `parse_tfmodisco_results.py`

```{bash}
source /staging/leuven/stg_00002/lcb/sdewin/Programs/anaconda3/etc/profile.d/conda.sh
conda activate tfmodsico-lite
```

Helper functions

```{python}
import h5py
import numpy as np
import pandas as pd
import modiscolite as modisco

# Helper function to trim by IC
def trim_pattern_by_ic(
    pattern: np.ndarray,
    min_v: float,
    background: list = [0.27, 0.23, 0.23, 0.27],
    pseudocount: float = 0.5,
    return_positions: bool = False
) -> np.ndarray:
    ic = modisco.util.compute_per_position_ic(
        ppm=pattern,
        background=background,
        pseudocount=pseudocount
    )
    _ = np.nan_to_num(ic, copy=False, nan=0.0)
    v = (abs(pattern) * ic[:, None]).max(1)
    v = (v - v.min()) / (v.max() - v.min())
    start_idx = min(np.where(np.diff((v > min_v) * 1))[0])
    end_idx = max(np.where(np.diff((v > min_v) * 1))[0]) + 1
    if return_positions:
        return pattern[start_idx:end_idx, :], start_idx, end_idx
    else:
        return pattern[start_idx:end_idx, :]

# Function to load TF modisco reuslts and trim by IC
def read_patterns_from_h5(h5_file_path, min_ic_threshold=0.1):
    """
    Reads and trims patterns from an HDF5 file.
    
    Parameters:
        h5_file_path (str): Path to the HDF5 file.
        min_ic_threshold (float): Minimum IC threshold for trimming.
    
    Returns:
        dict: A dictionary containing the trimmed patterns.
    """
    patterns = {}
    contributions = {}
    hypothetical_contributions = {}
    with h5py.File(h5_file_path, 'r') as f:
        if 'pos_patterns' in f:
            pattern_group = f['pos_patterns']
            for pattern_name in pattern_group:
                pattern = pattern_group[pattern_name]['sequence'][:]
                new_pattern_name = "pos_"+pattern_name
                # Trim by IC
                trimmed_positions = trim_pattern_by_ic(pattern, min_ic_threshold, return_positions=True)
                trimmed_pattern = trimmed_positions[0]
                patterns[new_pattern_name] = trimmed_pattern
                # Contribution scores
                start_idx = trimmed_positions[1]
                end_idx = trimmed_positions[2]
                contrib_scores = pattern_group[pattern_name]['contrib_scores'][start_idx:end_idx]
                contributions[new_pattern_name] = contrib_scores
                hypothetical_contrib_scores = pattern_group[pattern_name]['hypothetical_contribs'][start_idx:end_idx]
                hypothetical_contributions[new_pattern_name] = hypothetical_contrib_scores
        if 'neg_patterns' in f:
            pattern_group = f['neg_patterns']
            for pattern_name in pattern_group:
                pattern = pattern_group[pattern_name]['sequence'][:]
                new_pattern_name = "neg_"+pattern_name
                # Trim by IC
                trimmed_positions = trim_pattern_by_ic(pattern, min_ic_threshold, return_positions=True)
                trimmed_pattern = trimmed_positions[0]
                patterns[new_pattern_name] = trimmed_pattern
                # Contribution scores
                start_idx = trimmed_positions[1]
                end_idx = trimmed_positions[2]
                contrib_scores = pattern_group[pattern_name]['contrib_scores'][start_idx:end_idx]
                contributions[new_pattern_name] = contrib_scores
                hypothetical_contrib_scores = pattern_group[pattern_name]['hypothetical_contribs'][start_idx:end_idx]
                hypothetical_contributions[new_pattern_name] = hypothetical_contrib_scores
    patterns_dict = {
        'patterns': patterns,
        'contributions': contributions,
        'hypothetical_contributions': hypothetical_contributions
    }
    return patterns_dict

# Function to save patterns to text
def write_patterns_to_txt(patterns, txt_file_path,):
    """
    Writes patterns to a txt file.
    
    Parameters:
        patterns (dict): A dictionary containing the patterns.
        txt_file_path (str): Path to the txt file
    """
    # Save patterns
    f = open(txt_file_path, 'w')
    # Save names
    nms_file_path = txt_file_path.replace(".txt", ".names")
    n = open(nms_file_path, 'w')
    for pattern_name, pattern_matrix in patterns.items():
        pd.DataFrame(pattern_matrix).to_csv(f, sep="\t", header=False, index=False)
        f.write("\n")
        n.write(f'{pattern_name}\n')
    f.close()
    n.close()

```

TF Modisco results to load

```{python}
# Path to TF modisco results
name="muscle_1"
name="neuron_GATA_Islet_1"
name="neuron_Pou4_FoxL2_2"
name="cnidocyte"
name="epidermis_1"
input_len=1000
output_len=800
model_dir="chrombpnet_model/"+name+"_inputlen_"+str(input_len)+"_outputlen_"+str(output_len)+"/"
model_head="profile"
modisco_res=model_dir+"motifs/"+name+"_"+model_head+".modisco_results.h5"
```

Load patterns, trim by IC and save as txt files.

```{python}
# Read and trim both positive and negative patterns
patterns = read_patterns_from_h5(modisco_res, min_ic_threshold=0.1)

# Save patterns to text
for group in ['pos_pattern', 'neg_pattern']:
    # select patterns that have key starting with 'pos_pattern'
    patterns_ = {k: v for k, v in patterns['patterns'].items() if k.startswith(group)}
    if len(patterns_) > 0:
        print(f"Writing {len(patterns_)} {group} patterns to txt file")
        write_patterns_to_txt(
            patterns = patterns_,
            txt_file_path = model_dir+"motifs/"+name+"_"+model_head+f"_{group}s.txt"
        )

# Contributioin scores
contributions = {}
hypothetical_contributions = {}
for pattern_name, pattern_matrix in patterns['patterns'].items():
    contrib_scores = patterns['contributions'][pattern_name]
    hyp_contrib_scores = patterns['hypothetical_contributions'][pattern_name]
    contributions[pattern_name] = contrib_scores
    hypothetical_contributions[pattern_name] = hyp_contrib_scores

# Max contribution scores per pattern
max_contributions = {pattern_name: abs(contrib_scores).max() * np.sign(contrib_scores.mean()) for pattern_name, contrib_scores in contributions.items()}

# Save contribution scores to table
pd.DataFrame(
    max_contributions.items(),
    columns=['pattern', name]
).to_csv(
    model_dir+"motifs/"+name+"_"+model_head+'_pattern_to_task_contribution.tsv',
    sep="\t",
    index=False
)
```

## Combine results from multiple cell type models

```{bash}
conda activate chrombpnet
```

```{r}
cell_types_adult <- c(
  "cnidocyte",
  "gastro_circular_muscle_1",
  "gastro_circular_muscle_2",
  "gastro_parietal_muscle",
  "gastro_unknown_1",
  "gastro_unknown_2",
  "muscle_1",
  "muscle_2",
  "digestive_filaments_1",
  "digestive_filaments_2",
  "digestive_filaments_3",
  "epidermis_1",
  "epidermis_2",
  "precursors_1",
  "precursors_2",
  "precursors_3",
  "neuron_GATA_Islet_1",
  "neuron_GATA_Islet_2",
  "neuron_Pou4_FoxL2_1",
  "neuron_Pou4_FoxL2_2",
  "neuron_Pou4_FoxL2_3",
  "gland"
)
cell_types_gastrula <- c(
  "ectoderm",
  "EMS",
  "ecto_aboral",
  "ecto_pharynx",
  "EMS_ecto_boundary",
  "cnidocyte_gastrula",
  "gland_mucin",
  "neuronal",
  "NPC"
)

model_head <- "profile"

models <- c(cell_types_adult, cell_types_gastrula)
in_len <- 500
out_len <- 250

res_dir <- "results"
fig_dir <- "figures"
dir.create(res_dir, showWarnings = FALSE)
dir.create(fig_dir, showWarnings = FALSE)
```

Load data

```{r}
library(data.table)
library(stringr)
library(universalmotif)

# where are the results
models_dirs <- sapply(models, function(name) {
  file.path(
    "chrombpnet_model/", 
    sprintf("%s_inputlen_%s_outputlen_%s", name, in_len, out_len),
    "motifs/"
  )
}, USE.NAMES = TRUE)

# patterns to cell type topic contributions
pattern_dt <- rbindlist(lapply(models, function(name) {
  print(name)
  # where are the results
  tfmodisco_dir <- models_dirs[name]
  # patterns contribution file
  pattern_fn <- file.path(tfmodisco_dir, sprintf("%s_profile_pattern_to_task_contribution.tsv", name))
  # load contributions matrix
  tryCatch({
    pattern_dt <- fread(pattern_fn)
    pattern_dt <- melt.data.table(pattern_dt, id.vars = "pattern", variable.name = "cell_type", value.name = "contribution")
    pattern_dt[, pattern := sprintf("%s_%s", name, pattern)][]
  }, error = function(e) {
    message("Error reading ", pattern_fn)
    NULL
  })
}))

# save
fwrite(pattern_dt, file.path(res_dir, "profile_pattern_to_task_contribution.tsv"), sep = "\t")

# only keep models for which we have contributions
models <- as.character(unique(pattern_dt$cell_type))

# load patterns (motifs)
pattern_list <- sapply(models, function(name) {
  # where are the results
  tfmodisco_dir <- models_dirs[name]
  # motif files
  pats_pos_fns <- file.path(tfmodisco_dir, sprintf("%s_profile_pos_patterns.txt", name))
  pats_neg_fns <- file.path(tfmodisco_dir, sprintf("%s_profile_neg_patterns.txt", name))
  # load motifs
  pats_pos <- tryCatch(universalmotif::read_matrix(pats_pos_fns, positions = "rows", headers = FALSE), error = function(e) NULL)
  pats_neg <- tryCatch(universalmotif::read_matrix(pats_neg_fns, positions = "rows", headers = FALSE), error = function(e) NULL)
  if (!is.null(pats_pos)) {
    message(name, " positive patterns")
    names(pats_pos) <- fread(str_replace(pats_pos_fns, "txt", "names"), header = FALSE)[[1]]
    pats_pos_list <- sapply(names(pats_pos), function(x) {
      m <- pats_pos[[x]]
      m@name <- sprintf("%s_%s", name, x)
      m
    }, simplify = FALSE, USE.NAMES = TRUE)
  } else {
    pats_pos_list <- list()
  }
  if (!is.null(pats_neg)) {
    message(name, " negative patterns")
    names(pats_neg) <- fread(str_replace(pats_neg_fns, "txt", "names"), header = FALSE)[[1]]
    pats_neg_list <- sapply(names(pats_neg), function(x) {
      m <- pats_neg[[x]]
      m@name <- sprintf("%s_%s", name, x)
      m
    }, simplify = FALSE, USE.NAMES = TRUE)
  } else {
    pats_neg_list <- list()
  }
  list("positive_patterns" = pats_pos_list, "negative_patterns" = pats_neg_list)
  
}, USE.NAMES = TRUE, simplify = FALSE)

# save
saveRDS(pattern_list, file.path(res_dir, "patterns_motifs.rds"))
```

Cluster patterns

```{r}
library(data.table)
library(stringr)
library(universalmotif)
library(ComplexHeatmap)
library(ggplot2)
library(ggseqlogo)

source("../scripts/mta_downstream_functions.R")

pattern_dt <- fread(file.path(res_dir, "profile_pattern_to_task_contribution.tsv"), sep = "\t")
pattern_list <- readRDS(file.path(res_dir, "patterns_motifs.rds"))

pats_pos_list <- unlist(lapply(pattern_list, function(x) x[["positive_patterns"]]), recursive = FALSE)
names(pats_pos_list) <- sapply(pats_pos_list, function(x) x@name)
universalmotif::write_matrix(pats_pos_list, file.path(res_dir, "profile_pos_patterns.txt"), overwrite=TRUE)
writeLines(names(pats_pos_list), file.path(res_dir, "profile_pos_patterns.names"))
pats_neg_list <- unlist(lapply(pattern_list, function(x) x[["negative_patterns"]]), recursive = FALSE)
names(pats_neg_list) <- sapply(pats_neg_list, function(x) x@name)
universalmotif::write_matrix(pats_neg_list, file.path(res_dir, "profile_neg_patterns.txt"), overwrite=TRUE)
writeLines(names(pats_neg_list), file.path(res_dir, "profile_neg_patterns.names"))

pats_list <- c(pats_pos_list, pats_neg_list)


# remove motifs shorter than 3bp
short_pats <- which(sapply(pats_list, function(m) dim(m@motif)[2])<3)
pats_list <- pats_list[!names(pats_list) %in% names(short_pats)]

# cluster and merge positive patterns
merge_pats_list <- mta_merge_motifs_by_similarity(
  motifs = pats_list,
	matrix_class = "PWM",
	threshold = 0.7,
	dist_method = "WPCC",
	clus_method = "hclust",
	min_overlap = 5,
	eps_dbscan = 1,
	do_heatmap = TRUE
)

# how many motifs per cluster
clust_dt <- as.data.table(merge_pats_list$clusters)
setnames(clust_dt, "motif", "pattern")
clust_dt[, .N, cluster][order(-N)]
fwrite(clust_dt, file.path(res_dir, "patterns_clust.tsv"), sep="\t")

# save similarity matrix
sim_mat <- merge_pats_list$matrix
saveRDS(sim_mat, file.path(res_dir, "patterns_similarity_matrix.rds"))

# save heatmap
pdf(file.path(fig_dir, "hm_clustering_patterns.pdf"), width=8, height=6)
merge_pats_list$matrix_hm
dev.off()

# plot logos
mots <- pats_list[clust_dt$pattern]
pdf(file.path(fig_dir, "hm_clustering_patterns_motifs.pdf"), width = 12, height = 4)
for (i in seq_along(mots)) {
  m <- mots[[i]]
  tryCatch({
    gg_mot <- ggplot() + ggseqlogo::geom_logo(
        m@motif, seq_type = "dna", col_scheme = "nucleotide"
      ) + 
      labs(title = sprintf("%s; cluster %s", m@name, clust_dt[pattern==m@name]$cluster)) +
      theme_logo()
    print(gg_mot)
 }, error = function(e) warning(e))
}
dev.off()
```